---
title: "2024-09-03_assemble_metadata"
author: "Lucore"
date: "2024-09-03"
output: html_document
---

NOTES: This file compiles the metadata for the shotgun samples to be used in downstream analyses. 

# 1. Environmnet set up 

```{r}

library(tidyverse)

sample_log <- read_csv("data/metadata/sample_log.csv")

tax_table <- read_table("/Users/Jordan/Desktop/research_repos/tables_shotgun/bracken_abundance_species_mpa.txt")

```


# 2. Acquire basic metadata from sample log

```{r}

col_names <- colnames(tax_table) # extract sample names
  
samples <- data.frame(sampleid = col_names) # make new df

samples <- samples %>% 
  mutate(sampleid = str_replace(sampleid, "JL.*", "JL")) %>% # remove characters after JL
  filter(sampleid != "#Classification") %>% # remove old header row
  mutate(samplename = str_extract(sampleid, "(?<=-)[^-]+(?=-)"), # extract monkey name and sample number
         sample_number = str_extract(samplename, "\\d+"), # separate sample numbers
         sample_number = as.numeric(str_replace(sample_number, "^0+", "")), # remove leading zeros
         monkey_id = str_extract(sampleid, "[A-Za-z]+")) # extract monkey id
  
sample_log1 <- sample_log %>%
  rename(sample_number = Sample_number,
         collector = `Collector ID`,
         group = Group,
         date = Date,
         time = Time) %>%
  mutate(date = dmy(date),
         sample_number = as.numeric(sample_number)) %>%
  select(-ID)
  
m <- samples %>%
  left_join(sample_log1, by = "sample_number") %>%
  mutate(time = hm(time)) %>%
  rename(ID = monkey_id,
         sample = sampleid) %>%
  select(-group)


```


# 3.Add demography data 

## 3a. calculate age 

```{r}
#load data
d <- read_csv("data/metadata/2024-09-05_Life History.csv", show_col_types = FALSE)

#create list of id's from sample data
id <- m %>%
  select(ID) 
id <- unique(id)

#tidy data for age calculation
dob <- d %>%
  select(c('Name', 'ID', 'Group', 'Sex', 'Event', 'Date', 'Estimated?')) %>%
  rename(name = Name,
         group = Group, 
         sex = Sex, 
         event = Event, 
         date = Date, 
         estimated = 'Estimated?') %>%
  filter(event == "DOB") %>%
  mutate(date = dmy(date)) #the four that fail to parse are not monkeys with urine samples, do not need to fix this

#filter dob table for ID's 
dob2 <- inner_join(dob, id, by = "ID")

#finish cleaning dob table, join with neopterin table, calculate age
dob3 <- dob2 %>%
  select(c(ID, group, sex, date)) %>%
  rename(DOB = date) %>%
  mutate(group = case_when( # change PED to Tenori
    ID == "PED" ~ "Tenori",
    TRUE ~ group  
  ))

m2 <- left_join(m, dob3, by = "ID")
m2 <- m2 %>% 
  mutate(agedays = as.numeric(date - DOB)) %>%
  mutate(age = round(agedays/ 365, 1)) %>%
  select(-agedays, -DOB)

```

## 3b. add repro status

```{r}

#tidy data
r <- d %>%
  select(c('ID', 'Event', 'Date')) %>%
  rename(event = Event, 
         date = Date) %>%
  mutate(ID = str_squish(ID)) %>%
  mutate(date = dmy(date)) #don't worry about the lines that don't parse

# filter for study females
r1 <- inner_join(r, id, by = "ID") %>%
  filter(event == "Wean/Resumes Cycling" | event == "Conception" | event == "Birth" | event == "First Birth") %>%
  filter(date <= as.Date("2023-08-16")) %>%
  filter(date >= as.Date("2022-01-01"))

# add end study period line for each individual 78
new_rows <- r1 %>%
  group_by(ID) %>%
  summarise(event = "END_sample", date = "2023-07-20") %>%
  mutate(date = ymd(date))

r2 <- bind_rows(r1, new_rows) %>% # add ending lines
  arrange(ID)

# get list of dates for each repro state per ID
date_sequences <- r2 %>%
  group_by(ID) %>%
  mutate(next_date = lead(date) - days(1)) %>% # subtract 1 day from next state to get time period without overlap
  filter(!is.na(next_date)) %>%
  rowwise() %>%
  mutate(date_list = list(seq(date, next_date, by = "day"))) %>%
  ungroup()

date_sequences_unnested <- date_sequences %>%
  select(ID, event, date_list) %>%
  unnest(cols = c(date_list))

# Assign repro state for each date
repro_state <- date_sequences_unnested %>%
  mutate(repro = case_when(event == "Wean/Resumes Cycling" ~ "Cycling",
                           event == "Conception" ~ "Pregnant",
                           event == "Birth" ~ "Lactating",
                           event == "First Birth" ~ "Lactating")) %>%
  mutate(date = ymd(date_list)) %>%
  select(c(ID, repro, date))

# combine with metadata data
m3 <- m2 %>%
  left_join(repro_state, by = c("date", "ID")) %>%
  mutate(repro = case_when(sex == "M" ~ "M",
                           age <= 6 & is.na(repro) ~ "juv",
                           age >=6 & is.na(repro) ~ "Cycling",
                           TRUE ~ repro))
```

# 4. Add scfa data

```{r}

#load data
s <- read_csv("data/metadata/SCFA_results_Jordan_Lucore.csv", show_col_types = FALSE)

s1 <- s %>%
  rename(samplename = 'Sample name',
         PA = `PA (ug/g)`,
         BA = `BA (ug/g)`,
         iVA = `iVA (ug/g)`,
         AA = `AA (ug/g)`) %>%
  mutate(samplename = str_replace(samplename, "_.*", "")) %>%
  mutate(across(everything(), ~ replace(.x, .x == "ND", 0))) %>% #replace not detected with zeros 
  select(samplename, PA, BA, iVA, AA) %>%
  mutate(BA = as.numeric(BA)) %>%
  group_by(samplename) %>%
  mutate(all_scfa = sum(PA, BA, iVA, AA))
  
m4 <- m3 %>%
  left_join(s1, by = "samplename")

```

#  5. Add neopterin data

remember tat you need to do this with a sheet where no column have "days" in the name or else the code will break

```{r}

source("functions/generate_time_intervals.R")
source("functions/full_intervals.R")
source("functions/sample_averages.R")

# load and clean neopterin data
n <- read_csv("data/metadata/2024-07-25_compiled_neopterin.csv") %>%
  #select(sample:datetime1) %>%
  rename(sampleid = sample,
         date_range = date, # rename to work with sample_averages function
         neo = neo_nmol_SG) %>%
  select(date_range, ID, neo)

# first, generate time windows of interest, here choosing pre and post of 3, 5, and 7 days
interval_df <- generate_intervals(m4, c(-3, 3, 5, -5, -7, 7))

# next, combine pre and pot intervals to get a total before and after interval
df_total_3day <- create_full_intervals(interval_df, "3days", "total_3_day")
df_total_5day <- create_full_intervals(interval_df, "5days", "total_5_day")
df_total_7day <- create_full_intervals(interval_df, "7days", "total_7_day")

# combine dataframes from previous step with pre and post interval dataframe
df_combined <- bind_rows(interval_df, df_total_3day, df_total_5day, df_total_7day) %>%
  arrange(sample, date_range)

# calculate neopterin averages for each specific time interval
final <- sample_averages(df_combined, n, "neo")

# add averages to metadata
m5 <- m4 %>%
  left_join(final, by = "sample")

```


# 6. Add weather data

can adjust time periods whenever you want
remember you can also add min temp but sticking with max for now because we know they are highly correlated

```{r}

# load data
w <- read_csv("data/metadata/raw_paloverde_weather.csv") %>%
   mutate(date_range = date)

# load function to calculate avergaes
source("functions/calculate_weather_averages.R")

max_t <- calculate_weather_averages(m5, w, "max_t", c(1:3, 5, 10, 15, 20, 30))
rain <- calculate_weather_averages(m5, w, "rain", c(1:4, 5, 10, 15, 20, 30)) # 4 days is the first interval with no NA's

w_calc <- max_t %>%
  left_join(rain, by = "sample") 

m6 <- m5 %>%
  left_join(w_calc, by = "sample")


```


# 7. Add feeding data

## 7a. combine sample info with feed scans

```{r}

f <- read_csv("data/metadata/2024-02-15_FeedScan.csv") %>%
  rename(date_time2 = datetime,
         ID = monkey_id) %>%
  mutate(date_range = date) %>%
  select(c(date_range, date_time2, ID, food_type, plant_part, plant_name))

# add date variable 
dates <- m6 %>%
  mutate(days14 = date - weeks(2)) 

# Expand the date range for each sample
dates <- dates %>%
  select(c(sample, ID, time, date, days14)) %>%
  pivot_longer(cols = starts_with("days"),
               names_to = "time_window",
               values_to = "window_date") %>%
  mutate(datetime_end = paste(date, time, sep = " ")) %>%
  mutate(datetime_start = paste(window_date, time, sep = " "))

date_range <- dates %>%
  rowwise() %>%
  mutate(date_range = list(seq(window_date, date, by = "day")))

exploded_dates <- date_range %>%
  tidyr::unnest(date_range) %>%
  select(c(sample:date, ID, datetime_end:date_range))
  
# combine with feed data
d <- exploded_dates %>%
  left_join(f, by = c('date_range', 'ID'), relationship = "many-to-many") %>%
  group_by(sample) %>%
  filter(!is.na(date_time2)) %>%
  mutate(datetime_start = ymd_hms(datetime_start)) %>%
  mutate(datetime_end = ymd_hms(datetime_end)) %>%
  mutate(date_time2 = ymd_hms(date_time2)) %>%
  filter(date_time2 >= datetime_start & date_time2 <= datetime_end) # IMPORTANT filtering step to make sure you don't include scans from the same day of collection but after the time of collection

```

## 7b. calculate diet diversity index

calculating diet diversity by modeling richness ~ scan_total up to 2 weeks before sample collection, then extract the residuals and use the residuals as the diet diversity predictor. Negative number means less diversity than expected, positive means more diversity than expected. Doing this so I don't have to rarefy when using Shannon. See Marshall et al., 2021 for methods breakdown

```{r}

# calculate total unique diet items per sample
d1 <- d %>%
  mutate(plant_name = ifelse(food_type == "amphibian_reptile", "amphibian_reptile", plant_name)) %>%  #add not plant items to plant_name 
  mutate(plant_name = ifelse(food_type == "bird", "bird", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "egg", "egg", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "fungi", "fungi", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "gastropod", "gastropod", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "insect", "insect", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "mammal", "mammal", plant_name)) %>% 
  mutate(plant_name = ifelse(food_type == "nurse", "mammary_milk", plant_name)) %>%
  mutate(plant_name = ifelse(food_type == "poop", "poop", plant_name)) %>%
  filter(!is.na(plant_name)) %>%
  rename(food_name = plant_name) %>%
  select(sample, food_name) %>%
  group_by(sample, food_name) %>%
  summarise(total_count = n()) %>%
  group_by(sample) %>%
  summarise(richness = n())
  
# calculate total scans per sample 
total <- d %>%
  group_by(sample) %>%
  summarise(total_scans = n())

#combine data
richness <- d1 %>%
  left_join(total, by = "sample")

hist(richness$richness)

#model data 
m1 <- lm(richness ~ total_scans, d = richness)

summary(m1)

richness$residuals <- residuals(m1) # extract residuals

# visualize residuals - everything looks relatively normal
hist(richness$residuals)

ggplot(richness, aes(x = total_scans, y = richness)) +
  geom_point() + 
  geom_smooth(method = "loess")

# add to metadata 

m7 <- m6 %>% left_join(richness, by = "sample") 

```

## 7c. create part plant variable

Make a proportion of plant part versus animal for each sample, two weeks before the sample

```{r}

d2 <- d %>%
  mutate(food_category = case_when(food_type == "amphibian_reptile" ~ "animal",  #make a general category for plant versus animal
                                   food_type == "bird" ~ "animal",
                                   food_type == "egg" ~ "animal",
                                   food_type == "gastropod" ~ "animal",
                                   food_type == "insect" ~ "animal",
                                   food_type == "mammal" ~ "animal",
                                   food_type == "nurse" ~ NA,
                                   food_type == "poop" ~ NA,
                                   food_type == "unknown" ~ NA,
                                   food_type == "fungi" ~ "plant",
                                   food_type == "plant" ~ "plant")) %>%
  filter(!is.na(food_category)) %>%
  group_by(sample, food_category) %>%
  summarise(count_food = n())  # get category counts

# calculate the total with the NA's filtered out 
total2 <- d2 %>%
  group_by(sample) %>%
  summarise(total_scans = sum(count_food))

category <- d2 %>%
  left_join(total2, by = 'sample') %>%
  mutate(percent = (count_food/total_scans)*100) %>%
  select(sample, food_category, percent) %>%
  pivot_wider(names_from = food_category,
              values_from = percent) %>%
  mutate_all(~ replace_na(., 0))
  
# check that they are 100% correlated
plot(category$animal, category$plant)

cor(category$animal, category$plant)

# add to metadata 

m8 <- m7 %>% left_join(category, by = "sample") 

```


## 7d plant part variable

this compares all the types of plant parts to insect, nurse, or animal

```{r}

d3 <- d %>%
  mutate(food_part = case_when(plant_part == "flower" ~ "flower",  #make food part category
                                   plant_part == "fruit" ~ "fruit",
                                   plant_part == "leaves" ~ "leaves",
                                   plant_part == "nectar" ~ "nectar",
                                   plant_part == "pith" ~ "pith",
                                   plant_part == "seeds" ~ "seeds",
                                   plant_part == "flower" ~ "flower",
                                   food_type == "bird" ~ "animal",
                                   food_type == "egg" ~ "animal",
                                   food_type == "gastropod" ~ "animal",
                                   food_type == "mammal" ~ "animal",
                                   food_type == "amphibian_reptile" ~ "animal",
                                   food_type == "nurse" ~ "nurse",
                                   food_type == "fungi" ~ "fungi",
                                   food_type == "poop" ~ NA,
                                   food_type == "unknown" ~ NA,
                                   food_type == "insect" ~ "insect")) %>%
  filter(!is.na(food_part)) %>%  ## probably not the best way to do this, should account for NA's but doing it for now
  group_by(sample, food_part) %>%
  summarise(count_food_part = n())  # get category counts

# calculate the total with the NA's filtered out 
total2 <- d3 %>%
  group_by(sample) %>%
  summarise(total_scans = sum(count_food_part)) # there are a couple samples with only one scan that might need to be filtered out here

category <- d3 %>%
  left_join(total2, by = 'sample') %>%
  mutate(percent = (count_food_part/total_scans)*100) %>%
  select(sample, food_part, percent) %>%
  pivot_wider(names_from = food_part,
              values_from = percent) %>%
  mutate_all(~ replace_na(., 0))

# add to metadata 

m9 <- m8 %>% left_join(category, by = "sample")

```

# 8. Add number of paired reads

These are the total number of paired reads that make it through the kneaddata step, so the total reads used for the humann3 step. Chose this number instead of the total number before kneaddata because it is after the host filtering step, so it is the total reads used by humann3 for the gene assignment which is what I use for the anlaysis. 

```{r}

reads <- read_csv("bioinfo_processing/shotgun/kneaddata_read_counts.csv") %>%
  rename(reads = 'final pair1', 
         sample = Sample) %>%
  select(sample, reads) %>%
  filter(reads > 5)  %>% # remove the negative sample
  mutate(sample = str_replace(sample, "JL.*", "JL")) # clean sample name

m10 <- m9 %>% left_join(reads, by = "sample")

```


# 9. export metadata

```{r}

write_csv(m10, 'data/metadata/2025-02-17_shotgun_compiled_metadata.csv')

```

