---
title: "2025-02-20 kegg assignments"
author: "Lucore"
date: "2025-02-19"
output: html_document
---

Notes: In this script, I'm trying to figure out how to assign the KEGG Orthogroups (the abundances outputted from humann3) to higher level patheays and brite categories (a.k.a. functional groups). I haven't had much luck with the {KEGGREST} package but I was able to download a json of the KO (Kegg Orthology Database) that assigns function based on Kegg orthogroups. I'm trying to figure out how to wrangle this file now to get the different brite groups and their associated Kegg orthologs. 




# use the downloaded json from this website

websie that explains KEG orthology database https://www.genome.jp/kegg/ko.html
json download for KEG orthology assignment: https://www.genome.jp/kegg-bin/get_htext?ko00001.keg


# script to clean up the json file 

github page with function to clean json: https://gist.github.com/cmkobel/c2f76359eb17e0c00ddd4e83e90bdf99

```{r}

brite <- fromJSON("data/brite_map.json", flatten = F)

kegg_parse_json = function(kegg) {
    # Reads the kegg json (https://www.kegg.jp/kegg-bin/download_htext?htext=ko00001.keg&format=json&filedir=) and outputs a neat table
    
    unnest = kegg$children %>% unnest(children, names_repair = "universal") %>% # AB 57
        unnest(children, names_repair = "universal") %>% # C 551
        unnest(children, names_repair = "universal") %>% # D 59868
    
        rename(
            a_class = 1, # E.g. "09100 Metabolism"
            b_class = 2, # E.g. "09101 Carbohydrate metabolism"
            pathway = 3, 
            reaction = 4
        ) %>%
        separate_wider_regex(pathway, # E.g. "00010 Glycolysis \/ Gluconeogenesis [PATH:ko00010]"
            c(
                pathway = "^[^\\[]+", # Anything that isn't a [
                pathway_id = "\\[.+\\]$"),
            too_few = "align_start"
        ) %>%
        separate_wider_regex(reaction, # E.g. "K00844  HK; hexokinase [EC:2.7.1.1]"
            c( # "name":"K04022  eutG; alcohol dehydrogenase"
                reaction_ko = "^K[0-9]+",
                "  ",
                reaction_abbreviation = "[^;]+", # Anything that isn't a ;
                "; ",
                reaction_name = "[^\\[]+", # Anything that isn't a [
                " \\[",
                reaction_ec = ".+",
                "\\]$"), 
            too_few = "align_start"
        ) %>% 
        mutate(reaction_ec = str_trim(reaction_ec)) %>% 
        mutate(illegal_reaction = is.na(reaction_ko))
    
    # Check and report on the number of reactions with illegal names.
    illegals = unnest %>% filter(illegal_reaction)
    if (nrow(illegals) > 0) {
        message(paste("\nWarning:", nrow(illegals), "reaction(s) have an illegal name."))
        message(illegals$reaction %>% head())
    }
    
    # Return everything that isn't illegal.
    unnest %>% filter(!illegal_reaction) %>% 
        select(-illegal_reaction)
    
}


brite_parsed <- kegg_parse_json(brite)

```

























```{r}

brite <- fromJSON("data/brite_map.json")

brite_flat <- flatten(brite)
str(brite_flat)

brite_flat <- brite_flat %>%
  mutate(children = map(children, ~as.data.frame(.x)))


brite_tidy <- brite_flat %>%
  unnest_wider(children)


```




```{r}

library(jsonlite)

brite <- fromJSON("data/brite_map.json")


# Function to clean KO IDs
clean_ko_id <- function(ko_name) {
  ko_id <- sub(" .*", "", ko_name)  # Keep only the KO number
  return(ko_id)
}

ko_id <- clean_ko_id(brite$name)
results <- list(data.frame(KO = ko_id))


 if (is.list(brite) && !is.null(brite$name) && grepl("^K\\d{5}", brite$name)) {
    ko_id <- clean_ko_id(brite$name)
    results <- list(data.frame(KO = ko_id))
  }

 if (is.list(brite) && "children" %in% names(brite)) {
    children <- brite$children
    if (is.data.frame(children)) {
      children <- as.list(children)
    }
    for (child in children) {
      if (is.list(child)) {
        results <- append(results, tidy_json(child))
      }
    }
  }




# Example processing function (adjust to your JSON structure)
tidy_json <- function(node) {
  results <- list()

  # Check if the node is a KO (third level)
  if (is.list(node) && !is.null(node$name) && grepl("^K\\d{5}", node$name)) {
    ko_id <- clean_ko_id(node$name)
    results <- list(data.frame(KO = ko_id))
  }

  # Process children if they exist
  if (is.list(node) && "children" %in% names(node)) {
    children <- node$children
    if (is.data.frame(children)) {
      children <- as.list(children)
    }
    for (child in children) {
      if (is.list(child)) {
        results <- append(results, tidy_json(child))
      }
    }
  }

  return(results)
}

# Run the processing function on the loaded JSON
ko_mappings <- tidy_json(brite)


```


```{r}

extract_ko_mappings <- function(node, category_chain = character()) {
  results <- list()
  
  # Ensure node is a list and has a "name" field
  if (is.list(node) && !is.null(node$name)) {
    category_chain <- c(category_chain, node$name)
  }
  
  # If node contains children and children is a list, recursively process them
  if (is.list(node) && "children" %in% names(node)) {
    children <- node$children
    
    if (is.list(children)) {
      # Process each child if children is a list
      for (child in children) {
        if (is.list(child)) {
          results <- append(results, extract_ko_mappings(child, category_chain))
        } else {
          warning("Unexpected structure for child node (expected list).")
        }
      }
    } else if (is.data.frame(children)) {
      # If children is a data.frame, process each row as a node
      for (i in 1:nrow(children)) {
        child <- as.list(children[i, , drop = FALSE])  # Convert each row to a list
        results <- append(results, extract_ko_mappings(child, category_chain))
      }
    } else {
      warning("Unexpected structure for children (not list or data.frame).")
    }
  }
  # Handle leaf nodes that are KO IDs
  else if (is.character(node) && grepl("^K\\d{5}", node)) {
    ko_id <- sub(" .*", "", node)  # Keep only the KO number
    results <- list(data.frame(KO = ko_id,  
                               Category1 = category_chain[1],  
                               Category2 = category_chain[2],  
                               Category3 = category_chain[3],  
                               Category4 = category_chain[4],  
                               stringsAsFactors = FALSE))
  }
  
  return(results)  # Return the accumulated results
}
```




# KEGGREST stuff

```{r}

if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("KEGGREST")
BiocManager::install("EnrichmentBrowser")
library("KEGGREST")
library("EnrichmentBrowser")

ko_id <- "ko00001"

pathways <- keggList("pathway", ko_id)

brite_ko <- keggGet("br08901")

pathways <- keggList("brite", ko_id)

brite_map <- keggGet("ko00001")

keggList("brite")

brite_hierarchy <- keggGet("ko00001")

```



# script to clean up the json file 

https://gist.github.com/cmkobel/c2f76359eb17e0c00ddd4e83e90bdf99

```{r}

brite <- fromJSON("data/brite_map.json", flatten = F)

kegg_parse_json = function(kegg) {
    # Reads the kegg json (https://www.kegg.jp/kegg-bin/download_htext?htext=ko00001.keg&format=json&filedir=) and outputs a neat table
    
    unnest = kegg$children %>% unnest(children, names_repair = "universal") %>% # AB 57
        unnest(children, names_repair = "universal") %>% # C 551
        unnest(children, names_repair = "universal") %>% # D 59868
    
        rename(
            a_class = 1, # E.g. "09100 Metabolism"
            b_class = 2, # E.g. "09101 Carbohydrate metabolism"
            pathway = 3, 
            reaction = 4
        ) %>%
        separate_wider_regex(pathway, # E.g. "00010 Glycolysis \/ Gluconeogenesis [PATH:ko00010]"
            c(
                pathway = "^[^\\[]+", # Anything that isn't a [
                pathway_id = "\\[.+\\]$"),
            too_few = "align_start"
        ) %>%
        separate_wider_regex(reaction, # E.g. "K00844  HK; hexokinase [EC:2.7.1.1]"
            c( # "name":"K04022  eutG; alcohol dehydrogenase"
                reaction_ko = "^K[0-9]+",
                "  ",
                reaction_abbreviation = "[^;]+", # Anything that isn't a ;
                "; ",
                reaction_name = "[^\\[]+", # Anything that isn't a [
                " \\[",
                reaction_ec = ".+",
                "\\]$"), 
            too_few = "align_start"
        ) %>% 
        mutate(reaction_ec = str_trim(reaction_ec)) %>% 
        mutate(illegal_reaction = is.na(reaction_ko))
    
    # Check and report on the number of reactions with illegal names.
    illegals = unnest %>% filter(illegal_reaction)
    if (nrow(illegals) > 0) {
        message(paste("\nWarning:", nrow(illegals), "reaction(s) have an illegal name."))
        message(illegals$reaction %>% head())
    }
    
    # Return everything that isn't illegal.
    unnest %>% filter(!illegal_reaction) %>% 
        select(-illegal_reaction)
    
}


test <- kegg_parse_json(brite)

```

