---
title: "2025-03-11 shotgun analysis"
author: "Lucore"
date: "2025-03-11"
output: html_document
---


Notes: Most up to date analysis for the shotgun data. Have decided to use the uniref 50 data based on decisoins from the 2025-02-27_unmapped_reads scripts. We are also only going to work with the KO's for now since we have determined that the databases are redundant with one another, you can't just combine them because the points aren't independent. 

Here we are working on the DAA analysis. Left this where I ran the DAA analysis, found that the effect sizes aren't super reliable but the temp model fits best for the majority of lvl2 and lvl3 categories. Then found that GAM's aren't saying much here, most of the model weights are evenly split between the model with the temp spline and the model without the temp spline. Probably need to give up on the GAM's for this data but I would like to summarize them in a script for evidence that I did the compairson and it didn't work. 

If we aren't going to be able to say much about individual KO direction of effects, might be a god idea to bring back in the pathway data for the community analysis and possible run a DAA analysis on that data? 

finally, is there a way that makes sense to visualuze the results from the DAA? Even though the effect sizes aren't reliable, are there functional groups that are broadly negaitve and broadly positiev? Can we draw more conclusions from this data? I want to push this a bit more before giving up. 

Finally, we will need to figure out how to incoporate the neopterin data. Is there an analysis we could do or a figure I could make to summarize the stratefied data? 

# 1. Load environment

```{r}

library(tidyverse)
library(phyloseq)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(zCompositions)
library(jsonlite)

source("functions/my_scale.R")

source("functions/inverse_scale.R")

source("functions/kegg_parse_json.R")

conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)


ko_physeq_u50 <- readRDS("phyloseq_objects/ko_phylo_u50.rds")


```


# 2. Data set up

## a. Filter metadata

decided to use the 87 samples that have feed data associated with them because I felt like I couldn't run the models without fruit since there is such a strong relationship between fruit consumption and temperature

```{r}

m <- data.frame(sample_data(ko_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

m_filtered <- m[!is.na(m$fruit), ] # remove samples with no diet data, drops 8 samples

# filter phyloseq objects to remove samples with no diet data 
ko_physeq2 <- prune_samples(m_filtered$sample, ko_physeq_u50) # filter physeq object
ko_physeq2 <- prune_taxa(taxa_sums(ko_physeq2) > 0, ko_physeq2) # remove ASV's with zero abundance

```

## b. Prevalence filtering

hold off on prev filtering for now... our filter for the 16s data "remove features that are not present in at least 0.5% of total samples" doesn't remove anything here because that filter that the feature needs to be in at least one sample for this data. 

Unapped values represent the total abundance of reads that failed to align to any known gene sequences and ungrouped values represent reads that mapped to known gene sequences but could not be confidently assigned to a specific functional category.

We are going to remove the UNMAPPED and UNGROUPED reads so that we can just compared the mapped reads for each sample. Removing the UNGROUPED reads because we can think of this as taxa with no phylum assignment. 
Need to think about removing KO's that don't have any functional assignment (similar to the argument of removing ASV's with no phylum assignment)

```{r}

####### remove UNMAPPED and UNGROUPED reads #######

ko_filtered <- subset_taxa(ko_physeq2, 
                                !(grepl("UNMAPPED", gene_ko) | grepl("UNGROUPED", gene_ko)))

```

## c. alpha diversity

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

ignore the warning, it is upset because the data is not in counts. Have to calculate the observed metric manually because `estimate_richness()` only takes count data

```{r}

ko_shannon <- estimate_richness(ko_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(ko_shannon = Shannon) 

ko_alpha <- as.data.frame(colSums(otu_table(ko_filtered) > 0)) %>%
  rename(ko_observed = 'colSums(otu_table(ko_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(ko_shannon, by = "sample")

hist(ko_alpha$ko_shannon) 
hist(ko_alpha$ko_observed) 

m1 <- m_filtered %>% # add to metadata
  left_join(ko_alpha, by = "sample")

```

## d. CLR transformation

doing the CLR transformation on the RPK abundances 

```{r}

ko_table <- as.matrix(t(otu_table(ko_filtered)))
ko_table[1:5,1:5] # rows are samples 

ko_czm <- cmultRepl(ko_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
ko_czm[1:5,1:5] # rows are samples 

ko_clr <- t(apply(ko_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
ko_clr[1:5,1:5] # rows are samples 

ko_pca <- prcomp(ko_clr) # calc pca
plot(ko_pca$x[,1], ko_pca$x[,2]) # plot 

ko_d.mvar <- sum(ko_pca$sdev^2) # total variance
ko_PC1 <- paste("PC1 ","(",round(sum(ko_pca$sdev[1]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC1 30.8%
ko_PC2 <- paste("PC2 ","(",round(sum(ko_pca$sdev[2]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC2 12.5%

######### Add PCA values to metadata #########

ko_df<-data.frame(ko_pca$x)
ko_df$sample<-rownames(ko_df)
ko_df_small <- ko_df %>% 
  select(c(1:4,88)) %>%
  rename_with(~paste0("ko_", .), starts_with("PC"))

# add to metadata
m2 <- m1 %>%
  left_join(ko_df_small, by = "sample") 

```

## f. prep data for modeling

scale variables and make categorical variables factors

```{r}

table(m2$ID)

# scale variables
m_model <- m2 %>%
  mutate(across(c(age, temp, rain, fruit, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```


# 3. Community features analysis

## a. model community features

everything seems normally distributed enough to model with Gaussian except for the Shannon index of the eggnog families which is pretty right skewed but I'm going to move forward with the normal distribution for simplicity of these models and then I can come back to tweak this later if I think it is necessary. 

Going to need to think about how the observed effect size is on a totally difference scale than the shannon diversity effect, which is makeing all of the other effects look quire small

main results here: for the functional metrics, it does not seem like temperature is related to shannon's diversity but it is realted to the PC's and observed gene families. Why is it not related to alpha diversity? Even though taxonomic shannon diversity may decrease at high temperatures, functional alpha diversity may be more resilient to this because many taxa have similar genes and similar functions? However, we don't see the same thing for observed gene families, could speak more to that there is still a decrease in number of functional annotations but the evenness is relatively stable in the community  

```{r}

####### histogranms ########

hist(m_model$ko_shannon) 
hist(m_model$ko_observed) 
hist(m_model$ko_PC1)
hist(m_model$ko_PC2)
hist(m_model$ko_PC3)
hist(m_model$ko_PC4)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15:20)] # list of features to model


for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)

```

## b. plot results

```{r}

#### plot ####

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) #%>%
  #filter(metric != "nog_observed") %>%
  #filter(metric != "ko_observed")
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long <- com_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "PC1", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "PC1", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  


```

# 4. Data set up for DAA

## a. Agglomeration at brite levels

Agglomerate KO's at Brite level 2 and Pathway level from the Keg orthology database. Below are links of where to find the explanation of the database and the link to download the json file. Then the link to the github page with the code to parse the json file. 

websie that explains KEG orthology database https://www.genome.jp/kegg/ko.html
json download for KEG orthology assignment: https://www.genome.jp/kegg-bin/get_htext?ko00001.keg
github page with function to clean json: https://gist.github.com/cmkobel/c2f76359eb17e0c00ddd4e83e90bdf99


```{r}

brite <- fromJSON("data/brite_map.json", flatten = F)

brite_parsed <- kegg_parse_json(brite) %>%
  rename(lvl1 = a_class,
         lvl2 = b_class, 
         lvl3 = pathway)

#################### brite level2 ####################################

brite_level2 <- brite_parsed[,c(2,5)]
length(unique(brite_level2$lvl2)) # 57 unique brite level 2

# agglomerate at brite level 2
lvl2_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_level2, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl2 = replace_na(lvl2, "no_assignment"))

length(unique(lvl2_abund_ko$lvl2)) # 55 lvl2, 1 no_assignment

unique_lvl2 <- unique(lvl2_abund_ko$lvl2)
missing_lvl2 <- brite_level2[!(brite_level2$lvl2 %in% unique_lvl2), ]
unique(missing_lvl2$lvl2) # missing two lvl2 in our dataset

lvl2_abund <- lvl2_abund_ko %>%
  group_by(sample, lvl2) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl2")

unique_lvl2 <- length(unique(lvl2_abund$lvl2))
unique_samples <- length(unique(lvl2_abund$sample))
(unique_lvl2 * unique_samples) == nrow(lvl2_abund) # check that each sample has a value for all lvls


#################### brite level3 ####################################

# pathways (brite level 3)
brite_path <- brite_parsed[,c(3,5)]
length(unique(brite_path$lvl3)) # 561 unique pathways 

#agglomerate at pathway
lvl3_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_path, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl3 = replace_na(lvl3, "no_assignment"))

length(unique(lvl3_abund_ko$lvl3)) # 511 total, 510 lvl3, 1 no_assignment

unique_path <- unique(lvl3_abund_ko$lvl3)
missing_path <- brite_path[!(brite_path$lvl3 %in% unique_path), ]
length(unique(missing_path$lvl3)) # missing 51 pathways in our dataset

lvl3_abund <- lvl3_abund_ko %>%
  group_by(sample, lvl3) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl3")

unique_path <- length(unique(lvl3_abund$lvl3))
unique_samples <- length(unique(lvl3_abund$sample))
(unique_path * unique_samples) == nrow(lvl3_abund) # check that each sample has a value for all lvls

```

## b. Plot distribution

distribution of clr abundances looks kind of normal for the level 2 agglomerations, level 3 doesn't look great but we'll proceed with the normal distribution because I think it is the most appropriate for the clr abundances given that there are negative and positive values

```{r}

################## lvl2 ######################

# plot ditribution 
pdf("figures/lvl2_distributions.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(lvl2_abund$lvl2)) {
  
  data <- subset(lvl2_abund, lvl2 == i)
  
  p <- ggplot(data, aes(x = clr_abundance)) +
    geom_histogram(bins = 15) +
    theme_minimal() +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()


```

## c. Abundance filtering

not sure how to do this with clr transformed values....... maybe I just shouldn't. Lets move on the the daa of the lvl2's and see how it goes

maybe instead of abundance filtering we only filter for the lvls we are interested in, we should probably remove unknowns and unnasigns, and not well described etc. and the assignmentsf from the human disease caetgory

ex. not interested in: 

09190 Not Included in Pathway or Brite


## d. Scale variables

have to do this manually because scale function not working within the tidyverse mutate() command for some reason

```{r}

# Scale variables

lvl2_abund$temp_s <-  my_scale(lvl2_abund$temp) 
lvl2_abund$rain_s <-  my_scale(lvl2_abund$rain)
lvl2_abund$age_s <-  my_scale(lvl2_abund$age)
lvl2_abund$fruit_s <-  my_scale(lvl2_abund$fruit)

lvl2_abund <- lvl2_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex)) 

length(unique(lvl2_abund$lvl2)) # 56 unique groups


lvl3_abund$temp_s <-  my_scale(lvl3_abund$temp) 
lvl3_abund$rain_s <-  my_scale(lvl3_abund$rain)
lvl3_abund$age_s <-  my_scale(lvl3_abund$age)
lvl3_abund$fruit_s <-  my_scale(lvl3_abund$fruit)

lvl3_abund <- lvl3_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex)) 

length(unique(lvl3_abund$lvl3)) # 526 unique groups

```


# 5. Temperature abundance testing

Only using lvl2 now but should probably also try lvl3. Think about removing the groups you are not interested in, like unnasigned and the drug/disease groups

## a. run loop

Had to change the way I track convergence errors for this loop because I'm using lmer insteatd of glmer. 

two don't converge 

```{r}

coef <- list()
aic <- list()
model_results <- list()

for (i in unique(lvl2_abund$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl2_abund, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl2 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    aic[[i]] <- aic_df
    coef[[i]] <- results_df
    model_results[[i]] <- model_results_df

  })
}

# combine results
aic <- do.call(rbind, aic)
coef <- do.call(rbind, coef)
model_results <- do.call(rbind, model_results)


```

## b. select taxa

```{r}
# Remove taxa with convergence warnings 
convergence_warnings <- unique(aic$lvl2[aic$Warning == "Convergence Warning"]) # 2 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 100) %>%
  pull(lvl2) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 5 do not converge


# remove error taxa
lvl2_abund2 <- lvl2_abund %>% filter(!lvl2 %in% errors)
length(unique(lvl2_abund2$lvl2)) # 51 remaining 

coef2 <- coef %>% filter(!lvl2 %in% errors)
length(coef2$lvl2)

aic2 <- aic %>% filter(!lvl2 %in% errors)
length(aic2$lvl2) # 324/2 = 162

# filter for taxa where full model fits best
temp_filter <- aic2 %>%
  group_by(lvl2) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2) 
length(unique(temp_filter)) # temp model fits best for all


## only filter for ones where temp fits best
#filtered_coef <- coef2 %>% filter(lvl2 %in% temp_filter)
#length(filtered_coef$lvl3)

#filtered_aic <- aic2 %>%filter(lvl3 %in% temp_filter)
#length(filtered_aic$lvl3)

```


## c. coef plot

results: for every lvl2 category the temp model fits better, over 85% of the model weight. 35 out of 16 cetgroies have a negative coefficient but none have reliable confidence intervals. Seem like termperature is important for abundance of functional genes but we don't have a lot of certainty about the direction. Still, the maximum likelihood is the best estimate and the majority of the effects are negative so it seems like there is a negative effect on both taxa abundances and abundances of functional genes 

```{r}

plot_coef <- coef2 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl2 = reorder(lvl2, -coef))

table(plot_coef$direction)
         
p1 <- ggplot(plot_coef, aes(x = lvl2, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1

# Reshape data
aic_long <- aic2 %>%
  select(lvl2, model, weight) %>%
  spread(key = model, value = weight)

# Plot model weights
p2 <- ggplot(aic_long, aes(y = lvl2)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "no_assignment", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "no_assignment", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  

```


# 6. Fit GAMS 

do they have certain shapes over different temps? Try all first and then just the negative ones? 

## a. AIC comparison

There doesn't seem to be a purpose of doing this with the shotgun data. I out of 51 lvl2's the gAM fits SLIGHTLY etter and about 27 out of 441 of the lvl3 the gam fits slightly better. Not much certainty around this effect. Not sure if pursuing this is the right thing to do. 

```{r}

# filter abundance
#smooth_abund <- lvl3_abund3 %>% filter(lvl3 %in% temp_filter)
#length(unique(smooth_abund$lvl3))


gam_results <- list()

for (i in unique(lvl2_abund2$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl2_abund2, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)
    
    m_main <- gam(clr_abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)

  }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # Update the warning message
      } 
    })

    # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      #create data frame
      results_df <- data.frame(
        lvl2 = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE
      )
  
      # store results
      gam_results[[i]] <- results_df
}

# combine results 
gam_results <- do.call(rbind, gam_results)


#Filter for taxa where m_full best fit
gam_filter <- gam_results %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2)
length(gam_filter) # 36 taxa 
  
# filter abundance
smooth_abund <- negtemp_abund %>% filter(Taxa %in% gam_filter)
length(unique(smooth_abund$Taxa))










```



```{r}

t <- lvl2_abund2 %>% filter(lvl2 == "09142 Cell motility")


m_full <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = t)
    
m_main <- gam(clr_abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = t)

AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)   

summary(m_full)

plot(m_full, select = 1, rug = TRUE, shade = TRUE)

```
















# extra code

```{r}

 
t <- lvl2_abund %>% filter(lvl2 == '09122 Translation' | lvl2 == "09121 Transcription")

t <- lvl2_abund %>% filter(lvl2 == "09191 Unclassified: metabolism")
    

m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
               data = t)

m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
               data = t)





```


