---
title: "2025-03-11 shotgun analysis"
author: "Lucore"
date: "2025-03-11"
output: html_document
---


Overall notes: There are a ton of things in this script. essentially it goes through testing bot the KO's and the pathways on different hiararchical levesl. The community level data is congruent for both, PCA's show an effect and so does total obs number but shannon does not. The DAA data is a different story. For the KO's essentially everything is related to temperature (on both lvl 2 and lvl3) and this is true for the interaction effect. Since everything is related, you can't really say much about effects of temp. Also the CI's are huge so you can't say anything about direction of effect. Surpringsingly, Pathways tell a difference story, only some of them are related to temp and realted to the interaction effect. I tried with individual patheays and agglomerated on lvl2 of the pathays. The CI's for the individual pathways are unsprisingly more reliable. Further, then I tried the GAM's on the individual pathways and it's shoing two main clusters, one that is steady until 18degress and then declines and one that increases at 28 decres. Indiation of shift in metabolic profile with temp. This data is pretty overwhelming but I need to look at other papers and how they have shown these results. I think the main difference between the patheays and the KO's is that the patheays are more speciifc, they've taken the genes and clustered them into the different functional groups whereas the grouping of the KO's into functional groups is less precise. I just added up all the KO's associated with one level, much more error involved I think. I am going to move just the patheay analysis to a new sheet becaue that is what we are going to move forward with. 

NEW DISCOVERY: PATHWAYS are showing a relationship with temp unlike KO's. 

NOTES ON KO's
thinking about giving up on this since the DAA isn't showing anything with much certainty except that temp matters, but it's not really showing in what direction it matters, I think I need to use the stratefied data to provide some functional context for the interaction effect and that would be a more efficient way to use the data since I already have so many 16s result. (update: little overlapping taxa in stratefieid data so can't do this).

Could add in some very general conclusions from the results, unlike taxa alpha diversity, functional shannon diversity does not seem to be affected by temperature. However, there is strong evidence that temperature can affect the abundnace of functional pathways across diverse groups (all lvl2's fit bettr with temp in the model and only 64 lvl3 where the temp model does not fit best) spans numerous functions. Importance of temperature for the microbial profile of the gut but more research neded to understand specific effects. 

Have further refined this to include only samples that have have a neopterin value and ran the interacion model. Found very similar results to the temp model, that the interaction model fits the best for all of the lvl2 and most of the lvl3 but not suire about the direction or magnitude of the effect. Unclear whether I should pick out the ones with reliable CI's and focus on those? 

Remember: 

-Have decided to use the uniref 50 data based on decisoins from the 2025-02-27_unmapped_reads scripts. We are also only going to work with the KO's for now since we have determined that the databases are redundant with one another, you can't just combine them because the points aren't independent. 

-GAM's aren't saying much here, most of the model weights are evenly split between the model with the temp spline and the model without the temp spline. Can find this analysis in 2025-03-11_shotgun_analysis

# 1. Load environment

```{r}

library(tidyverse)
library(phyloseq)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(zCompositions)
library(jsonlite)
library(file2meco)

source("functions/my_scale.R")

source("functions/inverse_scale.R")

source("functions/kegg_parse_json.R")

conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)


ko_physeq_u50 <- readRDS("phyloseq_objects/ko_phylo_u50.rds")
path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

```


# 2. Data set up

## a. Filter metadata

decided to use the 71 samples that have feed data and neopterin associated with them because I felt like I couldn't run the models without fruit since there is such a strong relationship between fruit consumption and temperature

```{r}

m <- data.frame(sample_data(ko_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

m_filtered <- m[!is.na(m$fruit), ] # remove samples with no diet data, drops 8 samples

m_filtered <- m_filtered[!is.na(m_filtered$neo), ] # down to 71 samples


# filter phyloseq objects to remove samples with no diet data 
ko_physeq2 <- prune_samples(m_filtered$sample, ko_physeq_u50) # filter physeq object
ko_physeq2 <- prune_taxa(taxa_sums(ko_physeq2) > 0, ko_physeq2) # remove ASV's with zero abundance

path_physeq2 <- prune_samples(m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

## b. Prevalence filtering

hold off on prev filtering for now... our filter for the 16s data "remove features that are not present in at least 0.5% of total samples" doesn't remove anything here because that filter that the feature needs to be in at least one sample for this data. 

Unapped values represent the total abundance of reads that failed to align to any known gene sequences and ungrouped values represent reads that mapped to known gene sequences but could not be confidently assigned to a specific functional category.

We are going to remove the UNMAPPED and UNGROUPED reads so that we can just compared the mapped reads for each sample. Removing the UNGROUPED reads because we can think of this as taxa with no phylum assignment. 
Need to think about removing KO's that don't have any functional assignment (similar to the argument of removing ASV's with no phylum assignment)

```{r}

####### remove UNMAPPED and UNGROUPED reads #######

ko_filtered <- subset_taxa(ko_physeq2, 
                                !(grepl("UNMAPPED", gene_ko) | grepl("UNGROUPED", gene_ko)))

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

## c. alpha diversity

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

ignore the warning, it is upset because the data is not in counts. Have to calculate the observed metric manually because `estimate_richness()` only takes count data

```{r}

ko_shannon <- estimate_richness(ko_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(ko_shannon = Shannon) 

ko_alpha <- as.data.frame(colSums(otu_table(ko_filtered) > 0)) %>%
  rename(ko_observed = 'colSums(otu_table(ko_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(ko_shannon, by = "sample")

hist(ko_alpha$ko_shannon) 
hist(ko_alpha$ko_observed) 

path_shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(path_shannon = Shannon) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(path_observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(path_shannon, by = "sample")

hist(path_alpha$path_shannon) 
hist(path_alpha$path_observed)

m1 <- m_filtered %>% # add to metadata
  left_join(ko_alpha, by = "sample") %>%
  left_join(path_alpha, by = "sample")

```

## d. CLR transformation

doing the CLR transformation on the RPK abundances 

```{r}

ko_table <- as.matrix(t(otu_table(ko_filtered)))
ko_table[1:5,1:5] # rows are samples 

ko_czm <- cmultRepl(ko_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
ko_czm[1:5,1:5] # rows are samples 

ko_clr <- t(apply(ko_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
ko_clr[1:5,1:5] # rows are samples 

ko_pca <- prcomp(ko_clr) # calc pca
plot(ko_pca$x[,1], ko_pca$x[,2]) # plot 

ko_d.mvar <- sum(ko_pca$sdev^2) # total variance
ko_PC1 <- paste("PC1 ","(",round(sum(ko_pca$sdev[1]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC1 30.8%
ko_PC2 <- paste("PC2 ","(",round(sum(ko_pca$sdev[2]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC2 12.5%


path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 

path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 24.4%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 10.9%

######### Add PCA values to metadata #########

ko_df<-data.frame(ko_pca$x)
ko_df$sample<-rownames(ko_df)
ko_df_small <- ko_df %>% 
  select(c(1:4,72)) %>%
  rename_with(~paste0("ko_", .), starts_with("PC"))


path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:4,72)) %>%
  rename_with(~paste0("pathway_", .), starts_with("PC"))

# add to metadata
m2 <- m1 %>%
  left_join(ko_df_small, by = "sample") %>%
  left_join(path_df_small, by = "sample")

```

## f. prep data for modeling

scale variables and make categorical variables factors

```{r}

table(m2$ID)

# scale variables
m_model <- m2 %>%
  mutate(across(c(age, temp, rain, fruit, neo, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```


# 3. Community features analysis

## a. model community features

everything seems normally distributed enough to model with Gaussian except for the Shannon index of the eggnog families which is pretty right skewed but I'm going to move forward with the normal distribution for simplicity of these models and then I can come back to tweak this later if I think it is necessary. 

Going to need to think about how the observed effect size is on a totally difference scale than the shannon diversity effect, which is makeing all of the other effects look quire small

main results here: for the functional metrics, it does not seem like temperature is related to shannon's diversity but it is realted to the PC's and observed gene families. Why is it not related to alpha diversity? Even though taxonomic shannon diversity may decrease at high temperatures, functional alpha diversity may be more resilient to this because many taxa have similar genes and similar functions? However, we don't see the same thing for observed gene families, could speak more to that there is still a decrease in number of functional annotations but the evenness is relatively stable in the community  

```{r}

####### histogranms ########

hist(m_model$ko_shannon) 
hist(m_model$ko_observed) 
hist(m_model$ko_PC1)
hist(m_model$ko_PC2)
hist(m_model$ko_PC3)
hist(m_model$ko_PC4)

hist(m_model$path_shannon) 
hist(m_model$path_observed) 
hist(m_model$pathway_PC1)
hist(m_model$pathway_PC2)
hist(m_model$pathway_PC3)
hist(m_model$pathway_PC4)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15:25)] # list of features to model


for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)

```

## b. plot results

```{r}

#### plot ####

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) #%>%
  #filter(metric != "nog_observed") %>%
  #filter(metric != "ko_observed")
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long <- com_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "PC1", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "PC1", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  

```

## c. loop for interaction effect

```{r}

com_coef_int <- list() 
com_aic_int <- list()

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[10,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[10,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic_int[[i]] <- aic_df
    com_coef_int[[i]] <- results_df

}

# combine results 
com_aic_int <- do.call(rbind, com_aic_int)
com_coef_int <- do.call(rbind, com_coef_int)

```

## d. plot interaction effect

same resutls as temp results, full model fitx best for everything except shannon

```{r}

#### plot ####

plot_com_coef_int <- com_coef_int %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) #%>%
  #filter(metric != "nog_observed") %>%
  #filter(metric != "ko_observed")
         
p1 <- ggplot(plot_com_coef_int, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long_int <- com_aic_int %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long_int, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "PC1", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "PC1", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  


```

# 4. Data set up for DAA

## a. Agglomeration at brite levels

Agglomerate KO's at Brite level 2 and Pathway level from the Keg orthology database. Below are links of where to find the explanation of the database and the link to download the json file. Then the link to the github page with the code to parse the json file. 

website that explains KEG orthology database https://www.genome.jp/kegg/ko.html
json download for KEG orthology assignment: https://www.genome.jp/kegg-bin/get_htext?ko00001.keg
github page with function to clean json: https://gist.github.com/cmkobel/c2f76359eb17e0c00ddd4e83e90bdf99

```{r}

brite <- fromJSON("data/brite_map.json", flatten = F)

brite_parsed <- kegg_parse_json(brite) %>%
  rename(lvl1 = a_class,
         lvl2 = b_class, 
         lvl3 = pathway)

```


# 5. LVL 2 WORKFLOW

results: seems like there is a small negative effect for organismal systems and cellular processes. Genetic information seems to trend positive but metbolisms is all over the place, both negative and positive effects. The CI's are huge and even though the models fit better with the temp model I don't think we can conclude much from this 

results: for every lvl2 category the temp model fits better, over 85% of the model weight. 35 out of 16 cetgroies have a negative coefficient but none have reliable confidence intervals. Seem like termperature is important for abundance of functional genes but we don't have a lot of certainty about the direction. Still, the maximum likelihood is the best estimate and the majority of the effects are negative so it seems like there is a negative effect on both taxa abundances and abundances of functional genes 

############# lvl2 workflow ######################

## A. agglomerate levels

```{r}

brite_level2 <- brite_parsed[,c(2,5)]
length(unique(brite_level2$lvl2)) # 57 unique brite level 2

# agglomerate at brite level 2
lvl2_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_level2, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl2 = replace_na(lvl2, "no_assignment"))

length(unique(lvl2_abund_ko$lvl2)) # 55 lvl2, 1 no_assignment

unique_lvl2 <- unique(lvl2_abund_ko$lvl2)
missing_lvl2 <- brite_level2[!(brite_level2$lvl2 %in% unique_lvl2), ]
unique(missing_lvl2$lvl2) # missing two lvl2 in our dataset

lvl2_abund <- lvl2_abund_ko %>%
  group_by(sample, lvl2) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl2") 

unique_lvl2 <- length(unique(lvl2_abund$lvl2))
unique_samples <- length(unique(lvl2_abund$sample))
(unique_lvl2 * unique_samples) == nrow(lvl2_abund) # check that each sample has a value for all lvls

```

## B. Abundance filtering

Remove categories that you are not interested in including lvls that are not assigned or poorly characterized and human diseases

```{r}

# extract and adjust lvl 1 categories

lvl1 <- brite_parsed %>% select(lvl1, lvl2) %>%
  distinct() %>%
  filter(lvl2 != "09194 Poorly characterized" &
           lvl2 != "no_assignment") %>%
  mutate(lvl1 = case_when(lvl2 == "09192 Unclassified: genetic information processing" ~ "09120 Genetic Information Processing",
                          lvl2 == "09193 Unclassified: signaling and cellular processes" ~ "09140 Cellular Processes",
                          lvl2 == "09191 Unclassified: metabolism" ~ "09100 Metabolism",
                          TRUE ~ lvl1)) # remove the unclassified lvl1 for simplicity
  

length(unique(lvl2_abund$lvl2)) # 56 starting 

lvl2_abund <- lvl2_abund %>% left_join(lvl1, by = "lvl2") %>%
  filter(lvl1 != "09160 Human Diseases") %>%
  filter(lvl2 != "09194 Poorly characterized" &
           lvl2 != "no_assignment") 

length(unique(lvl2_abund$lvl2)) # 42, removed 14 lvl2's

```


## C. Scale variables

have to do this manually because scale function not working within the tidyverse mutate() command for some reason

```{r}

# Scale variables

lvl2_abund$temp_s <-  my_scale(lvl2_abund$temp) 
lvl2_abund$rain_s <-  my_scale(lvl2_abund$rain)
lvl2_abund$age_s <-  my_scale(lvl2_abund$age)
lvl2_abund$fruit_s <-  my_scale(lvl2_abund$fruit)
lvl2_abund$neo_s <-  my_scale(lvl2_abund$neo)

lvl2_abund <- lvl2_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex)) 

length(unique(lvl2_abund$lvl2)) 

```


## D. run loop

Had to change the way I track convergence errors for this loop because I'm using lmer insteatd of glmer. 

two don't converge 

```{r}

coef <- list()
aic <- list()
model_results <- list()

for (i in unique(lvl2_abund$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl2_abund, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl2 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    aic[[i]] <- aic_df
    coef[[i]] <- results_df
    model_results[[i]] <- model_results_df

  })
}

# combine results
aic <- do.call(rbind, aic)
coef <- do.call(rbind, coef)
model_results <- do.call(rbind, model_results)


```

## D. select taxa

of the 38 final lvl2 groups, the temp model fits best for ALL of them

```{r}
# Remove taxa with convergence warnings 
convergence_warnings <- unique(aic$lvl2[aic$Warning == "Convergence Warning"]) # 2 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 100) %>%
  pull(lvl2) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 5 do not converge


# remove error taxa
lvl2_abund2 <- lvl2_abund %>% filter(!lvl2 %in% errors)
length(unique(lvl2_abund2$lvl2)) # 38 remaining 

coef2 <- coef %>% filter(!lvl2 %in% errors)
length(coef2$lvl2)

aic2 <- aic %>% filter(!lvl2 %in% errors)
length(aic2$lvl2) # 76/2 = 38

# filter for taxa where full model fits best
temp_filter <- aic2 %>%
  group_by(lvl2) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2) 
length(unique(temp_filter)) # temp model fits best for all

```

## E. plot loop

loop for coefficient plots by lvl1 catgeory

results: seems like there is a small negative effect for organismal systems and cellular processes. Genetic information seems to trend positive but metbolisms is all over the place, both negative and positive effects. The CI's are huge and even though the models fit better with the temp model I don't think we can conclude much from this 

```{r}

plot_coef <- coef2 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl2 = reorder(lvl2, -coef)) %>%
  left_join(lvl1, by = "lvl2")

# plot ditribution 
pdf("figures/lvl2_KO_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl1)) {
  
  data <- subset(plot_coef, lvl1 == i)
  
  p <- ggplot(data, aes(x = lvl2, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()



```

## F. Interaction effect

```{r}

coef_int <- list()
aic_int <- list()
model_results_int <- list()

for (i in unique(lvl2_abund2$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl2_abund, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl2 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    aic_int[[i]] <- aic_df
    coef_int[[i]] <- results_df
    model_results_int[[i]] <- model_results_df

  })
}

# combine results
aic_int <- do.call(rbind, aic_int)
coef_int <- do.call(rbind, coef_int)
model_results_int <- do.call(rbind, model_results_int)

```


## G. select interaction groups

Interaction model fits best for all lvl 2's

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(aic_int$lvl2[aic_int$Warning == "Convergence Warning"]) # 0 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results_int %>% 
  filter(StdError > 100) %>%
  pull(lvl2) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 4 do not converge


# remove error taxa
lvl2_abund3 <- lvl2_abund2 %>% filter(!lvl2 %in% errors)
length(unique(lvl2_abund3$lvl2)) # 33 remaining 

coef_int2 <- coef_int %>% filter(!lvl2 %in% errors)
length(coef_int2$lvl2)

aic_int2 <- aic_int %>% filter(!lvl2 %in% errors)
length(aic_int2$lvl2) # 76/2 = 38

# filter for taxa where full model fits best
temp_filter <- aic_int2 %>%
  group_by(lvl2) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2) 
length(unique(temp_filter)) # interaction model fits best for all


```


## H Plot interaction

Similar to temp modeling, super wide CI's, not a lot of certainty of effect

```{r}

plot_coef_int <- coef_int2 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl2 = reorder(lvl2, -coef)) %>%
  left_join(lvl1, by = "lvl2")

# plot ditribution 
pdf("figures/lvl2_KO_interaction_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef_int$lvl1)) {
  
  data <- subset(plot_coef_int, lvl1 == i)
  
  p <- ggplot(data, aes(x = lvl2, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()


```


# 6. LVL 3 WORKFLOW

## A. agglomerate levels

```{r}

# pathways (brite level 3)
brite_path <- brite_parsed[,c(3,5)]
length(unique(brite_path$lvl3)) # 561 unique pathways 

#agglomerate at pathway
lvl3_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_path, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl3 = replace_na(lvl3, "no_assignment"))

length(unique(lvl3_abund_ko$lvl3)) # 511 total, 510 lvl3, 1 no_assignment

unique_path <- unique(lvl3_abund_ko$lvl3)
missing_path <- brite_path[!(brite_path$lvl3 %in% unique_path), ]
length(unique(missing_path$lvl3)) # missing 51 pathways in our dataset

lvl3_abund <- lvl3_abund_ko %>%
  group_by(sample, lvl3) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl3")

unique_path <- length(unique(lvl3_abund$lvl3))
unique_samples <- length(unique(lvl3_abund$sample))
(unique_path * unique_samples) == nrow(lvl3_abund) # check that each sample has a value for all lvls

```

## B. Abundance filtering

Remove categories that you are not interested in including lvls that are not assigned or poorly characterized and human diseases

```{r}

# extract and adjust lvl 1 categories

lvl1 <- brite_parsed %>% select(lvl1, lvl2, lvl3) %>%
  distinct() 

length(unique(lvl3_abund$lvl3)) # 526 starting 

lvl3_abund <- lvl3_abund %>% left_join(lvl1, by = "lvl3") %>%
  filter(lvl1 != "09160 Human Diseases") %>%
  filter(lvl3 != "no_assignment" &
           lvl3 != "99997 Function unknown" &
           lvl3 != "99999 Others") 

length(unique(lvl3_abund$lvl3)) # 430, removed 96 lvl2's

```

## C. Scale variables

have to do this manually because scale function not working within the tidyverse mutate() command for some reason

```{r}

# Scale variables

lvl3_abund$temp_s <-  my_scale(lvl3_abund$temp) 
lvl3_abund$rain_s <-  my_scale(lvl3_abund$rain)
lvl3_abund$age_s <-  my_scale(lvl3_abund$age)
lvl3_abund$fruit_s <-  my_scale(lvl3_abund$fruit)
lvl3_abund$neo_s <-  my_scale(lvl3_abund$neo)

lvl3_abund <- lvl3_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex)) 

length(unique(lvl3_abund$lvl3)) 

```


## D. run loop

Had to change the way I track convergence errors for this loop because I'm using lmer insteatd of glmer. 

```{r}

coef <- list()
aic <- list()
model_results <- list()

for (i in unique(lvl3_abund$lvl3)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl3_abund, lvl3 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl3 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl3 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl3 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    aic[[i]] <- aic_df
    coef[[i]] <- results_df
    model_results[[i]] <- model_results_df

  })
}

# combine results
aic <- do.call(rbind, aic)
coef <- do.call(rbind, coef)
model_results <- do.call(rbind, model_results)


```

## E. select taxa

temp model fits best for 342 of them (52 where temp model doesn't fit best)

```{r}
# Remove taxa with convergence warnings 
convergence_warnings <- unique(aic$lvl3[aic$Warning == "Convergence Warning"]) # 35 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 100) %>%
  pull(lvl3) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 36 do not converge

# remove error taxa
lvl3_abund2 <- lvl3_abund %>% filter(!lvl3 %in% errors)
length(unique(lvl3_abund2$lvl3)) # 394 remaining 

coef2 <- coef %>% filter(!lvl3 %in% errors)
length(coef2$lvl3)

aic2 <- aic %>% filter(!lvl3 %in% errors)
length(aic2$lvl3) 

# filter for taxa where full model fits best
temp_filter <- aic2 %>%
  group_by(lvl3) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl3) 
length(unique(temp_filter)) # temp model fits best for 342 

# filter df's for where temp model fits best 
lvl3_abund3 <- lvl3_abund2 %>% filter(lvl3 %in% temp_filter)
length(unique(lvl3_abund3$lvl3))  

coef3 <- coef2 %>% filter(lvl3 %in% temp_filter)
length(coef3$lvl3)

aic3 <- aic2 %>% filter(lvl3 %in% temp_filter)
length(aic3$lvl3)

```

## F. coef plot loop

loop for coefficient plots by lvl3 catgeory

results: moST OF THE 50% CI's cross zero and their doesn't seem to be a large trend by either lvl1 or lvl2 groups

```{r}

plot_coef <- coef3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl3 = reorder(lvl3, -coef)) %>%
  left_join(lvl1, by = "lvl3")

table(plot_coef$lvl2)

# plot ditribution 
pdf("figures/lvl3_KO_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl2)) {
  
  data <- subset(plot_coef, lvl2 == i)
  
  p <- ggplot(data, aes(x = lvl3, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 2)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```


## F. Interaction effect

```{r}

coef_int <- list()
aic_int <- list()
model_results_int <- list()

for (i in unique(lvl3_abund3$lvl3)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(lvl3_abund3, lvl3 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl3 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl3 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl3 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    aic_int[[i]] <- aic_df
    coef_int[[i]] <- results_df
    model_results_int[[i]] <- model_results_df

  })
}

# combine results
aic_int <- do.call(rbind, aic_int)
coef_int <- do.call(rbind, coef_int)
model_results_int <- do.call(rbind, model_results_int)

```


## G. select interaction groups

only 11 where interaction model does not fit best

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(aic_int$lvl3[aic_int$Warning == "Convergence Warning"]) # 4 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results_int %>% 
  filter(StdError > 100) %>%
  pull(lvl3) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 5 do not converge


# remove error taxa
lvl3_abund4 <- lvl3_abund3 %>% filter(!lvl3 %in% errors)
length(unique(lvl3_abund4$lvl3)) # 337 remaining 

coef_int2 <- coef_int %>% filter(!lvl3 %in% errors)
length(coef_int2$lvl3)

aic_int2 <- aic_int %>% filter(!lvl3 %in% errors)
length(aic_int2$lvl3) 

# filter for taxa where full model fits best
temp_filter <- aic_int2 %>%
  group_by(lvl3) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl3) 
length(unique(temp_filter)) # 326 interaction model fits best for all


```


## H Plot interaction

Similar to temp modeling, super wide CI's, not a lot of certainty of effect

```{r}

plot_coef_int <- coef_int2 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl3 = reorder(lvl3, -coef)) %>%
  left_join(lvl1, by = "lvl3")

# plot ditribution 
pdf("figures/lvl3_KO_interaction_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef_int$lvl2)) {
  
  data <- subset(plot_coef_int, lvl2 == i)
  
  p <- ggplot(data, aes(x = lvl3, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()


```


# 7. PATHWAY WORKFLOW


## A. Prep data

ARG-POLYAMINE-SYN does not have level assignments 

```{r}

# wrangle pathway levels

pathway_levels <- as.data.frame(MetaCyc_pathway_map) %>%
  rownames_to_column("path_id") %>%
  rename(lvl1 = Superclass1,
         lvl2 = Superclass2) %>%
  select(lvl1, lvl2, pathway, path_id)

# wrangle clr abundances

path_abund <- path_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "pathway", value = "clr_abundance", -sample) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(pathway = str_remove(pathway, "^X")) %>% 
  separate(pathway, into = c("path_id", "pathway_name"), sep = "\\.\\.", extra = "merge") %>%
  mutate(path_id = str_replace_all(path_id, "\\.", "-")) %>%
  left_join(pathway_levels, by = "path_id") %>%
  rename(path_name = pathway) %>%
  select(sample, path_id, path_name, lvl1, lvl2, clr_abundance:reads)
  

length(unique(path_abund$path_id)) # 548 pathways
table(path_abund$sample) # each sample has 548 instances

# scale variables

path_abund$temp_s <-  my_scale(path_abund$temp) 
path_abund$rain_s <-  my_scale(path_abund$rain)
path_abund$age_s <-  my_scale(path_abund$age)
path_abund$fruit_s <-  my_scale(path_abund$fruit)
path_abund$neo_s <-  my_scale(path_abund$neo)

path_abund <- path_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex),
         group = as.factor(group))


```

## B. run loop

```{r}

path_coef <- list()
path_aic <- list()
path_model_results <- list()

for (i in unique(path_abund$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df
    path_model_results[[i]] <- model_results_df

  })
}

# combine results
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)
path_model_results <- do.call(rbind, path_model_results)


```

## C. select taxa

165 do not converge so we tested 373 total (started with 538)

Temp model fits best for 169 pathways which means that it doesn't fit best for 214... different from the KO's, might actually have some functional significance here

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic$path_id[path_aic$Warning == "Convergence Warning"]) # 165 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no huge SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 165 do not converge

# remove error taxa
path_abund2 <- path_abund %>% filter(!path_id %in% errors)
length(unique(path_abund2$path_id)) # 383 remaining 

path_coef2 <- path_coef %>% filter(!path_id %in% errors)
length(path_coef2$path_id)

path_aic2 <- path_aic %>% filter(!path_id %in% errors)
length(path_aic2$path_id) 

# filter for taxa where full model fits best
temp_filter <- path_aic2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(temp_filter)) # temp model fits best for 169 pathways

# filter df's for where temp model fits best 
path_abund3 <- path_abund2 %>% filter(path_id %in% temp_filter)
length(unique(path_abund3$path_id))  

path_coef3 <- path_coef2 %>% filter(path_id %in% temp_filter)
length(path_coef3$path_id)

path_aic3 <- path_aic2 %>% filter(path_id %in% temp_filter)
length(path_aic3$path_id)

```

## D. coef plot loop

```{r}

plot_coef <- path_coef3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef)) %>%
  left_join(pathway_levels, by = "path_id")

# plot ditribution 
pdf("figures/pathway_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl2)) {
  
  data <- subset(plot_coef, lvl2 == i)
  
  p <- ggplot(data, aes(x = pathway, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 4)) {
  
  batch_plots <- plot_list[i:min(i + 4, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

## GAM's 

19 pathways with non-linear relationship... 

```{r}

gam_results <- list()

for (i in unique(path_abund3$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)
    
    m_main <- gam(clr_abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)

  }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # Update the warning message
      } 
    })

    # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      #create data frame
      results_df <- data.frame(
        path_id = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE
      )
  
      # store results
      gam_results[[i]] <- results_df
}

# combine results 
gam_results <- do.call(rbind, gam_results)


#Filter for taxa where m_full best fit
gam_filter <- gam_results %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id)
length(gam_filter) # 19 
  
# filter abundance
smooth_abund <- path_abund3 %>% filter(path_id %in% gam_filter) %>%
  filter(path_id != "NA")
length(unique(smooth_abund$path_id)) 

```


## c. Generate GAM preds


```{r}

# Initialize an empty list to store results for all taxa
gam_preds <- list()

# Define the range of temp_s values for prediction
temp_seq <- seq(from = min(smooth_abund$temp_s, na.rm = TRUE), 
                to = max(smooth_abund$temp_s, na.rm = TRUE), length.out = 100)


# Iterate over unique values in final_temp_abund$path_id
for (i in unique(smooth_abund$path_id)) {
    
    # Subset your data based on current value of i
    data <- droplevels(subset(smooth_abund, path_id == i))
    
    # run model 
    gam_m <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex, data = data)
    
    # Generate predictions over the specified range of temp_s
    pred_data <- data.frame(
      age_s = mean(data$age_s, na.rm = TRUE),
      temp_s = temp_seq,
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = factor("Tenori", levels = levels(data$group)), # assuming group is a factor
      sex = "M" # assuming sex is a factor
    )
    
    # Predict abundance values
    predicted_values <- predict(gam_m, newdata = pred_data)
    
    # Store the predicted values along with temperature sequence and Taxa label
    df <- data.frame(
      path_id = i,
      temp_s = temp_seq,
      preds = predicted_values
    )
    
    # Store the results in the list
    gam_preds[[i]] <- df
    
}

# Combine all results into a single dataframe
gam_preds <- do.call(rbind, gam_preds)
row.names(gam_preds) <- NULL # clean up row names

length(unique(gam_preds$path_id))
```

## GAM cluster

```{r}

# Create a wide format dataframe
predicted_wide <- reshape(gam_preds, idvar = "temp_s", timevar = "path_id", direction = "wide")

# clean data
predicted_wide <- predicted_wide %>% 
  rename_with(~ str_replace(., "^preds.", ""), everything()) %>% # clean column names 
  select(-temp_s) # remove the temp_s column for correlation calculation and positive taxa

# elbow method
clust_plot <- fviz_nbclust(predicted_wide, kmeans, method = "wss", k.max = 10) + 
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank()) +
  labs(x = "Number of clusters", y = "Within-cluster sum of squares", title = "")
clust_plot

# set number of clusters
k=4

# distnce matrix
clus.distance = as.dist(1 - cor(predicted_wide, use = "pa")) # create distance matrix

h.clust = hclust(clus.distance,method='complete') # calculate clusters 

dendogram <- plot(h.clust, cex = 0.6, main = "", xlab = "", ylab = "") # plot dendogram
rect.hclust(h.clust, k = k, border = 2:6) # add clusters 


# Separate temp-relationships into k clusters 
clusters <- cutree(h.clust, k = k)

# Add cluster assignments to the original data
gam_preds$Cluster <- clusters[match(gam_preds$path_id, names(clusters))]

### Plot clusters ###

# clean data
clean_preds <- gam_preds %>% mutate(Cluster = as.factor(Cluster))

# by cluster 
c1 <- clean_preds %>% filter(Cluster == 1)
c2 <- clean_preds %>% filter(Cluster == 2)
c3 <- clean_preds %>% filter(Cluster == 3)
c4 <- clean_preds %>% filter(Cluster == 4)

p1 <- ggplot(c1, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p1

p2 <- ggplot(c2, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p2

p3 <- ggplot(c3, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p3

p4 <- ggplot(c4, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p4

# Arrange the plots in a 2x2 layout
cluster_plot <- ggarrange(p1, p2,
                           ncol = 1, nrow = 2,
                          labels = c("A", "B"))
```


# plot two main clusters 

```{r}

mean_temp <- mean(smooth_abund$temp)
sd_temp <- sd(smooth_abund$temp) 

cluster1 <- ggplot(c1, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal() +
  labs(y = "Predicted log abundance", x = "Temperature (°C)") +
  scale_x_continuous(
   breaks = seq(min(c1$temp_s), max(c1$temp_s), length.out = 6), 
   labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 16, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 

cluster2 <- ggplot(c2, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal() +
  labs(y = "Predicted log abundance", x = "Temperature (°C)") +
  scale_x_continuous(
   breaks = seq(min(c1$temp_s), max(c1$temp_s), length.out = 6), 
   labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 16, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 

```















## E. interaction effect

```{r}

path_coef_int <- list()
path_aic_int <- list()
path_model_results_int <- list()

for (i in unique(path_abund3$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic_int[[i]] <- aic_df
    path_coef_int[[i]] <- results_df
    path_model_results_int[[i]] <- model_results_df

  })
}

# combine results
path_aic_int <- do.call(rbind, path_aic_int)
path_coef_int <- do.call(rbind, path_coef_int)
path_model_results_int <- do.call(rbind, path_model_results_int)

```


## F. select interaction taxa

tested 169, 53 of those didn't converge so tested 116

out of 116 the interaction model first best for 65 of them 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic_int$path_id[path_aic_int$Warning == "Convergence Warning"]) # 53 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results_int %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no big SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 53 do not converge


# remove error taxa
path_abund4 <- path_abund3 %>% filter(!path_id %in% errors)
length(unique(path_abund4$path_id)) # 116 remaining 

path_coef_int2 <- path_coef_int %>% filter(!path_id%in% errors)
length(path_coef_int2$path_id)

path_aic_int2 <- path_aic_int %>% filter(!path_id %in% errors)
length(path_aic_int2$path_id) 

# filter for taxa where full model fits best
temp_filter <- path_aic_int2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(temp_filter)) # 65 interaction model fits best for all

# filter df's for where temp model fits best 
path_abund5 <- path_abund4 %>% filter(path_id %in% temp_filter)
length(unique(path_abund5$path_id))  

path_coef_int3 <- path_coef_int2 %>% filter(path_id %in% temp_filter)
length(path_coef_int3$path_id)

path_aic_int2 <- path_aic_int2 %>% filter(path_id %in% temp_filter)
length(path_aic_int2$path_id)

```


## G. plot interaction

```{r}

plot_coef <- path_coef_int3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef)) %>%
  left_join(pathway_levels, by = "path_id")

# plot ditribution 
pdf("figures/pathway_interaction_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl2)) {
  
  data <- subset(plot_coef, lvl2 == i)
  
  p <- ggplot(data, aes(x = pathway, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 4)) {
  
  batch_plots <- plot_list[i:min(i + 4, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```


# 8. Pathways again but at lvl2 hierarchy

## A. data set up

```{r}

# wrangle pathway levels

pathway_levels <- as.data.frame(MetaCyc_pathway_map) %>%
  rownames_to_column("path_id") %>%
  rename(lvl1 = Superclass1,
         lvl2 = Superclass2) %>%
  select(lvl1, lvl2, pathway, path_id)

# wrangle clr abundances

path_abund <- path_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "pathway", value = "clr_abundance", -sample) %>%
  mutate(pathway = str_remove(pathway, "^X")) %>% 
  separate(pathway, into = c("path_id", "pathway_name"), sep = "\\.\\.", extra = "merge") %>%
  mutate(path_id = str_replace_all(path_id, "\\.", "-")) %>%
  left_join(pathway_levels, by = "path_id") %>%
  rename(path_name = pathway) %>%
  select(sample, path_id, path_name, lvl1, lvl2, clr_abundance) %>%
  filter(path_id != "ARG-POLYAMINE-SYN") # remove path with no characterization
  

length(unique(path_abund$path_id)) # 547 pathways
length(unique(path_abund$lvl2)) # 81 lvl2's
table(path_abund$sample) # each sample has 547 instances

path_abund <- path_abund %>%
  group_by(sample, lvl2) %>%
  summarise(clr_abundance = sum(clr_abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  filter(lvl2 != "NA") # remove NA's

length(unique(path_abund$lvl2)) # 80
table(path_abund$sample)

# scale variables

path_abund$temp_s <-  my_scale(path_abund$temp) 
path_abund$rain_s <-  my_scale(path_abund$rain)
path_abund$age_s <-  my_scale(path_abund$age)
path_abund$fruit_s <-  my_scale(path_abund$fruit)
path_abund$neo_s <-  my_scale(path_abund$neo)

path_abund <- path_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex))

```


## B. run loop

```{r}

path_coef <- list()
path_aic <- list()
path_model_results <- list()

for (i in unique(path_abund$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl2 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df
    path_model_results[[i]] <- model_results_df

  })
}

# combine results
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)
path_model_results <- do.call(rbind, path_model_results)


```

## C. select taxa

26 do not converge so we tested 54 total (started with 80)

Temp model fits best for 169 pathways which means that it doesn't fit best for 214... different from the KO's, might actually have some functional significance here

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic$lvl2[path_aic$Warning == "Convergence Warning"]) # 26 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results %>% 
  filter(StdError > 100) %>%
  pull(lvl2) %>%
  unique() # no huge SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 26 do not converge

# remove error taxa
path_abund2 <- path_abund %>% filter(!lvl2 %in% errors)
length(unique(path_abund2$lvl2)) # 54 remaining 

path_coef2 <- path_coef %>% filter(!lvl2 %in% errors)
length(path_coef2$lvl2)

path_aic2 <- path_aic %>% filter(!lvl2 %in% errors)
length(path_aic2$lvl2) 

# filter for taxa where full model fits best
temp_filter <- path_aic2 %>%
  group_by(lvl2) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2) 
length(unique(temp_filter)) # temp model fits best for 32 pathways

# filter df's for where temp model fits best 
path_abund3 <- path_abund2 %>% filter(lvl2 %in% temp_filter)
length(unique(path_abund3$lvl2))  

path_coef3 <- path_coef2 %>% filter(lvl2 %in% temp_filter)
length(path_coef3$lvl2)

path_aic3 <- path_aic2 %>% filter(lvl2 %in% temp_filter)
length(path_aic3$lvl2)

```

## D. coef plot loop

```{r}

plot_coef <- path_coef3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(lvl2, -coef)) %>%
  left_join(pathway_levels, by = "lvl2")

# plot ditribution 
pdf("figures/pathway_DAA_lvl2.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl1)) {
  
  data <- subset(plot_coef, lvl1 == i)
  
  p <- ggplot(data, aes(x = lvl2, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 1)) {
  
  batch_plots <- plot_list[i:min(i, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

## GAM's 

6 lvl2's that have a non-linear relationship... and no real trend when you plot them. I think I should keep this simple and leave out the gams

```{r}

gam_results <- list()

for (i in unique(path_abund3$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)
    
    m_main <- gam(clr_abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)

  }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # Update the warning message
      } 
    })

    # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      #create data frame
      results_df <- data.frame(
        lvl2 = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE
      )
  
      # store results
      gam_results[[i]] <- results_df
}

# combine results 
gam_results <- do.call(rbind, gam_results)


#Filter for taxa where m_full best fit
gam_filter <- gam_results %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2)
length(gam_filter) # 36 taxa 
  
# filter abundance
smooth_abund <- path_abund3 %>% filter(lvl2 %in% gam_filter)
length(unique(smooth_abund$lvl2))

```


## predict GAM's and plot

no real trend

```{r}

test <- smooth_abund %>% filter(lvl2 == "Secondary Metabolite Biosynthesis")


gam <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = test)

summary(gam)

plot(gam, select = 1, rug = TRUE, shade = TRUE)


```
























## E. interaction effect

```{r}

path_coef_int <- list()
path_aic_int <- list()
path_model_results_int <- list()

for (i in unique(path_abund3$lvl2)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, lvl2 == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      lvl2 = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic_int[[i]] <- aic_df
    path_coef_int[[i]] <- results_df
    path_model_results_int[[i]] <- model_results_df

  })
}

# combine results
path_aic_int <- do.call(rbind, path_aic_int)
path_coef_int <- do.call(rbind, path_coef_int)
path_model_results_int <- do.call(rbind, path_model_results_int)

```


## F. select interaction taxa

tested 28 total, started with 32 and four did not converge 


```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic_int$lvl2[path_aic_int$Warning == "Convergence Warning"]) # 4 do not converge
length(convergence_warnings)


# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results_int %>% 
  filter(StdError > 100) %>%
  pull(lvl2) %>%
  unique() # no big SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 4 do not converge


# remove error taxa
path_abund4 <- path_abund3 %>% filter(!lvl2 %in% errors)
length(unique(path_abund4$lvl2)) # 28 remaining 

path_coef_int2 <- path_coef_int %>% filter(!lvl2%in% errors)
length(path_coef_int2$lvl2)

path_aic_int2 <- path_aic_int %>% filter(!lvl2 %in% errors)
length(path_aic_int2$lvl2) 

# filter for taxa where full model fits best
temp_filter <- path_aic_int2 %>%
  group_by(lvl2) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(lvl2) 
length(unique(temp_filter)) # 25 interaction model fits best for all

# filter df's for where temp model fits best 
path_abund5 <- path_abund4 %>% filter(lvl2 %in% temp_filter)
length(unique(path_abund5$lvl2))  

path_coef_int3 <- path_coef_int2 %>% filter(lvl2 %in% temp_filter)
length(path_coef_int3$lvl2)

path_aic_int2 <- path_aic_int2 %>% filter(lvl2 %in% temp_filter)
length(path_aic_int2$lvl2)

```


## G. plot interaction

```{r}

plot_coef <- path_coef_int3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         lvl2 = reorder(lvl2, -coef)) %>%
  left_join(pathway_levels, by = "lvl2")

# plot ditribution 
pdf("figures/pathway_interaction_DAA_lvl2.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl1)) {
  
  data <- subset(plot_coef, lvl1 == i)
  
  p <- ggplot(data, aes(x = lvl2, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 1)) {
  
  batch_plots <- plot_list[i:min(i, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```
