---
title: "2025-03-11 shotgun analysis"
author: "Lucore"
date: "2025-03-11"
output: html_document
---


Notes: Most up to date analysis for the shotgun data. Have decided to use the uniref 50 data based on decisoins from the 2025-02-27_unmapped_reads scripts. We are also only going to work with the KO's for now since we have determiend that the databases are redunant with one another, you can't just combine them because the points aren't independent. 

Here we are working on the DAA analysis and left this where the warning message won't update because four groups aren't converging but I can't figure out which ones they are. 

Next up is to figure this out and then filter out the ones that haven't converged based on large CI's and then interpret the results. Then think about doing this for the lvl three groups and figuring out a good visualization. 


# 1. Load environment

```{r}

library(tidyverse)
library(phyloseq)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(zCompositions)
library(jsonlite)

source("functions/my_scale.R")

source("functions/inverse_scale.R")

source("functions/kegg_parse_json.R")

conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)


ko_physeq_u50 <- readRDS("phyloseq_objects/ko_phylo_u50.rds")


```


# 2. Data set up

## a. Filter metadata

decided to use the 87 samples that have feed data associated with them because I felt like I couldn't run the models without fruit since there is such a strong relationship between fruit consumption and temperature

```{r}

m <- data.frame(sample_data(ko_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

m_filtered <- m[!is.na(m$fruit), ] # remove samples with no diet data, drops 8 samples

# filter phyloseq objects to remove samples with no diet data 
ko_physeq2 <- prune_samples(m_filtered$sample, ko_physeq_u50) # filter physeq object
ko_physeq2 <- prune_taxa(taxa_sums(ko_physeq2) > 0, ko_physeq2) # remove ASV's with zero abundance

```

## b. Prevalence filtering

hold off on prev filtering for now... our filter for the 16s data "remove features that are not present in at least 0.5% of total samples" doesn't remove anything here because that filter that the feature needs to be in at least one sample for this data. 

Unapped values represent the total abundance of reads that failed to align to any known gene sequences and ungrouped values represent reads that mapped to known gene sequences but could not be confidently assigned to a specific functional category.

We are going to remove the UNMAPPED and UNGROUPED reads so that we can just compared the mapped reads for each sample. Removing the UNGROUPED reads because we can think of this as taxa with no phylum assignment. 
Need to think about removing KO's that don't have any functional assignment (similar to the argument of removing ASV's with no phylum assignment)

```{r}

####### remove UNMAPPED and UNGROUPED reads #######

ko_filtered <- subset_taxa(ko_physeq2, 
                                !(grepl("UNMAPPED", gene_ko) | grepl("UNGROUPED", gene_ko)))

```

## c. alpha diversity

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

ignore the warning, it is upset because the data is not in counts. Have to calculate the observed metric manually because `estimate_richness()` only takes count data

```{r}

ko_shannon <- estimate_richness(ko_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(ko_shannon = Shannon) 

ko_alpha <- as.data.frame(colSums(otu_table(ko_filtered) > 0)) %>%
  rename(ko_observed = 'colSums(otu_table(ko_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(ko_shannon, by = "sample")

hist(ko_alpha$ko_shannon) 
hist(ko_alpha$ko_observed) 

m1 <- m_filtered %>% # add to metadata
  left_join(ko_alpha, by = "sample")

```

## d. CLR transformation

doing the CLR transformation on the RPK abundances 

```{r}

ko_table <- as.matrix(t(otu_table(ko_filtered)))
ko_table[1:5,1:5] # rows are samples 

ko_czm <- cmultRepl(ko_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
ko_czm[1:5,1:5] # rows are samples 

ko_clr <- t(apply(ko_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
ko_clr[1:5,1:5] # rows are samples 

ko_pca <- prcomp(ko_clr) # calc pca
plot(ko_pca$x[,1], ko_pca$x[,2]) # plot 

ko_d.mvar <- sum(ko_pca$sdev^2) # total variance
ko_PC1 <- paste("PC1 ","(",round(sum(ko_pca$sdev[1]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC1 30.8%
ko_PC2 <- paste("PC2 ","(",round(sum(ko_pca$sdev[2]^2)/ko_d.mvar, 3)*100,"%",")",sep="") # PC2 12.5%

######### Add PCA values to metadata #########

ko_df<-data.frame(ko_pca$x)
ko_df$sample<-rownames(ko_df)
ko_df_small <- ko_df %>% 
  select(c(1:4,88)) %>%
  rename_with(~paste0("ko_", .), starts_with("PC"))

# add to metadata
m2 <- m1 %>%
  left_join(ko_df_small, by = "sample") 

```

## f. prep data for modeling

scale variables and make categorical variables factors

```{r}

table(m2$ID)

# scale variables
m_model <- m2 %>%
  mutate(across(c(age, temp, rain, fruit, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```


# 3. Community features analysis

## a. model community features

everything seems normally distributed enough to model with Gaussian except for the Shannon index of the eggnog families which is pretty right skewed but I'm going to move forward with the normal distribution for simplicity of these models and then I can come back to tweak this later if I think it is necessary. 

Going to need to think about how the observed effect size is on a totally difference scale than the shannon diversity effect, which is makeing all of the other effects look quire small

main results here: for the functional metrics, it does not seem like temperature is related to shannon's diversity but it is realted to the PC's and observed gene families. Why is it not related to alpha diversity? Even though taxonomic shannon diversity may decrease at high temperatures, functional alpha diversity may be more resilient to this because many taxa have similar genes and similar functions? However, we don't see the same thing for observed gene families, could speak more to that there is still a decrease in number of functional annotations but the evenness is relatively stable in the community  

```{r}

####### histogranms ########

hist(m_model$ko_shannon) 
hist(m_model$ko_observed) 
hist(m_model$ko_PC1)
hist(m_model$ko_PC2)
hist(m_model$ko_PC3)
hist(m_model$ko_PC4)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15:20)] # list of features to model


for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)

```

## b. plot results

```{r}

#### plot ####

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) #%>%
  #filter(metric != "nog_observed") %>%
  #filter(metric != "ko_observed")
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long <- com_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "PC1", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "PC1", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  


```

# 4. Data set up for DAA

## a. Agglomeration at brite levels

Agglomerate KO's at Brite level 2 and Pathway level from the Keg orthology database. Below are links of where to find the explanation of the database and the link to download the json file. Then the link to the github page with the code to parse the json file. 

websie that explains KEG orthology database https://www.genome.jp/kegg/ko.html
json download for KEG orthology assignment: https://www.genome.jp/kegg-bin/get_htext?ko00001.keg
github page with function to clean json: https://gist.github.com/cmkobel/c2f76359eb17e0c00ddd4e83e90bdf99


```{r}

brite <- fromJSON("data/brite_map.json", flatten = F)

brite_parsed <- kegg_parse_json(brite) %>%
  rename(lvl1 = a_class,
         lvl2 = b_class, 
         lvl3 = pathway)

#################### brite level2 ####################################

brite_level2 <- brite_parsed[,c(2,5)]
length(unique(brite_level2$lvl2)) # 57 unique brite level 2

# agglomerate at brite level 2
lvl2_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_level2, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl2 = replace_na(lvl2, "no_assignment"))

length(unique(lvl2_abund_ko$lvl2)) # 55 lvl2, 1 no_assignment

unique_lvl2 <- unique(lvl2_abund_ko$lvl2)
missing_lvl2 <- brite_level2[!(brite_level2$lvl2 %in% unique_lvl2), ]
unique(missing_lvl2$lvl2) # missing two lvl2 in our dataset

lvl2_abund <- lvl2_abund_ko %>%
  group_by(sample, lvl2) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl2")

unique_lvl2 <- length(unique(lvl2_abund$lvl2))
unique_samples <- length(unique(lvl2_abund$sample))
(unique_lvl2 * unique_samples) == nrow(lvl2_abund) # check that each sample has a value for all lvls


#################### brite level3 ####################################

# pathways (brite level 3)
brite_path <- brite_parsed[,c(3,5)]
length(unique(brite_path$lvl3)) # 561 unique pathways 

#agglomerate at pathway
lvl3_abund_ko <- ko_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "reaction_ko", value = "abundance", -sample) %>%
  left_join(brite_path, by = "reaction_ko", relationship = "many-to-many") %>%
  mutate(lvl3 = replace_na(lvl3, "no_assignment"))

length(unique(lvl3_abund_ko$lvl3)) # 511 total, 510 lvl3, 1 no_assignment

unique_path <- unique(lvl3_abund_ko$lvl3)
missing_path <- brite_path[!(brite_path$lvl3 %in% unique_path), ]
length(unique(missing_path$lvl3)) # missing 51 pathways in our dataset

lvl3_abund <- lvl3_abund_ko %>%
  group_by(sample, lvl3) %>%
  summarise(clr_abundance = sum(abundance)) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(level = "lvl3")

unique_path <- length(unique(lvl3_abund$lvl3))
unique_samples <- length(unique(lvl3_abund$sample))
(unique_path * unique_samples) == nrow(lvl3_abund) # check that each sample has a value for all lvls

```

## b. Plot distribution

distribution of clr abundances looks kind of normal for the level 2 agglomerations, level 3 doesn't look great but we'll proceed with the normal distribution because I think it is the most appropriate for the clr abundances given that there are negative and positive values

```{r}

################## lvl2 ######################

# plot ditribution 
pdf("figures/lvl2_distributions.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(lvl2_abund$lvl2)) {
  
  data <- subset(lvl2_abund, lvl2 == i)
  
  p <- ggplot(data, aes(x = clr_abundance)) +
    geom_histogram(bins = 15) +
    theme_minimal() +
    labs(title = paste("Distribution for", i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()


```

## c. Abundance filtering

not sure how to do this with clr transformed values....... maybe I just shouldn't. Lets move on the the daa of the lvl2's and see how it goes

maybe instead of abundance filtering we only filter for the lvls we are interested in, we should probably remove unknowns and unnasigns, and not well described etc. 

ex. not interested in: 

09190 Not Included in Pathway or Brite


## d. Scale variables

have to do this manually because scale function not working within the tidyverse mutate() command for some reason

```{r}

# Scale variables

lvl2_abund$temp_s <-  my_scale(lvl2_abund$temp) 
lvl2_abund$rain_s <-  my_scale(lvl2_abund$rain)
lvl2_abund$age_s <-  my_scale(lvl2_abund$age)
lvl2_abund$fruit_s <-  my_scale(lvl2_abund$fruit)

lvl2_abund <- lvl2_abund %>%
  mutate(ID = as.factor(ID),
         sex = as.factor(sex)) 

length(unique(lvl2_abund$lvl2)) # 56 unique groups

```


# 5. Temperature abundance testing

my warning message isn't working, four of them aren't converging but the warning message isn't updating to tell me which four aren't converging

```{r}

coef <- list()
aic <- list()
model_results <- list()

# loop
for (i in unique(lvl2_abund$lvl2)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(lvl2_abund, lvl2 == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
               data = data)

      m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
               data = data) 
      
    }, warning = function(w) {
   
      warning_msg <<- conditionMessage(w)
  
      if (grepl("boundary (singular) fit: see help('isSingular')", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
     # extract taxa coef
    coef_value <- round(summary(m_full)$coefficient[7, 1], 4)
    
    # extract se
    std_value <- round(summary(m_full)$coefficient[7, 2], 4)
    
    # calc CI's
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value
      
      #create data frames
    aic_df <- data.frame(
        lvl2 = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
      lvl2 = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE)
    
  options(scipen = 999)  

    model_results_df <- data.frame(
      lvl2 = i,
      Predictor = rownames(summary(m_full)$coefficients)[2:7],
      Coefficient = round(summary(m_full)$coeff[c(2:7), 1], 4),
      StdError = round(summary(m_full)$coeff[c(2:7), 2], 4),
      stringsAsFactors = FALSE)
  
      # save results
    aic[[i]] <- aic_df
    coef[[i]] <- results_df
    model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
aic <- do.call(rbind, aic)
coef <- do.call(rbind, coef)
model_results <- do.call(rbind, model_results)

```







# extra code

```{r}

 
t <- lvl2_abund %>% filter(lvl2 == '09122 Translation' | lvl2 == "09121 Transcription")

t <- lvl2_abund %>% filter(lvl2 == "09121 Transcription")
    

m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
               data = t)

m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
               data = t)



```


