---
title: "Untitled"
author: "Lucore"
date: "2025-05-01"
output: html_document
---

Notes: This is the shotgun analysis that only uses the pathway (metacyc) data. Decided to forgo the KO data because it wasn't showing much (see script 2024-04-28 shotgun analysis). This analysis follows the same steps as the 16s analysis. We use 71 samples that have neo and feeding data associated with them. 

Have decided to use the uniref 50 data based on decisoins from the 2025-02-27_unmapped_reads scripts.

# 1. Load environment

```{r}

library(tidyverse)
library(phyloseq)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(zCompositions)
library(jsonlite)
library(file2meco)

source("functions/my_scale.R")

source("functions/inverse_scale.R")

source("functions/kegg_parse_json.R")

conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)

path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

```

# 2. Data set up

## a. Filter metadata

decided to use the 71 samples that have feed data and neopterin associated with them

```{r}

m <- data.frame(sample_data(path_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

m_filtered <- m[!is.na(m$fruit), ] # remove samples with no diet data, drops 8 samples

m_filtered <- m_filtered[!is.na(m_filtered$neo), ] # down to 71 samples


# filter phyloseq objects to remove samples with no diet data 
path_physeq2 <- prune_samples(m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

## b. Prevalence filtering

Umapped values represent the total abundance of reads that failed to align to any known gene sequences and ungrouped values represent reads that mapped to known gene sequences but could not be confidently assigned to a specific functional category.

Unintegrated values represent genes that did not belong to a known pathway

```{r}

####### remove UNMAPPED and UNGROUPED reads #######

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

## c. alpha diversity

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

ignore the warning, it is upset because the data is not in counts. Have to calculate the observed metric manually because `estimate_richness()` only takes count data, gives you exactly the same values as if you calculated it with vegan

```{r}

path_shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(path_shannon = Shannon) 

path_simpson <- estimate_richness(path_filtered, measures = "Simpson") %>%
  rownames_to_column(var = "sample") %>%
  rename(path_simpson = Simpson) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(path_observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(path_shannon, by = "sample") %>%
  full_join(path_simpson, by = "sample")

hist(path_alpha$path_shannon) 
hist(path_alpha$path_observed)
hist(path_alpha$path_simpson)

m1 <- m_filtered %>% # add to metadata
  left_join(path_alpha, by = "sample")

```

## d. CLR transformation

doing the CLR transformation on the RPK abundances 

```{r}

######### transformation and PCA #########

path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 


path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 21.5%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 13.1%

######### Add PCA values to metadata #########

path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:4,72)) %>%
  rename_with(~paste0("pathway_", .), starts_with("PC"))

# add to metadata
m2 <- m1 %>%
  left_join(path_df_small, by = "sample")

```

## f. prep data for modeling

scale variables and make categorical variables factors

```{r}

table(m2$ID)

# scale variables
m_model <- m2 %>%
  mutate(across(c(age, temp, rain, fruit, neo, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```

# 3. Community features analysis

## a. model community features

normally distributed enough to model with Gaussian (PC1, shannon, and simpson are most concerning)

main results here: based on model weight temp is not related to shannon or simpson diversity but it is related observed paths and to the PC's. When we look at the coefficient plot the CI for observed PC's is wide so we don't have a lot of certainty about the direction of the effect. 

Why isn't temp related to shannon or simpson? - temperature does not substantialy change how evenly distributed patheays are. It is likely that low abundance patheays decrease with rising temperature due to the negative effect of temperature on observed number of pathways, but those decreases are likely in low abundant paths that don't affect the overall distribution of pathway in the community. Influences functional diversity but not the emphasis on priotiy functions. Even though taxonomic shannon diversity may decrease at high temperatures, functional alpha diversity may be more resilient to this because many taxa have similar genes and similar functions? 

```{r}

####### histogranms ########

hist(m_model$path_shannon) 
hist(m_model$path_simpson)
hist(m_model$path_observed) 
hist(m_model$pathway_PC1)
hist(m_model$pathway_PC2)
hist(m_model$pathway_PC3)
hist(m_model$pathway_PC4)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15:21)] # list of features to model

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)

```

## b. plot results

```{r}

#### plot ####

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long <- com_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "path_observed", label = "temp model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "path_observed", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  

```

## c. model interaction effect

```{r}

com_coef_int <- list() 
com_aic_int <- list()

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[10,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[10,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic_int[[i]] <- aic_df
    com_coef_int[[i]] <- results_df

}

# combine results 
com_aic_int <- do.call(rbind, com_aic_int)
com_coef_int <- do.call(rbind, com_coef_int)

```

## d. plot interaction effect

same results as temp results, full model fits best for everything except shannon and simpson

```{r}

#### plot ####

plot_com_coef_int <- com_coef_int %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) #%>%
  #filter(metric != "nog_observed") %>%
  #filter(metric != "ko_observed")
         
p1 <- ggplot(plot_com_coef_int, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "community features") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1


# Reshape data
com_data_long_int <- com_aic_int %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long_int, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.7, y = "path_observed", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "path_observed", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  

```


# 4. Pathway temp DAA

Temp DAA analysis for the individual pathways

## a. Prep data

starting with 547 pathways

```{r}

# wrangle pathway levels
pathway_levels <- as.data.frame(MetaCyc_pathway_map) %>%
  rownames_to_column("path_id") %>%
  rename(lvl1 = Superclass1,
         lvl2 = Superclass2) %>%
  select(lvl1, lvl2, pathway, path_id)

# wrangle clr abundances
path_abund <- path_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "pathway", value = "clr_abundance", -sample) %>%
  left_join(m_filtered, by = "sample") %>%
  mutate(pathway = str_remove(pathway, "^X")) %>% 
  separate(pathway, into = c("path_id", "pathway_name"), sep = "\\.\\.", extra = "merge") %>%
  mutate(path_id = str_replace_all(path_id, "\\.", "-")) %>%
  left_join(pathway_levels, by = "path_id") %>%
  rename(path_name = pathway) %>%
  select(sample, path_id, path_name, lvl1, lvl2, clr_abundance:reads) %>%
  filter(path_id != "ALL-CHORISMATE-PWY") # remove path with no lvl 2 assignment 
  
length(unique(path_abund$path_id)) # 547 pathways
table(path_abund$sample) # each sample has 547 instances

# scale variables

path_abund <- path_abund %>%
  mutate(across(c(age, temp, rain, fruit, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         sex = as.factor(sex),
         ID = as.factor(ID))

```

## b. run temp loop

```{r}

path_coef <- list()
path_aic <- list()
path_model_results <- list()

for (i in unique(path_abund$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df
    path_model_results[[i]] <- model_results_df

  })
}

# combine results
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)
path_model_results <- do.call(rbind, path_model_results)


```

## c. select taxa

164 do not converge so we tested 383 total (started with 547)

Temp model fits best for 169 pathways which means that it doesn't fit best for 214

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic$path_id[path_aic$Warning == "Convergence Warning"]) # 164 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no huge SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 164 do not converge

# remove error taxa
path_abund2 <- path_abund %>% filter(!path_id %in% errors)
length(unique(path_abund2$path_id)) # 383 remaining 

path_coef2 <- path_coef %>% filter(!path_id %in% errors)
length(path_coef2$path_id)

path_aic2 <- path_aic %>% filter(!path_id %in% errors)
length(path_aic2$path_id) 

# filter for taxa where full model fits best
temp_filter <- path_aic2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(temp_filter)) # temp model fits best for 169 pathways

# filter df's for where temp model fits best 
path_abund3 <- path_abund2 %>% filter(path_id %in% temp_filter)
length(unique(path_abund3$path_id))  

path_coef3 <- path_coef2 %>% filter(path_id %in% temp_filter)
length(path_coef3$path_id)

path_aic3 <- path_aic2 %>% filter(path_id %in% temp_filter)
length(path_aic3$path_id)

```

## d. plot figure

```{r}

# data prep
groupings <- read.csv("data/path_group_assignment.csv")

temp_path_info <- path_coef3 %>%
  left_join(pathway_levels, by = "path_id") %>%
  select(path_id, pathway, coef:high_ci_50) %>% 
  left_join(groupings, by = "pathway") %>%
  filter(!is.na(group)) %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef),
         pathway = str_remove_all(pathway, " \\(.*?\\)"),
         pathway = str_replace_all(pathway, "&beta;", "β"),
         pathway = case_when(pathway == "pyruvate fermentation to acetate and-lactate I" ~ "pyruvate fermentation to acetate and (S)-lactate I",
                             pathway == "superpathway of2-lipid A biosynthesis" ~ "superpathway of (Kdo)2-lipid A biosynthesis",
                             pathway == "fatty acid elongation -- saturated" ~ "fatty acid elongation",
                             pathway == "palmitoleate biosynthesis I-dodec-5-enoate)" ~ "palmitoleate biosynthesis I",
                             TRUE ~pathway))

# order for immune groups
subgroup_order <- c("increased anti-inflammatory signals", "decreased inflammatory signals", "decrease anti-inflammatory signals")

immune <- temp_path_info %>% filter(group == "immune regulation") %>% arrange(desc(coef)) %>%
  mutate(subgroup = factor(subgroup, levels = subgroup_order)) %>%
  arrange(subgroup, desc(coef)) %>%
  mutate(pathway = factor(pathway, levels = unique(pathway)))

# make other functional groups
redox <- temp_path_info %>% filter(group == "redox maintenance") %>% arrange(desc(coef)) 
lipid <- temp_path_info %>% filter(group == "lipid catabolism and remodeling") %>% arrange(desc(coef)) 
nucleotide <- temp_path_info %>% filter(group == "nucleotide mobilization") %>% arrange(desc(coef)) 
oxygen <- temp_path_info %>% filter(group == "oxygen-driven metabolism") %>% arrange(desc(coef)) 



# write function
# remmeber to add text size to this

make_coef_plot <- function(data) {
  ggplot(data, aes(x = fct_reorder(pathway, coef, .desc = TRUE), y = coef, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
    geom_point(position = position_dodge(width = 0.3), size = 1) +
    geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                   position = position_dodge(width = 0.3), linewidth = .35) +
    geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                   position = position_dodge(width = 0.3), linewidth = .55) +
    scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
    coord_flip(clip = "off") +
    ylim(-5, 5) +
    labs(x = NULL, y = NULL) +  
    theme(axis.line = element_line(color = "black", linewidth = .28),
          axis.text.y = element_text(size = 8, color = "black"), 
          text = element_text(size = 10, color = "black", family = "sans"), 
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          legend.position = "none",
          panel.background = element_blank(),
          plot.margin = margin(t = 5, r = 110, b = 5, l = 5))
}


# plot
redox_p <- make_coef_plot(redox) +
  ggtitle("A. Redox maintenance") +
  annotate("segment", x = "assimilatory sulfate reduction IV",  xend = "L-methionine biosynthesis IV", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "superpathway of methylglyoxal degradation", y = 4.85, label = "reduced redox\nmaintenance", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "tetrapyrrole biosynthesis I",  xend = "L-ascorbate degradation I", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "pentose phosphate pathway I", y = 4.85, label = "increased redox\nmaintenance", hjust = 0, size = 5, color = "#80b1d3") +
  theme(axis.text.x = element_blank())
redox_p
  
nucleotide_p <- make_coef_plot(nucleotide) +
ggtitle("B. Nucleotide mobilization") +
  annotate("segment", x = "pyrimidine deoxyribonucleosides salvage",  xend = "adenosine nucleotides degradation I", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "UMP biosynthesis III", y = 4.85, label = "increased synthesis\nand salvage", hjust = 0, size = 5, color = "#80b1d3") +
  theme(axis.text.x = element_blank())
nucleotide_p 

oxygen_p <- make_coef_plot(oxygen) +
  ggtitle("E. Oxygen-driven metabolism") +
  ylab("Effect size") +
  annotate("segment", x = "superpathway of glycolysis, pyruvate dehydrogenase, TCA, and glyoxylate bypass",  xend = "glyoxylate cycle", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "superpathway of glycol metabolism and degradation", y = 4.85, label = "decreased aerobic\nmetabolism", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "pyruvate fermentation to acetate and (S)-lactate I",  xend = "formaldehyde assimilation II", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "glycogen degradation II", y = 4.85, label = "increased anerobic\nmetabolism", hjust = 0, size = 5, color = "#80b1d3") +
   theme(axis.text.x = element_text(size = 10, color = "black"))
oxygen_p

lipid_p <- make_coef_plot(lipid) +
  ggtitle("D. Lipid metabolism") +
  ylab("Effect size") +
  annotate("segment", x = "superpathway of glyoxylate cycle and fatty acid degradation",  xend = "9-cis, 11-trans-octadecadienoyl-CoA degradation", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "fatty acid salvage", y = 4.85, label = "decreased catabolism\nand recycling", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "stearate biosynthesis II",  xend = "petroselinate biosynthesis", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "gondoate biosynthesis", y = 4.85, label = "increased synthesis\nand remodeling", hjust = 0, size = 5, color = "#80b1d3") +
   theme(axis.text.x = element_text(size = 10, color = "black"))
lipid_p


immune_p <- ggplot(immune, aes(x = pathway, y = coef, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
    geom_point(position = position_dodge(width = 0.3), size = 1) +
    geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                   position = position_dodge(width = 0.3), linewidth = .35) +
    geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                   position = position_dodge(width = 0.3), linewidth = .55) +
    scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
    coord_flip(clip = "off") +
    ylim(-5, 5) +
    labs(y = "Effect size", x = NULL) +  # suppress axis labels
    theme(axis.line = element_line(color = "black", linewidth = .28),
          axis.text.x = element_text(size = 10, color = "black"),               
          axis.text.y = element_text(size = 10, color = "black"), 
          text = element_text(size = 10, color = "black", family = "sans"), 
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          legend.position = "none",
          panel.background = element_blank(),
          plot.margin = margin(t = 5, r = 110, b = 5, l = 5)) +
  ggtitle("C. Immune regulation") +
  annotate("segment", x = "acetyl-CoA fermentation to butanoate II",  xend = "L-glutamate degradation XI", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "L-glutamate degradation VII", y = 4.85, label = "decreased\nanti-inflammatory\nsignals", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "enterobacterial common antigen biosynthesis",  xend = "superpathway of (Kdo)2-lipid A biosynthesis", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = 3.5, y = 4.85, label = "decreased\ninflammatory\nsignals", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "(5Z)-dodecenoate biosynthesis I",  xend = "L-tryptophan biosynthesis", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = 1.5, y = 4.85, label = "increased\nanti-inflammatory\nsignals", hjust = 0, size = 5, color = "#80b1d3") 

immune_p



combined <- ggarrange(redox_p, nucleotide_p, immune_p, lipid_p, oxygen_p, ncol = 3, nrow = 2, align = "v")

ggsave("figures/temp_path_da.pdf", combined, width =25, height = 13)




# find midpoint betwen two discrete values 
levels <- levels(immune$pathway)

# Get numeric positions of those two
a <- which(levels == "enterobacterial common antigen biosynthesis")
b <- which(levels == "superpathway of (Kdo)2-lipid A biosynthesis")

# Find midpoint
mid <- (a + b) / 2


```


## d. plot model results

plot by level2 groupings

```{r}

plot_coef <- path_coef3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef)) %>%
  left_join(pathway_levels, by = "path_id")

# plot ditribution 
pdf("figures/pathway_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl2)) {
  
  data <- subset(plot_coef, lvl2 == i)
  
  p <- ggplot(data, aes(x = pathway, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 4)) {
  
  batch_plots <- plot_list[i:min(i + 4, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```


# 5. GAM analysis 

## a. run loop 

all converge

19 pathways where smooth model has more of the model weight

```{r}

gam_results <- list()

for (i in unique(path_abund3$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)
    
    m_main <- gam(clr_abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re"), data = data)

  }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # Update the warning message
      } 
    })

    # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      #create data frame
      results_df <- data.frame(
        path_id = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE
      )
  
      # store results
      gam_results[[i]] <- results_df
}

# combine results 
gam_results <- do.call(rbind, gam_results)

#Filter for taxa where m_full best fit
gam_filter <- gam_results %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id)
length(gam_filter) # 19 
  
# filter abundance
smooth_abund <- path_abund3 %>% filter(path_id %in% gam_filter) %>%
  filter(path_id != "NA")
length(unique(smooth_abund$path_id)) 

```


## b. generate predictions


```{r}

# Initialize an empty list to store results for all taxa
gam_preds <- list()

# Define the range of temp_s values for prediction
temp_seq <- seq(from = min(smooth_abund$temp_s, na.rm = TRUE), 
                to = max(smooth_abund$temp_s, na.rm = TRUE), length.out = 100)


# Iterate over unique values in final_temp_abund$path_id
for (i in unique(smooth_abund$path_id)) {
    
    # Subset your data based on current value of i
    data <- droplevels(subset(smooth_abund, path_id == i))
    
    # run model 
    gam_m <- gam(clr_abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex, data = data)
    
    # Generate predictions over the specified range of temp_s
    pred_data <- data.frame(
      age_s = mean(data$age_s, na.rm = TRUE),
      temp_s = temp_seq,
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = "Tenori", # assuming group is a factor
      sex = "M" # assuming sex is a factor
    )
    
    # Predict abundance values
    predicted_values <- predict(gam_m, newdata = pred_data)
    
    # Store the predicted values along with temperature sequence and Taxa label
    df <- data.frame(
      path_id = i,
      temp_s = temp_seq,
      preds = predicted_values
    )
    
    # Store the results in the list
    gam_preds[[i]] <- df
    
}

# Combine all results into a single dataframe
gam_preds <- do.call(rbind, gam_preds)
row.names(gam_preds) <- NULL # clean up row names

length(unique(gam_preds$path_id))

```

## c. custer analysis

four clusters fits best with cluster 1 (9 paths) and 2 (7 paths) including the most pathways 

```{r}

# Create a wide format dataframe
predicted_wide <- reshape(gam_preds, idvar = "temp_s", timevar = "path_id", direction = "wide")

# clean data
predicted_wide <- predicted_wide %>% 
  rename_with(~ str_replace(., "^preds.", ""), everything()) %>% # clean column names 
  select(-temp_s) # remove the temp_s column for correlation calculation and positive taxa

# elbow method
clust_plot <- fviz_nbclust(predicted_wide, kmeans, method = "wss", k.max = 10) + 
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank()) +
  labs(x = "Number of clusters", y = "Within-cluster sum of squares", title = "")
clust_plot

# set number of clusters
k=4

# distnce matrix
clus.distance = as.dist(1 - cor(predicted_wide, use = "pa")) # create distance matrix

h.clust = hclust(clus.distance,method='complete') # calculate clusters 

dendogram <- plot(h.clust, cex = 0.6, main = "", xlab = "", ylab = "") # plot dendogram
rect.hclust(h.clust, k = k, border = 2:6) # add clusters 


# Separate temp-relationships into k clusters 
clusters <- cutree(h.clust, k = k)

# Add cluster assignments to the original data
gam_preds$Cluster <- clusters[match(gam_preds$path_id, names(clusters))]

### Plot clusters ###

# clean data
clean_preds <- gam_preds %>% mutate(Cluster = as.factor(Cluster))

# by cluster 
c1 <- clean_preds %>% filter(Cluster == 1)
c2 <- clean_preds %>% filter(Cluster == 2)
c3 <- clean_preds %>% filter(Cluster == 3)
c4 <- clean_preds %>% filter(Cluster == 4)

p1 <- ggplot(c1, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p1

p2 <- ggplot(c2, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p2

p3 <- ggplot(c3, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p3

p4 <- ggplot(c4, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal()
p4

# Arrange the plots in a 2x2 layout
cluster_plot <- ggarrange(p1, p2, p3, p4,
                           ncol = 2, nrow = 2,
                          labels = c("C1", "C2", "C3", "C4"))
```

```{r}

c1_info <- c1 %>% select(path_id) %>%
  distinct() %>%
  left_join(pathway_levels, by = "path_id") %>%
  select(path_id, pathway)

```



## d. plot clusters

plotting the two main clusters 1 & 2

```{r}

mean_temp <- mean(smooth_abund$temp)
sd_temp <- sd(smooth_abund$temp) 

cluster1 <- ggplot(c1, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal() +
  labs(y = "Predicted log abundance", x = "Temperature (°C)") +
  scale_x_continuous(
   breaks = seq(min(c1$temp_s), max(c1$temp_s), length.out = 6), 
   labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 16, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
cluster1 

cluster2 <- ggplot(c2, aes(x = temp_s, y = preds, color = path_id)) + 
  geom_line() +
  theme_minimal() +
  labs(y = "Predicted log abundance", x = "Temperature (°C)") +
  scale_x_continuous(
   breaks = seq(min(c1$temp_s), max(c1$temp_s), length.out = 6), 
   labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 16, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
cluster2

```


# 6. Neopterin * temp DAA

## a. run interaction loop

```{r}

path_coef_int <- list()
path_aic_int <- list()
path_model_results_int <- list()

for (i in unique(path_abund3$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic_int[[i]] <- aic_df
    path_coef_int[[i]] <- results_df
    path_model_results_int[[i]] <- model_results_df

  })
}

# combine results
path_aic_int <- do.call(rbind, path_aic_int)
path_coef_int <- do.call(rbind, path_coef_int)
path_model_results_int <- do.call(rbind, path_model_results_int)

```


## b. select interaction taxa

tested 169, 53 of those didn't converge so tested 116

out of 116 the interaction model first best for 65 of them 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic_int$path_id[path_aic_int$Warning == "Convergence Warning"]) # 53 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results_int %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no big SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 53 do not converge

# remove error taxa
path_abund4 <- path_abund3 %>% filter(!path_id %in% errors)
length(unique(path_abund4$path_id)) # 116 remaining 

path_coef_int2 <- path_coef_int %>% filter(!path_id%in% errors)
length(path_coef_int2$path_id)

path_aic_int2 <- path_aic_int %>% filter(!path_id %in% errors)
length(path_aic_int2$path_id) 

# filter for taxa where full model fits best
interaction_filter <- path_aic_int2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(interaction_filter)) # 65 interaction model fits best for all

# filter df's for where temp model fits best 
path_abund5 <- path_abund4 %>% filter(path_id %in% interaction_filter)
length(unique(path_abund5$path_id))  

path_coef_int3 <- path_coef_int2 %>% filter(path_id %in% interaction_filter)
length(path_coef_int3$path_id)

path_aic_int2 <- path_aic_int2 %>% filter(path_id %in% interaction_filter)
length(path_aic_int2$path_id)

```

## c. plot model results

```{r}

plot_coef <- path_coef_int3 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef)) %>%
  left_join(pathway_levels, by = "path_id")

length(unique(plot_coef$lvl2)) # 27
length(unique(plot_coef$lvl1)) # 6

# plot distribution 
pdf("figures/pathway_interaction_DAA.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(plot_coef$lvl2)) {
  
  data <- subset(plot_coef, lvl2 == i)
  
  p <- ggplot(data, aes(x = pathway, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
    labs(title = paste(i)) 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 4)) {
  
  batch_plots <- plot_list[i:min(i + 4, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

## d. make coplot

Only using two bins because we don't have a wide range of neopterin values for the shotgun data

```{r}

# Create neo bins
path_abund5$neo_bin <- cut(path_abund5$neo_s, breaks = 2, labels = c("Bin1", "Bin2"), include.lowest = TRUE)
bin_limits <- levels(cut(path_abund5$neo_s, breaks = 2, include.lowest = TRUE)) # look at bin range
bin_avg <- tapply(path_abund5$neo_s, path_abund5$neo_bin, mean, na.rm = TRUE) # use for co plot

# Create temp seq
temp_seq <- seq(min(path_abund5$temp_s, na.rm = TRUE), max(path_abund5$temp_s, na.rm = TRUE), length.out = 100)

# list
neo_preds <- list()

# loop
for (i in unique(path_abund5$path_id)) {
  
  data <- droplevels(subset(path_abund5, path_id == i))
  
  m <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s,
            data = data)
  
  coef <- round(summary(m)$coefficients["temp_s:neo_s", 1], 6)
  
  # neo_s values and bins
  neo_values <- c(-0.04380621, 3.06643502)
  bins <- paste0("neopterin_avg", 1:2)
  
  # Generate predictions for each neo_s value
  preds <- lapply(seq_along(neo_values), function(j) {
    neo_s <- neo_values[j]
    bin <- bins[j]
    
    # preds dataset
    pred <- expand.grid(
      temp_s = temp_seq,
      neo_s = neo_s,
      age_s = mean(data$age_s, na.rm = TRUE),
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = "Tenori",
      sex = "M"
    )
    
    # get model predictions
    preds_model <- predict(m, newdata = pred, se.fit = TRUE)
    
    # store preds and CI's 
    df <- data.frame(
      temp_s = pred$temp_s,  
      neo_s = neo_s,
      fit = preds_model$fit,
      lower = preds_model$fit - 1.96 * preds_model$se.fit,
      upper = preds_model$fit + 1.96 * preds_model$se.fit,
      bin = bin,
      coef = coef
    )
    
    return(df)
  })
  
  # combine predictions for all bins 
  preds_df <- do.call(rbind, preds)
  preds_df$path_id <- i 
  
  # store preds
  neo_preds[[i]] <- preds_df
}

# Combine all predictions into a single dataframe
neo_preds <- do.call(rbind, neo_preds)
row.names(neo_preds) <- NULL # clean up row names

```

## e. coplot by patheay

Plot with the CI's for each taxa for the supplement

```{r}

pdf("figures/pathway_interaction.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(neo_preds$path_id)) {
  
  data <- subset(neo_preds, path_id == i)

  p <- ggplot(data, aes(x = temp_s, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "grey") +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(
      title = paste("predicted Abundance for", i),
      x = "temperature (scaled)",
      y = "predicted log abundance") 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

## f. coplot by effect direction

### i. data prep

```{r}

# Define effect type

slope <- neo_preds %>%
  group_by(path_id, bin) %>%
  do({m <- lm(fit ~ temp_s, data = .)
    slope <- coef(m)["temp_s"]
    data.frame(slope = slope)}) %>%
  ungroup() %>%
  pivot_wider(names_from = bin, values_from = slope,
                     names_prefix = "slope_") %>%
  mutate(effect_type = case_when(
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 > 0 & slope_neopterin_avg2 > slope_neopterin_avg1 ~ "positive_amplifying",
      slope_neopterin_avg1 < 0 & slope_neopterin_avg2 < 0 & slope_neopterin_avg1 > slope_neopterin_avg2 ~ "negative_amplifying",
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 > 0 & slope_neopterin_avg2 < slope_neopterin_avg1 ~ "positive_dampening",
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 < 0 ~ "negative_reverse",
      slope_neopterin_avg1 < 0 & slope_neopterin_avg2 > 0 ~ "positive_reverse",
      TRUE ~ "other")) %>%
  select(path_id, effect_type)
      
      
# Prep data

mean_neo <- mean(path_abund5$neo)
sd_neo <- sd(path_abund5$neo) 

neo_reverse <- function(x) {
  return(round(x * 2 * sd_neo + mean_neo))  # reverse scale in figure axes
}

bin_limits <- levels(cut(path_abund5$neo_s, breaks = 2, include.lowest = TRUE)) # look at bin range
neo_reverse(1.24)

neo_preds2 <- neo_preds %>% mutate(bin = case_when(bin == "neopterin_avg1" ~ "neopterin: 107-2812 (nmol/L)",
                                                   bin == "neopterin_avg2" ~ "neopterin: 2812-5524 (nmol/L)")) %>%
  left_join(slope, by = "path_id") %>%
  left_join(pathway_levels, by = "path_id")
  
# Create df's for each effect                               
                
positive_amplifying <- neo_preds2 %>% filter(effect_type == "positive_amplifying")
negative_amplifying <- neo_preds2 %>% filter(effect_type == "negative_amplifying")
positive_dampening <- neo_preds2 %>% filter(effect_type == "positive_dampening")
negative_reverse <- neo_preds2 %>% filter(effect_type == "negative_reverse")
positive_reverse <- neo_preds2 %>% filter(effect_type == "positive_reverse")

# info for each interaction type
length(unique(positive_amplifying$path_id)) # 10
length(unique(negative_amplifying$path_id)) # 7
length(unique(positive_dampening$path_id)) # 3
length(unique(negative_reverse$path_id)) # 18
length(unique(positive_reverse$path_id)) # 27

info_positive_amplifying <- positive_amplifying %>%
  select(path_id, lvl1, lvl2, pathway, effect_type) %>%
  distinct()

info_negative_amplifying <- negative_amplifying %>%
  select(path_id, lvl1, lvl2, pathway, effect_type) %>%
  distinct()

info_positive_dampening <- positive_dampening %>%
  select(path_id, lvl1, lvl2, pathway, effect_type) %>%
  distinct()

info_negative_reverse <- negative_reverse %>%
  select(path_id, lvl1, lvl2, pathway, effect_type) %>%
  distinct()

info_positive_reverse <- positive_reverse %>%
  select(path_id, lvl1, lvl2, pathway, effect_type) %>%
  distinct()



# bin values
unique_bins <- unique(neo_preds2$bin)

# to rescale x axis
mean_temp <- mean(path_abund5$temp)
sd_temp <- sd(path_abund5$temp) 

```

### ii. plot

```{r}

# plot variables
ylab = "predicted\nclr-transformed\nabundance"


################## positive_amplifying ###########################
y_min <- min(positive_amplifying$fit)
y_max <- max(positive_amplifying$fit)

positive_amp_list <- list()

# loop
for (i in unique_bins) {
  
  pos_amp_filtered <- positive_amplifying[positive_amplifying$bin == i, ]
  
  # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 107-2812 (nmol/L)") {
    ylab
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(pos_amp_filtered, aes(x = temp_s, y = fit, color = path_id)) +
    geom_line() +
    labs(y = y_axis_label, x = "", title = paste(i)) +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  positive_amp_list[[as.character(i)]] <- bin_plot
  
}

positive_amp_plot <- do.call(ggarrange, c(positive_amp_list, nrow = 1))
positive_amp_plot


################### ngeaitve_amplifying ######################

y_min <- min(negative_amplifying$fit)
y_max <- max(negative_amplifying$fit)

negative_amp_list <- list()

# loop
for (i in unique_bins) {
  
  negative_amp_filtered <- negative_amplifying[negative_amplifying$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 107-2812 (nmol/L)") {
    ylab
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(negative_amp_filtered, aes(x = temp_s, y = fit, color = path_id)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  negative_amp_list[[as.character(i)]] <- bin_plot
  
}

negative_amp_plot <- do.call(ggarrange, c(negative_amp_list, nrow = 1))
negative_amp_plot


################## positive_dampening ###########################

y_min <- min(positive_dampening$fit)
y_max <- max(positive_dampening$fit)

positive_dampening_list <- list()

# loop
for (i in unique_bins) {
  
  positive_dampening_filtered <- positive_dampening[positive_dampening$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 107-2812 (nmol/L)") {
    ylab
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(positive_dampening_filtered, aes(x = temp_s, y = fit, color = path_id)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  positive_dampening_list[[as.character(i)]] <- bin_plot
  
}

positive_dampening_plot <- do.call(ggarrange, c(positive_dampening_list, nrow = 1))
positive_dampening_plot


################## negative_reverse ###########################

y_min <- min(negative_reverse$fit)
y_max <- max(negative_reverse$fit)

negative_reverse_list <- list()

# loop
for (i in unique_bins) {
  
  negative_reverse_filtered <- negative_reverse[negative_reverse$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 107-2812 (nmol/L)") {
    ylab
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(negative_reverse_filtered, aes(x = temp_s, y = fit, color = path_id)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  negative_reverse_list[[as.character(i)]] <- bin_plot
  
}

negative_reverse_plot <- do.call(ggarrange, c(negative_reverse_list, nrow = 1))
negative_reverse_plot


################## positive_reverse ###########################

y_min <- min(positive_reverse$fit)
y_max <- max(positive_reverse$fit)

positive_reverse_list <- list()

# loop
for (i in unique_bins) {
  
  positive_reverse_filtered <- positive_reverse[positive_reverse$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 107-2812 (nmol/L)") {
    ylab
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(positive_reverse_filtered, aes(x = temp_s, y = fit, color = path_id)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    scale_x_continuous(
      breaks = seq(min(neo_preds2$temp_s), max(neo_preds2$temp_s), length.out = 3),
      labels = inverse_scale) +
      theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  positive_reverse_list[[as.character(i)]] <- bin_plot
  
}

positive_reverse_plot <- do.call(ggarrange, c(positive_reverse_list, nrow = 1))
positive_reverse_plot


######## combine ##########

combined_plot <- ggarrange(positive_amp_plot, negative_amp_plot, positive_dampening_plot, negative_reverse_plot, positive_reverse_plot, 
                           nrow = 5, ncol = 1, 
                           labels = c("A", "B", "C", "D", "E"))
combined_plot

#ggsave("figures/fig7.pdf", combined_plot, width =13, height = 9)


```


