---
title: "2025-01-16 GM outcome compiled"
author: "Lucore"
date: "2025-01-13"
output: html_document
---

Notes: this is where we compile all of the results where GM is the outcome variable
We are using the smaller subset of samples that have a neo sample three days before or after so the total is 277 samples. This script is different from 2024-01-13_GM_outcome_compiled because do the full analysis on the neopterin samples instead of just the temp analysis with the larger subset of samples. The answers for the temp analysis questions are essentially the same with the smaller subset than with the larger subset and the workflow is easier to use the same samples for everything. 

To do: 
Investigate Gam convergence becaue some of the individual plots look wonky
incorporate neopterin plot to show that neopterin is relaitvely stable over 2 weeks, justification for sample pairing

change names of models to match methods writing


# 1. Load environment

```{r}

library(tidyverse)
library(phyloseq)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(ggnewscale)
library(ggtree)
library(tidytree)
library(ape)

physeq <- readRDS("phyloseq_objects/2024-12-18_phyloseq_16s.rds")

source("functions/my_scale.R")

source("functions/inverse_scale.R")

conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)

```


# 2. Data set up

## a. Set up phyloseq object

Subset for samples that have diet information so that we can control for fruit and neopterin, using neopterin averaged 3 days before and after GM sample. 277 remaining samples

```{r}

m <- data.frame(sample_data(physeq)) %>% # extract metadata 
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals, neo)

m_filtered <- m[!is.na(m$fruit), ] # remove samples with no diet data
m_filtered <- m_filtered[!is.na(m_filtered$neo), ] # remove samples with no neo

physeq2 <- prune_samples(rownames(m_filtered), physeq) # filter physeq object

physeq2 <- prune_taxa(taxa_sums(physeq2) > 0, physeq2) # remove ASV's with zero abundance

physeq2

```

## b. Prevalence filtering 

Identifies taxa that are not present in at least 0.5% of the total samples (at least two samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth.  

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(physeq2),
                MARGIN = ifelse(taxa_are_rows(physeq2), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(physeq2),
                      tax_table(physeq2))

# define prevalence threshold
prevalenceThreshold = 0.005 * nsamples(physeq2)  
prevalenceThreshold # must appear in at least two samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

ggsave("figures/supp/figS_prev.pdf", prev_fig, width = 6, height = 4)

## summary stats ##
asv_onesample <- nrow(prevdf[prevdf$Prevalence == 1,]) # number of ASV's present only in one sample
total_asv <- nrow(prevdf)
percent_onesample <- (asv_onesample/total_asv) * 100 # perent ASV's in only one sample

asv_nophyla <- nrow(prevdf[is.na(prevdf$Phylum),]) # number of ASV's with no assigned phylum 
percent_nophyla <- (asv_nophyla/total_asv) * 100 # percent number of ASV's with no phylum

## execute prevalence filter ##
phylo_filtered <- prune_taxa((prev0 > prevalenceThreshold), physeq2) 
phylo_filtered

## remove taxa with unnasigned phyla ## 
phylo_filtered2 <- subset_taxa(phylo_filtered, !is.na(Phylum)) # remove 102 taxa 
phylo_filtered2

# calc reads per sample
reads_filtered <- as.data.frame(sample_sums(phylo_filtered2)) %>% 
  rownames_to_column(var = "sample") %>%
  rename(filtered_reads = 'sample_sums(phylo_filtered2)')
  
# add reads per sample to metadata
m_filtered2 <- m_filtered %>%
  left_join(reads_filtered, by = "sample") 

# % assignment for remaining levels

asign_class <- 100-(((sum(is.na(tax_table(phylo_filtered2)[, "Class"])))/1159)*100)

asign_order <- 100-(((sum(is.na(tax_table(phylo_filtered2)[, "Order"])))/1159)*100)

asign_family <- 100-(((sum(is.na(tax_table(phylo_filtered2)[, "Family"])))/1159)*100)

asign_genus <- 100-(((sum(is.na(tax_table(phylo_filtered2)[, "Genus"])))/1159)*100)

```

##  b. Calc alpha diversity

calculate alpha diversity metrics, Shannon and observed with the filtered phyloseq object. The results are the same if you calculate before the filtering

```{r}

shannon <- estimate_richness(phylo_filtered2, measures = "Shannon") # calc shannon
shannon <- shannon %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

hist(shannon$shannon) # normal dist

observed <- estimate_richness(phylo_filtered2, measures = "Observed") # calc observed
observed <- observed %>%
  rownames_to_column(var = "sample") %>%
  rename(observed = Observed) %>%
  mutate( sample = str_replace_all(sample, "^X", ""),  
          sample = str_replace_all(sample, "\\.", "-"))

hist(observed$observed) # gamma dist


m_filtered3 <- m_filtered2 %>% # add to metadata
  left_join(shannon, by = "sample") %>%
  left_join(observed, by = "sample")

```


## d. CLR transformation

ROWS NEED TO BE SAMPLES WHEN DOING A PCA BECAUSE IT WILL TELL YOU HOW THE SAMPLE CLUSTER TOGETHER BASED ON THE ASV's instead of samples

we use the CLR transformation and PCA analysis because of Gloor et al., 2017, Baniel publications, and Gloor and Ried 2016 and this tutorial https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot They suggest that normalizing for sequencing depth is best done with a CLR transformation (as opposed to rarefaction) and that sample distances are best estimated using Aichisons distances as opposed to Bray Curtis or Unifrac because they do not reflect the compositional nature of the data. Aichison distance is the euclidean distances between the samples after the CLR transformation.

```{r}

library(zCompositions)

otu_table <- as.matrix(t(otu_table(phylo_filtered)))
otu_table[1:5,1:5] # rows are samples 

set.seed(123)
czm <- cmultRepl(otu_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
czm[1:5,1:5] # rows are samples 

clr <- t(apply(czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
clr[1:5,1:5] # rows are samples 

pca <- prcomp(clr) # calc pca
plot(pca$x[,1],pca$x[,2]) # plot 

d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="") # PC1 14.4%
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 8.1%

```

## e. Add PCA values to metadata

Choose how many axes to include using a scree plot which plots the percent explained by each axis. Axes 1-4 explain the largest proportion of the variance. 

```{r}

eigenvalues <- pca$sdev^2  # square the standard deviations to get eigenvalues
percent_explained <- 100 * eigenvalues / sum(eigenvalues) 

# scree plot
pe_plot <- tibble(pe = percent_explained, axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  geom_point() +
  labs(x = "Principal component axis", y = "Percent variance explained") +
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig1S.pdf", pe_plot, width = 6, height = 6)

df<-data.frame(pca$x)
df$sample<-rownames(df)
df_small <- df[, c(1:4,278)]

# add to metadata
m_filtered4 <- m_filtered3 %>%
  left_join(df_small, by = "sample") 

```

## f. prep data for modeling

scale variables and make categorical variables factors

make all palmas sampels tenori, only 6 palmas samples not enough to estimate effect 

```{r}

# scale variables
m_model <- m_filtered4 %>%
  mutate(across(c(age, temp, rain, fruit, residuals, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)),
         ID = as.factor(ID))  

```

## g. Suppelemenatary food plots

show that temp and fruit are correlated, to be able to evaluate temp, need to control for fruit.  

```{r}

# fruit ~ temp
fruit_plot <- ggplot(m_model, aes(x = temp, y = fruit)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") +
  labs(y = "Count of fruit eaten", x = "Temperature (°C)") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig2S.pdf", fruit_plot, width = 6, height = 6)

```


# 3. Community features analysis

## a. model community features

Compare m_main with group, age, sex, fruit, diet diversity, rain, and ID as predictors to m_full which adds a temperature predictor to show important of temp as predictor for community features. Both model include an offset of reads to control for seq depth. Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed. 

When you run the loop there is a singular fit error, this is from PC2, but I the output looks okay and it is only for the m_main model not the m_full model. We are using the coefficient from m_full so I think this is fine. 

```{r}

com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15, 17:20)] # list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      
      m_main <- lmer(formula_main, data = m_model)
      m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)


#### model observed ######

o_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = m_model)

o_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = m_model)

AICc <- as.data.frame(AICctab(o_full, o_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(o_full)$coeff[8,1], 4)
std <- round(summary(o_full)$coeff[8,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

com_coef <- rbind(com_coef, results_df)
com_aic <- rbind(com_aic, aic_df)

```

## b. plot community features

```{r}

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) 
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 18, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1

ggsave("figures/fig1.pdf", p1, width = 8, height = 6)


```

## c. plot alpha diversity

```{r}

# set up predictions df
temp_seq <- seq(from = min(m_model$temp_s, na.rm = TRUE), 
                to = max(m_model$temp_s, na.rm = TRUE), length.out = 100)

pred_data <- data.frame(
  group = unique(m_model$group)[1], 
  age_s = mean(m_model$age_s, na.rm = TRUE),
  sex = unique(m_model$sex)[1], 
  fruit_s = mean(m_model$fruit_s, na.rm = TRUE),
  residuals_s = mean(m_model$residuals_s, na.rm = TRUE),
  rain_s = mean(m_model$rain_s, na.rm = TRUE),
  temp_s = temp_seq,
  filtered_reads = mean(m_model$filtered_reads, na.rm = TRUE))

########## shannon ##########
s_temp <- lm(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + offset(log(filtered_reads)), 
               data = m_model)
s_preds <- predict(s_temp,  newdata = pred_data, interval = "confidence", level = 0.95 ) %>%
  data.frame() %>%
  rename(s_preds = fit,
         s_lower = lwr,
         s_upper = upr)
pred_data <- cbind(pred_data, s_preds) 

# to rescales x axis
mean_temp <- mean(m_model$temp)
sd_temp <- sd(m_model$temp) 

s_plot <- ggplot(pred_data, aes(x = temp_s, y = s_preds)) +
  geom_point(data = m_model, aes(x = temp_s, y = shannon),alpha = 0.5, size = 4, color = "#80b1d3") +
  geom_smooth(method = "lm", lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = s_lower, ymax = s_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  labs(y = "Shannon",
       x = "temperature (°C)") +
  scale_x_continuous(
    breaks = seq(min(m_model$temp_s), max(m_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.line = element_line(color = "black"),
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text = element_text(size = 12, color = "black"))

########## observed ##########
o_temp <- glm(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s +
                offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = m_model)

o_preds <- data.frame(predict(o_temp,  newdata = pred_data, type = "response", se.fit = TRUE)) %>%
  select(-residual.scale) %>%
  rename(o_preds = fit,
         o_se = se.fit)

pred_data <- cbind(pred_data, o_preds) %>%
  mutate(o_lower = o_preds - 0.97 * o_se, # calc CI's 
         o_upper = o_preds + 0.97 * o_se)

o_plot <- ggplot(pred_data, aes(x = temp_s, y = o_preds)) +
  geom_point(data = m_model, aes(x = temp_s, y = observed),alpha = 0.5, size = 4, color = "#80b1d3") +
  geom_line(lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = o_lower, ymax = o_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  labs(y = "Observed",
       x = "temperature (°C)") +
  scale_x_continuous(
    breaks = seq(min(m_model$temp_s), max(m_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.line = element_line(color = "black"),
  axis.title.x = element_text(size = 20),
  axis.title.y = element_text(size = 20),
  axis.text = element_text(size = 12, color = "black"))

# add plots together  

combo <- ggarrange(s_plot, o_plot, ncol = 2, nrow = 1, labels = c("A", "B"))

ggsave("figures/fig2.pdf", combo, width =12, height = 6)

```


# 4. Data set up for Taxa abundance testing

## a. Preprocessing

clean up taxa names in phyloseq object, use `tax_glom()` to combine taxa at the genus level and at the family level (i.e. sum abundances of the ASV's that are grouped together into genus or family). Change row names from ASV ID to taxa name. 

```{r}

# get rid of all the weird, unnassigned names and turn them all into NA's
tax_table(phylo_filtered2)<-apply(tax_table(phylo_filtered2),2, function(x) (gsub("Unassigned*",NA,x)))
tax_table(phylo_filtered2)<-apply(tax_table(phylo_filtered2),2, function(x) (gsub("uncultured*",NA, x)))
tax_table(phylo_filtered2)<-apply(tax_table(phylo_filtered2),2, function(x) (gsub("unidentified*",NA, x)))
tax_table(phylo_filtered2)<-apply(tax_table(phylo_filtered2),2, function(x) (gsub("gut metagenome",NA, x)))
tax_table(phylo_filtered2)<-apply(tax_table(phylo_filtered2),2, function(x) (gsub("metagenome",NA, x)))

# Tax glom 
genus_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered2, "Genus"))
family_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered2, "Family"))

# extract tables
genus_otu_table <- as.data.frame(otu_table(genus_aggregated))
genus_tax_table <- as.data.frame(tax_table(genus_aggregated))
genus_metadata <- sample_data(genus_aggregated)

family_otu_table <- as.data.frame(otu_table(family_aggregated))
family_tax_table <- as.data.frame(tax_table(family_aggregated))
family_metadata <- sample_data(family_aggregated)

#change OTU rownames to ASV family names
# Ensure that the OTU IDs are the same in both tables (rows in otu_df and tax_df)
genus_otu_ids <- rownames(genus_otu_table)
genus_tax_ids <- rownames(genus_tax_table)

family_otu_ids <- rownames(family_otu_table)
family_tax_ids <- rownames(family_tax_table)

# Extract the taxa names from the taxonomy table
genus_names <- genus_tax_table[genus_otu_ids, "Genus"]
family_names <- family_tax_table[family_otu_ids, "Family"]

# Set the row names of the OTU table to the family names
rownames(genus_otu_table) <- genus_names
rownames(family_otu_table) <- family_names

```

## b. Calc rel abundance 

Calculate the relative abundance by dividing the raw abundance counts by the total reads per sample. Using the "filtered_reads" column because we are using the phyloseq object `phylo_filtered2` that has already been filtered at a 0.5% threshold. 

```{r}

# extract table for DA testing
genus_da <-psmelt(genus_aggregated)
family_da <-psmelt(family_aggregated)

# Extract variables of interest 
genus_da1 <- genus_da %>% 
  mutate(Taxa = Genus) %>% # make genus the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>% 
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Genus") %>%
  mutate(RA = (Abundance/filtered_reads)*100)

family_da1 <- family_da %>% 
  mutate(Taxa = Family) %>% # make family the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>%
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Family") %>%
  mutate(RA = (Abundance/filtered_reads)*100) # calculates relative abundance

# n = 277
length(unique(genus_da1$sample))  
length(unique(family_da1$sample))

```

## c. Filter out low abundance taxa

Calculate the average relative abundance for each taxa across all of the samples and then keep the taxa that have at lease 0.2% relative abundance across all samples. Left with 134 genera and 79 families. 

```{r}

#Calculate average relative abundance of each taxa across all samples
genus_a1<-sapply(unique(genus_da1$Taxa),function(x) mean(genus_da1$RA[genus_da1$Taxa==x],na.rm=TRUE))
genus_a2<-data.frame(cbind(Taxa=as.character(unique(genus_da1$Taxa)),RA=genus_a1))
genus_a2$RA<-as.numeric(as.character(genus_a2$RA))

family_a1<-sapply(unique(family_da1$Taxa),function(x) mean(family_da1$RA[family_da1$Taxa==x],na.rm=TRUE))
family_a2<-data.frame(cbind(Taxa=as.character(unique(family_da1$Taxa)),RA=family_a1))
family_a2$RA<-as.numeric(as.character(family_a2$RA))


#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

genus_da2 <-droplevels(subset(genus_da1, genus_da1$Taxa %!in% NA)) # remove asv NAs
genus_da3 <-droplevels(subset(genus_da2, genus_da2$Taxa %in% unique(genus_a2$Taxa[genus_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(genus_da2$Taxa))
length(unique(genus_da3$Taxa)) # 134 genera

family_da2 <-droplevels(subset(family_da1, family_da1$Taxa %!in% NA)) # remove asv NAs
family_da3 <-droplevels(subset(family_da2, family_da2$Taxa %in% unique(family_a2$Taxa[family_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(family_da2$Taxa))
length(unique(family_da3$Taxa)) # 79 families

# Combine family and genus data frames 
da4 <- rbind(genus_da3, family_da3)

```

## d. Scale variables

scale, combine tenori and palmas, 211 total taxa for data 

```{r}

# Scale variables
da5 <- da4 %>%
   mutate(across(c(temp, rain, age, fruit, neo), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID),
         sex = as.factor(sex)) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)))

length(unique(da5$Taxa)) # 211 taxa total
length(unique(da5$ID))

```

## phylum barplot

```{r}

phylum_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered2, "Phylum"))
  
phylum_otu_table <- as.data.frame(otu_table(phylum_aggregated))
phylum_tax_table <- as.data.frame(tax_table(phylum_aggregated))

phylum_otu_ids <- rownames(phylum_otu_table)
phylum_tax_ids <- rownames(phylum_tax_table)

phylum_names <- phylum_tax_table[phylum_otu_ids, "Phylum"]

rownames(phylum_otu_table) <- phylum_names


# by sample
phylum_otu_table2 <- phylum_otu_table %>%
  rownames_to_column("Phylum") %>%
  pivot_longer(-Phylum, names_to = "sample", values_to = "Count") %>%
  left_join(reads_filtered, by = "sample") %>%
  group_by(sample) %>%
  mutate(RelativeAbundance = Count / sum(Count),
          Phylum_grouped = case_when(
           str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
           RelativeAbundance > 0.01 ~ Phylum,  
           TRUE ~ "Less than 1% rel abund"  
         )) %>%
  ungroup()

ggplot(phylum_otu_table2 , aes(x = sample, y = RelativeAbundance, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Relative Abundance", fill = "Phylum") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



# across samples
df_total <- phylum_otu_table2 %>%
  mutate(Phylum_grouped = case_when(
    str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
    TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarise(TotalCount = sum(Count)) %>%
  ungroup() %>%
  mutate(RelativeAbundance = TotalCount / sum(TotalCount)) %>%
  mutate(Phylum_grouped = case_when(RelativeAbundance > 0.01 ~ Phylum_grouped, 
                                    TRUE ~ "Less than 1% rel abund")) %>%
  group_by(Phylum_grouped) %>%
  summarise(RA = sum(RelativeAbundance)) %>%
  ungroup() %>%
  mutate(RA = 100*RA,
         Phylum_grouped = fct_reorder(Phylum_grouped, RA))
                                    

ggplot(df_total, aes(x = Phylum_grouped, y = RA, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Relative Abundance (%)", fill = "Phylum") +
  theme_minimal() +
  coord_flip() 

```


# 5. Temperature taxa abundance testing

## a. AIC comparison

this loops takes a while to run. 

Compare m_main with all predictors except temp and m_full which includes temp for all taxa. Not including diet diversity in these models to help simplify the models. Diet diversity is more importation for alpha diversity. 

```{r}

# create lists
taxa_coef <- list()
taxa_aic <- list()
model_results <- list()

# loop
for (i in unique(da5$Taxa)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(da5, Taxa == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- gam(Abundance ~ age_s + temp_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 

      m_main <- gam(Abundance ~ age_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 
      
    }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
     # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      # extract temp coef
      coef <- round(summary(m_full)$p.coeff[3], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[3], 4)
    
      # clac CI'
      ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
      ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
      # bind data
      df <- cbind(ci_95, ci_50)
      
      #create data frames
    aic_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
        Taxa = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3),
        stringsAsFactors = FALSE)

    options(scipen = 999)  

    model_results_df <- data.frame(
      Taxa = i,
      Predictor = names(summary(m_full)$p.coeff[c(2:7)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:7)], 4),
      StdError = round(summary(m_full)$se[c(2:7)], 4),
      stringsAsFactors = FALSE)
    
    
      # save results
    taxa_aic[[i]] <- aic_df
    taxa_coef[[i]] <- results_df
    model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
taxa_aic <- do.call(rbind, taxa_aic)
taxa_coef <- do.call(rbind, taxa_coef)
model_results <- do.call(rbind, model_results)



```

## b. taxa selection

First, we remove all taxa with explicit convergence warnings, then we remove all taxa that look like they didn't converge based on SE in the model results. 48 number of taxa did not converge. After removing the taxa that don't converge we tested 163 taxa in total and 93 of those have m_full with a model weight  > 0.5

Then, we select all taxa where m_full (the model with the temp predictor) fits the best. 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(taxa_aic$Taxa[taxa_aic$Warning == "Convergence Warning"]) # 22 do not converge

# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 80) %>%
  pull(Taxa) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 48 do not converge

# remove error taxa
da6 <- da5 %>% filter(!Taxa %in% errors)
length(unique(da6$Taxa)) # 163 remaining taxa

taxa_coef2 <- taxa_coef %>% filter(!Taxa %in% errors)
length(taxa_coef2$Taxa)

taxa_aic2 <- taxa_aic %>% filter(!Taxa %in% errors)
length(taxa_aic2$Taxa) # 324/2 = 162

# filter for taxa where full model fits best
temp_filter <- taxa_aic2 %>%
  group_by(Taxa) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa) # 93 taxa
length(unique(temp_filter))

filtered_coef <- taxa_coef2 %>%filter(Taxa %in% temp_filter)
length(filtered_coef$Taxa)

filtered_aic <- taxa_aic2 %>%filter(Taxa %in% temp_filter)
length(filtered_aic$Taxa)

```

## c. coef plot

```{r}

# get level info

levels <- da6 %>%
  select(Taxa, level) %>%
  unique()

plot_taxa_coef <- filtered_coef %>% 
  left_join(levels, by = "Taxa") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         Taxa = reorder(Taxa, -coef))

table(plot_taxa_coef$level)
table(plot_taxa_coef$direction)
         
p1 <- ggplot(plot_taxa_coef, aes(x = Taxa, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  facet_wrap(~level, nrow = 2, ncol = 1)

p1

ggsave("scripts/compiled/gm_outcome/figures/taxa_temp_coef.pdf", p1, width = 8, height = 16)

```

## phylum analysis

seeing a decrease across almost all phyla, different from what the literature shows on other species with a decrease in firmicutes and an increase in proteo bacteria across species (moeller review) and the increase in bacteroidota shown in meerkats. 

```{r}

phyla <- da6 %>% select(Taxa, Phylum) %>%
  unique()

length(phyla$Taxa)

filtered_coef2 <- filtered_coef %>% left_join(phyla, by = "Taxa")

plot_phyla_coef <- filtered_coef2 %>% 
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         Taxa = reorder(Taxa, -coef))

table(plot_phyla_coef$Phylum)
table(plot_taxa_coef$direction)

ggplot(plot_phyla_coef, aes(x = Taxa, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(position = position_dodge(width = 0.3), size = 1) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95),
                 position = position_dodge(width = 0.3), linewidth = .35) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50),
                 position = position_dodge(width = 0.3), linewidth = .55) +
  scale_color_manual(values = c("#bc80bd", "#fdb462"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "Taxa") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 8, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  facet_wrap(~Phylum, nrow = 5, ncol = 2)

```


## d. tree plot

Create cladogram that shows the results of the temp DAA for genera

### i. data prep

extract the necessary metadata and filter the phyloseq object for all genera included in the DA temp analysis. Then extract the tree. 

```{r}

# get list of taxa with ne
direction <- filtered_coef %>% 
  mutate(effect_direction = ifelse(coef > 0, "positive", "negative")) %>%
  select(Taxa, effect_direction)

# add temp information to da table
da7 <- da6 %>%
  mutate(taxa_temp_sensitive = ifelse(Taxa %in% temp_filter, "sensitive", "nuetral")) %>%
  left_join(direction, by = "Taxa") %>%
  mutate(effect_direction = case_when(
    is.na(effect_direction) ~ "neutral",
    TRUE ~ effect_direction))
  
# get genus names
genus_names <- da7 %>%
  filter(level == "Genus") %>%
  pull(Taxa) %>%
  unique()

# aggregate by taxa
genus_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered, "Genus"))

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
temp_genus_phylo <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)
genus_tax <- as.data.frame(tax_table(temp_genus_phylo))

# prep final tree metadata 
genus_meta <- da7 %>%
  filter(level == "Genus") %>%
  select(Taxa, taxa_temp_sensitive, effect_direction) %>%
  distinct() %>%
  left_join(taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight), by = "Taxa")

# create phylum groups for clade labels
phylum_groups <- genus_tax %>%
  rownames_to_column("OTU") %>%
  mutate(Genus = as.character(Genus), Phylum = as.character(Phylum)) %>%
  filter(!is.na(Phylum), !is.na(Genus)) %>%
  group_by(Phylum) %>%
  summarise(genus_list = list(Genus)) %>%
  unnest_longer(genus_list, values_to = "Genus")

```

### ii. assign tip labels

First assign the tip labels as genus names instead of otu numbers. Add the metadata for temp sensitivity and change the tip labels to the model weight varible

```{r}

# tip labels are still otu numbers
head(genus_tree$tip.label)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels
  
# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# now make tip labels the model weight
genus_tree$tip.label <- tip_data$weight
head(genus_tree$tip.label)

```

### iii. ancestral state reconstruction

reconstruction returns the maximum likelihood estimates of the model weights for each node, assumes brownian motion. Higher weights are more temperature sensitive than lower weights. 

```{r}

# run ace function
ace_result <- ace(tip_data$weight, genus_tree, type = "continuous")

# extract node probs
node_df <- data.frame(
  node = as.numeric(names(ace_result$ace)),
  ace_weight = ace_result$ace) # ignore warning will fix in next step

# assign node probs to node
node_df$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

tree_data <- left_join(tree_data, node_df, by = "node") %>%
  mutate(ace_weight = ifelse(isTip, weight, ace_weight)) # assign weights to tips for plotting branch color

```

### iv. Plot tree

consider only labeling phyla with > 1% rel abundance 

```{r}

tree <- ggtree(genus_tree, branch.length = "none") %<+% tree_data +
  geom_tree(aes(color = ace_weight), size = 1) +
  
  scale_x_continuous(
  expand = expansion(mult = c(0, 0)),
  limits = c(NA, 45)  # or whatever max x value you need
) +
  
  scale_color_gradient2(low = "#7570b3", mid = "grey90", high = "#d95f02", midpoint = 0.5, name = "Model weight") +
  
  ggnewscale::new_scale_color() + # add a new color scale for the tip circles
  
  geom_tiplab(aes(label = Taxa), size = 3.5, offset = 0.5) +
  
  geom_point2(aes(subset = isTip, color = effect_direction), shape = 19, size = 4, stroke = 0, show.legend = FALSE) +

  scale_color_manual(values = c("positive" = "#1f78b4", "negative" = "#e41a1c", "neutral"  = "#666666"), name = "Effect direction") +
  
  geom_text2(aes(subset = isTip, label = case_when(effect_direction == "positive" ~ "+", effect_direction == "negative" ~ "–")),
  size = 3.6, vjust = 0.4, hjust = 0.5, color = "white") +
  
  theme_tree2() + 
  
  theme(legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()) +
  
  # Direction effect legend
  annotate("text", x = 1.2, y = 82, label = "Effect direction", size = 3.7, hjust = 0) +
  
  annotate("point", x = 1.5, y = 80.2, shape = 19, size = 4, color = "#1f78b4") +
  annotate("text", x = 2.1, y = 80.25, label = "positive", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 80.3, label = "+", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 78.4, shape = 19, size = 4, color = "#e41a1c") +
  annotate("text", x = 2.1, y = 78.45, label = "negative", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 78.5, label = "–", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 76.6, shape = 19, size = 4, color = "#666666") +
  annotate("text", x = 2.1, y = 76.65, label = "no effect", size = 3.3, hjust = 0) +
  
  # Bacteroidota
  annotate("segment", x = 32.5, xend = 32.5, y = 31, yend = 43, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 31, yend = 31, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 43, yend = 43, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 37, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 31, y = 1, yend = 1, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 1, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  # Proteobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 7, yend = 30, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 7, yend = 7, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 30, yend = 30, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 18, label = "Proteobacteria", size = 5.5, hjust = 0) +
  
  # Cyanobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 2, yend = 5, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 2, yend = 2, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 5, yend = 5, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 3.5, label = "Cyanobacteria", size = 5.5, hjust = 0) +
  
  #Fusobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 50, yend = 52, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 50, yend = 50, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 52, yend = 52, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 51, label = "Fusobacteriota", size = 5.5, hjust = 0) +
  
  #Actinobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 47, yend = 49, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 47, yend = 47, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 49, yend = 49, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 48, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 67, yend = 65, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 67, yend = 67, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 65, yend = 65, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 66, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  #Firmicutes
  annotate("segment", x = 32.5, xend = 32.5, y = 68, yend = 97, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 68, yend = 68, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 97, yend = 97, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 82.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 53, yend = 64, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 53, yend = 53, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 64, yend = 64, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 59.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  #Desulfobacterota
  annotate("segment", x = 32.5, xend = 32.5, y = 45, yend = 46, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 45, yend = 45, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 46, yend = 46, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 45.5, label = "Desulfobacterota", size = 5.5, hjust = 0) + 
  
  #Spirochaetota
  annotate("segment", x = 32.5, xend = 31, y = 44, yend = 44, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 44, label = "Spirochaetota", size = 5.5, hjust = 0)  +
  
  #Campylobacterota
  annotate("segment", x = 32.5, xend = 31, y = 6, yend = 6, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 6, label = "Campylobacterota", size = 5.5, hjust = 0) 
  
tree

ggsave("figures/tree.pdf", tree, width =8, height = 12.5)


```


## 6. Fit GAM's

## a. filter abundances

filter the abundances for the temp taxa (taxa where the model with temp fit best) that have a negative temp coefficient, 81 total taxa, to specifically explore if there is a "tipping point" or a point where abundance tends to drop off. 36 taxa where the smooth fits better. 

```{r}

negtemp_filter <- filtered_coef %>%
  filter(coef < 0) %>%
  pull(Taxa)
length(negtemp_filter) # 81 taxa 

# filter abundances for temp taxa
negtemp_abund <- da6 %>% filter(Taxa %in% negtemp_filter )
length(unique(negtemp_abund$Taxa))

```

## b. AIC comparison

Compare a m_main with linear temp predictor versus m_full with a smooth temp predictor.Then filter for taxa where m_full has > 0.5 of model weight. 

```{r}

# create empty list
gam_results <- list()

# loop
for (i in unique(negtemp_abund$Taxa)) {
   tryCatch({
    
    # subset by taxa 
    data <- droplevels(subset(negtemp_abund, Taxa == i))
    
    # make warning message
    warning_msg <- "No Warning"
    
    # models
    m_full <- gam(Abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log"))
    
    m_main <- gam(Abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log"))

   }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # Update the warning message
      } 
    })
  
      # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      #create data frame
      results_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE
      )
  
      # store results
      gam_results[[i]] <- results_df
    
}

# combine results 
gam_results <- do.call(rbind, gam_results)

################ Filter ############ 

#Filter for taxa where m_full best fit
gam_filter <- gam_results %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa)
length(gam_filter) # 36 taxa 
  
# filter abundance
smooth_abund <- negtemp_abund %>% filter(Taxa %in% gam_filter)
length(unique(smooth_abund$Taxa))

```

## c. Generate GAM preds

For the taxa where m_full fits best, generate model predictors to be able to cluster by shape in the next step and then plot. Generate model predictions on the log scale to plot all taxa on the same scale.

Need to check the full model output to check the convergence for some of these I think. 

```{r}

# Initialize an empty list to store results for all taxa
gam_preds <- list()

# Define the range of temp_s values for prediction
temp_seq <- seq(from = min(smooth_abund$temp_s, na.rm = TRUE), 
                to = max(smooth_abund$temp_s, na.rm = TRUE), length.out = 100)


# Iterate over unique values in final_temp_abund$Taxa
for (i in unique(smooth_abund$Taxa)) {
    
    # Subset your data based on current value of i
    data <- droplevels(subset(smooth_abund, Taxa == i))
    
    # run model 
    gam_m <- gam(Abundance ~ age_s + s(temp_s) + rain_s + fruit_s + group + sex + offset(log(filtered_reads)), 
                   data = data, method = "REML", family = nb(link = "log"))
    
    # Generate predictions over the specified range of temp_s
    pred_data <- data.frame(
      age_s = mean(data$age_s, na.rm = TRUE),
      temp_s = temp_seq,
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = unique(data$group)[1], # assuming group is a factor
      sex = unique(data$sex)[1], # assuming sex is a factor
      filtered_reads = mean(data$filtered_reads, na.rm = TRUE)
    )
    
    # Predict abundance values
    predicted_values <- predict(gam_m, newdata = pred_data)
    
    # Store the predicted values along with temperature sequence and Taxa label
    df <- data.frame(
      Taxa = i,
      temp_s = temp_seq,
      preds = predicted_values
    )
    
    # Store the results in the list
    gam_preds[[i]] <- df
    
}

# Combine all results into a single dataframe
gam_preds <- do.call(rbind, gam_preds)
row.names(gam_preds) <- NULL # clean up row names


```

## d. Cluster gam preds by shape 

compute pairwise distances and cluster using `hclust()` 

make supplemental figure with the three clusters

```{r}

# Create a wide format dataframe
predicted_wide <- reshape(gam_preds, idvar = "temp_s", timevar = "Taxa", direction = "wide")

# clean data
predicted_wide <- predicted_wide %>% 
  rename_with(~ str_replace(., "^preds.", ""), everything()) %>% # clean column names 
  select(-temp_s) # remove the temp_s column for correlation calculation and positive taxa

# elbow method
clust_plot <- fviz_nbclust(predicted_wide, kmeans, method = "wss", k.max = 10) + 
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank()) +
  labs(x = "Number of clusters", y = "Within-cluster sum of squares", title = "")

ggsave("figures/supp/fig3S.pdf", clust_plot, width = 6, height = 6)

# set number of clusters
k=3

# distnce matrix
clus.distance = as.dist(1 - cor(predicted_wide, use = "pa")) # create distance matrix

h.clust = hclust(clus.distance,method='complete') # calculate clusters 

pdf("figures/supp/fig5S.pdf", width = 10, height = 6)
dendogram <- plot(h.clust, cex = 0.6, main = "", xlab = "", ylab = "") # plot dendogram
rect.hclust(h.clust, k = k, border = 2:6) # add clusters 
dev.off()

# Separate temp-relationships into k clusters 
clusters <- cutree(h.clust, k = k)

# Add cluster assignments to the original data
gam_preds$Cluster <- clusters[match(gam_preds$Taxa, names(clusters))]

### Plot clusters ###

# clean data
clean_preds <- gam_preds %>% mutate(Cluster = as.factor(Cluster))

# by cluster 
c1 <- clean_preds %>% filter(Cluster == 1)
c2 <- clean_preds %>% filter(Cluster == 2)
c3 <- clean_preds %>% filter(Cluster == 3)

p1 <- ggplot(c1, aes(x = temp_s, y = preds, color = Taxa)) + 
  geom_line() +
  theme_minimal()
p1

p2 <- ggplot(c2, aes(x = temp_s, y = preds, color = Taxa)) + 
  geom_line() +
  theme_minimal()
p2

p3 <- ggplot(c3, aes(x = temp_s, y = preds, color = Taxa)) + 
  geom_line() +
  theme_minimal()
p3

# Arrange the plots in a 2x2 layout
cluster_plot <- ggarrange(p1, p2,
                           ncol = 1, nrow = 2,
                          labels = c("A", "B"))
```

phylum check 

```{r}

gam_phyla <- c1 %>%
  select(Taxa) %>%
  distinct() %>%
  left_join(phyla, by = "Taxa")


```


## e. plot 

Plot predictions

```{r}

mean_temp <- mean(smooth_abund$temp)
sd_temp <- sd(smooth_abund$temp) 

gam_plot <- ggplot(c1, aes(x = temp_s, y = preds, color = Taxa)) + 
  geom_line() +
  theme_minimal() +
  labs(y = "Predicted log abundance", x = "Temperature (°C)") +
  scale_x_continuous(
   breaks = seq(min(c1$temp_s), max(c1$temp_s), length.out = 6), 
   labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 16, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 

gam_plot

ggsave("figures/fig4.pdf", gam_plot, width =6, height = 6)



```

## phyla plot

8 out of 24 of these are firmicutes but 6 are bacteroidta so I don't think there is a strong phylum trend here

```{r}

length(unique(c1$Taxa))

c1_taxa <- c1 %>%
  select(Taxa) %>%
  unique() %>%
  left_join(phyla, by = "Taxa") %>%
  group_by(Phylum) %>%
  summarise(phyla_count = n())

```

## g. supplement

Plot the smooths for each taxa with the CI's for the supplement

```{r}

# Start PDF
pdf("scripts/compiled/gm_outcome/figures_small/all_smooths.pdf", width = 8, height = 11)  

# Iterate over unique Taxa
for (j in seq(1, length(unique(smooth_abund$Taxa)), by = 4)) {
    
    # Set up a 3x3 grid 
    par(mfrow = c(2, 2))  
    
    # Loop through six taxa at a time
    for (i in unique(smooth_abund$Taxa)[j:min(j + 3, length(unique(smooth_abund$Taxa)))]) {
        
        # Subset your data based on the current value of i
        data <- droplevels(subset(smooth_abund, Taxa == i))
        
        # Fit model
        gam_m <- gam(Abundance ~ age_s + s(temp_s) + rain_s + group + sex + s(ID, bs = "re") + offset(log(filtered_reads)), 
                     data = data, method = "REML", family = nb(link = "log"))
        
        # Plot the smooth effect for temperature
        plot(gam_m, select = 1, rug = TRUE, shade = TRUE, main = paste(i, "- Smooth for temp"))
    }
}

# Close device
dev.off()

```


# 7. Neopterin * temp interaction 

## a. Community feature analysis

### i. model community features

Compare m_main that has the base predictors plus the temp predictor and m_full which has an interaction between neopterin and temperature. Both model include an offset of total reads to control for seq depth. Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed.

Fit observed model outside of the loop because loop environment doesn't have enough memory to let model converge

```{r}

# create empty lists
com_coef <- list() 
com_aic <- list()

com_features <- colnames(m_model)[c(15, 17:20)]# list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(filtered_reads))"))
      
    m_main <- lmer(formula_main, data = m_model)
    m_full <- lmer(formula_full, data = m_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[10,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[10,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
com_aic <- do.call(rbind, com_aic)
com_coef <- do.call(rbind, com_coef)


#### model observed ######

m_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = m_model)

m_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = m_model)

AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(m_full)$coeff[10,1], 4)
std <- round(summary(m_full)$coeff[10,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

com_coef <- rbind(com_coef, results_df)

com_aic <- rbind(com_aic, aic_df) 
```

### ii. plot community feature

```{r}

plot_com_coef <- com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric)) 
         
p1 <- ggplot(plot_com_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 18, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1

# Reshape data
com_data_long <- com_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 18, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        axis.text.y = element_blank()) +
  annotate("text", x = -0.5, y = "PC1", label = "interaction model", color = "white", size = 5, fontface = "bold") +
  annotate("text", x = 0.3, y = "PC1", label = "base model", color = "#fdb462", size = 5, fontface = "bold")
p2  

# add plots together  
combo_com <- ggarrange(p1, p2, ncol = 2, nrow = 1, labels = c("A", "B"))

ggsave("figures/fig5.pdf", combo_com, width =12, height = 6)

```


## b. Taxa analysis

### i. AIC comparison

Take all temp taxa (taxa that had a relationship with temp) and fit with m_main model with all the base predictors plus temp and m_full model with base predictors and a temp * neo interaction. 

```{r}

# filter abundance for temp taxa
temp_abund <- da6 %>% filter(Taxa %in% temp_filter )
length(unique(temp_abund$Taxa)) # 93

# create lists
neo_coef <- list()
neo_aic <- list()
neo_model_results <- list()

for (i in unique(temp_abund$Taxa)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(temp_abund, Taxa == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- gam(Abundance ~ age_s + temp_s * neo_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 

      m_main <- gam(Abundance ~ age_s + rain_s + temp_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 
      
    }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
     # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      # extract temp coef
      coef <- round(summary(m_full)$p.coeff[9], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[9], 4)
    
      # clac CI'
      ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
      ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
      # bind data
      df <- cbind(ci_95, ci_50)
      
      #create data frames
    aic_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
        Taxa = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3),
        stringsAsFactors = FALSE)

    options(scipen = 999)  

    model_results_df <- data.frame(
      Taxa = i,
      Predictor = names(summary(m_full)$p.coeff[c(2:9)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:9)], 4),
      StdError = round(summary(m_full)$se[c(2:9)], 4),
      stringsAsFactors = FALSE)
    
    
      # save results
    neo_aic[[i]] <- aic_df
    neo_coef[[i]] <- results_df
    neo_model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
neo_aic <- do.call(rbind, neo_aic)
neo_coef <- do.call(rbind, neo_coef)
neo_model_results <- do.call(rbind, neo_model_results)

```

### ii. taxa selection

same as previous AIC taxa filtered. Remove convergence warnings and any taxa that look like they didn't converge from SE. Filter for taxa where m_full (model with temp * neo interaction) have > .50 model weight

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(neo_aic$Taxa[neo_aic$Warning == "Convergence Warning"]) # 2 warnings

# remove all taxa where one or more predictos have huge SE
convergence_errors <- neo_model_results %>%
  filter(StdError > 100) %>%
  pull(Taxa) %>%
  unique()

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 2 do not converge

# remove error taxa
temp_abund2 <- temp_abund %>% filter(!Taxa %in% errors)
length(unique(temp_abund2$Taxa)) # 91 remaining taxa

neo_coef2 <- neo_coef %>% filter(!Taxa %in% errors)
length(neo_coef2$Taxa)

neo_aic2 <- neo_aic %>% filter(!Taxa %in% errors)
length(neo_aic2$Taxa) # 180/2 = 91

# filter for taxa where full model fits best
neo_filter <- neo_aic2 %>%
  group_by(Taxa) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa) 
length(neo_filter) # 21 taxa

# filter abundances for temp taxa
neo_abund <- temp_abund2 %>% filter(Taxa %in% neo_filter)
length(unique(neo_abund$Taxa)) 

filtered_neo_coef <- neo_coef2 %>% filter(Taxa %in% neo_filter)
length(filtered_neo_coef$Taxa)

filtered_neo_aic <- neo_aic2 %>% filter(Taxa %in% neo_filter)
length(filtered_neo_aic$Taxa) 

```

look at phylum 

```{r}

phyla_neo <- filtered_neo_coef %>% left_join(phyla, by = "Taxa") %>%
  mutate(Phylum_grouped = case_when(str_detect(Phylum, "Firmicutes") ~ "Firmicutes",
         TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarize(count = n())

level <- da6 %>%
  select(Taxa, level, Phylum) %>%
  distinct()

neo_taxa_level <- filtered_neo_coef %>%
  left_join(level, by = "Taxa")

```


### iii. plot 

work on big multi-panel and think about taxa with large CI's 
add back in family versus genus? 

```{r}

plot_neo_coef <- filtered_neo_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(Taxa, levels = Taxa)) 
         
p1 <- ggplot(plot_neo_coef, aes(x = order, y = coef, color = direction)) +
  geom_hline(yintercept = 0, linetype = "dashed", size = .5, color = "grey") +
  geom_point(size = 3) +
  geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
  geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
  scale_color_manual(values = c("#80b1d3", "#fb8072"), breaks = c("positive", "negative")) + 
  coord_flip() +
  labs(y = "Effect", x = "") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 18, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 
p1

# Reshape data
aic_long <- filtered_neo_aic %>%
  select(Taxa, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(Taxa, levels = levels(plot_neo_coef$order))) %>%
  arrange(order)


# Plot model weights
p2 <- ggplot(aic_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 20, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        axis.text.y = element_blank()) +
  annotate("text", x = -0.75, y = "Alistipes_A_871400", label = "interaction model", color = "white", size = 4, fontface = "bold") +
  annotate("text", x = 0.2, y = "Alistipes_A_871400", label = "base model", color = "#fdb462", size =4, fontface = "bold")
p2  

# add plots together  
combo_taxa <- ggarrange(p1, p2, ncol = 2, nrow = 1, labels = c("A", "B"))

ggsave("figures/fig6.pdf", combo_taxa, width =15, height = 7)


```

## c. Interaction co-plot

### i. get model predictions

```{r}

# Create neo bins
neo_abund$neo_bin <- cut(neo_abund$neo_s, breaks = 4, labels = c("Bin1", "Bin2", "Bin3", "Bin4"), include.lowest = TRUE)
bin_limits <- levels(cut(neo_abund$neo_s, breaks = 4, include.lowest = TRUE)) # look at bin range
bin_avg <- tapply(neo_abund$neo_s, neo_abund$neo_bin, mean, na.rm = TRUE) # use for co plot

# Create temp seq
temp_seq <- seq(min(neo_abund$temp_s, na.rm = TRUE), max(neo_abund$temp_s, na.rm = TRUE), length.out = 100)

# list
neo_preds <- list()

# loop
for (i in unique(neo_abund$Taxa)) {
  
  data <- droplevels(subset(neo_abund, Taxa == i))
  
  m <- gam(Abundance ~ age_s + temp_s * neo_s + rain_s + rain_s + sex + group + fruit_s + offset(log(filtered_reads)), 
           data = data, 
           method = "REML", 
           family = nb(link = "log"))
  
  coef <- round(summary(m)$p.coeff[9], 4)
  
  # neo_s values and bins
  neo_values <- c(-0.1250571,  0.7789195,  1.7335661,  2.6339655)
  bins <- paste0("neopterin_avg", 1:4)
  
  # Generate predictions for each neo_s value
  preds <- lapply(seq_along(neo_values), function(j) {
    neo_s <- neo_values[j]
    bin <- bins[j]
    
    # preds dataset
    pred <- expand.grid(
      temp_s = temp_seq,
      neo_s = neo_s,
      age_s = mean(data$age_s, na.rm = TRUE),
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = unique(data$group)[1],
      sex = unique(data$sex)[1],
      filtered_reads = mean(data$filtered_reads, na.rm = TRUE)
    )
    
    # get model predictions
    preds_model <- predict(m, newdata = pred, type = "link", se.fit = TRUE)
    
    # store preds and CI's 
    df <- data.frame(
      temp_s = pred$temp_s,  
      neo_s = neo_s,
      fit = preds_model$fit,
      lower = preds_model$fit - 1.96 * preds_model$se.fit,
      upper = preds_model$fit + 1.96 * preds_model$se.fit,
      bin = bin,
      coef = coef
    )
    
    return(df)
  })
  
  # combine predictions for all bins 
  preds_df <- do.call(rbind, preds)
  preds_df$Taxa <- i 
  
  # store preds
  neo_preds[[i]] <- preds_df
}

# Combine all predictions into a single dataframe
neo_preds <- do.call(rbind, neo_preds)
row.names(neo_preds) <- NULL # clean up row names

```

### ii. plot by Taxa

Plot with the CI's for each taxa for the supplement

```{r}

pdf("figures/taxa_interaction.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(neo_preds$Taxa)) {
  
  data <- subset(neo_preds, Taxa == i)

  p <- ggplot(data, aes(x = temp_s, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "grey") +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(
      title = paste("predicted Abundance for", i),
      x = "temperature (scaled)",
      y = "predicted log abundance") 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

### iii. Plot by effect direction

```{r}

########### Prep data ################

mean_neo <- mean(neo_abund$neo)
sd_neo <- sd(neo_abund$neo) 

neo_reverse <- function(x) {
  return(round(x * 2 * sd_neo + mean_neo))  # reverse scale in figure axes
}

bin_limits <- levels(cut(neo_abund$neo_s, breaks = 4, include.lowest = TRUE)) # look at bin range
neo_reverse(3.26)


neo_preds2 <- neo_preds %>% mutate(direction = ifelse(coef > 0, "positive", "negative"),
                                       direction = case_when(Taxa == "Traorella"~ "neg_pos",
                                                             Taxa == "CAG-317_146760" ~ "neg_pos",
                                                             Taxa == "Eubacterium_I" ~ "neg_pos",
                                                             Taxa == "Erysipelotrichaceae"~ "neg_pos",
                                                             TRUE ~ direction),
                                      bin = case_when(bin == "neopterin_avg1" ~ "neopterin: 108-1464 (nmol/L)",
                                                   bin == "neopterin_avg2" ~ "neopterin: 1464-2818 (nmol/L)",
                                                   bin == "neopterin_avg3" ~ "neopterin: 2818-4164 (nmol/L)",
                                                   bin == "neopterin_avg4" ~ "neopterin: 4164-5525 (nmol/L)")) 




pos <- neo_preds2 %>% filter(direction == "positive")
neg_pos <- neo_preds2 %>% filter(direction == "neg_pos")
neg <- neo_preds2 %>% filter(direction == "negative")

length(unique(pos$Taxa))
length(unique(neg_pos$Taxa))
length(unique(neg$Taxa))

# bin values
unique_bins <- unique(neo_preds2$bin)

# to rescale x axis
mean_temp <- mean(neo_abund$temp)
sd_temp <- sd(neo_abund$temp) 

################## POSITIVE ###########################
y_min <- min(pos$fit)
y_max <- max(pos$fit)

pos_list <- list()

# loop
for (i in unique_bins) {
  
  pos_filtered <- pos[pos$bin == i, ]
  
  # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 108-1464 (nmol/L)") {
    "predicted log abundance"
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 108-1464 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(pos_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = y_axis_label, x = "", title = paste(i)) +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  pos_list[[as.character(i)]] <- bin_plot
  
}

pos_plot <- do.call(ggarrange, c(pos_list, nrow = 1))
pos_plot


################### NEG PLOS PLOT ######################

y_min <- min(neg_pos$fit)
y_max <- max(neg_pos$fit)

neg_pos_list <- list()

# loop
for (i in unique_bins) {
  
  neg_pos_filtered <- neg_pos[neg_pos$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 108-1464 (nmol/L)") {
    "predicted log abundance"
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 108-1464 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(neg_pos_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  neg_pos_list[[as.character(i)]] <- bin_plot
  
}

neg_pos_plot <- do.call(ggarrange, c(neg_pos_list, nrow = 1))
neg_pos_plot



################### NEG PLOT ######################

y_min <- min(neg$fit)
y_max <- max(neg$fit)

neg_list <- list()

# loop
for (i in unique_bins) {
  
  neg_filtered <- neg[neg$bin == i, ]
  
   # set y-axis labels for bin 1
  y_axis_label <- if (i == "neopterin: 108-1464 (nmol/L)") {
    "predicted log abundance"
  } else {
    ""
  }
  
  y_tick_labels <- if (i == "neopterin: 108-1464 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(neg_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = y_axis_label, x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    scale_x_continuous(
      breaks = seq(min(neo_preds2$temp_s), max(neo_preds2$temp_s), length.out = 3),
      labels = inverse_scale) +
      theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none")
  
  neg_list[[as.character(i)]] <- bin_plot
  
}

neg_plot <- do.call(ggarrange, c(neg_list, nrow = 1))
neg_plot


######## combine ##########

combined_plot <- ggarrange(neg_plot, pos_plot, neg_pos_plot, 
                           nrow = 3, ncol = 1, 
                           labels = c("A", "B", "C"))
combined_plot

ggsave("figures/fig7.pdf", combined_plot, width =13, height = 9)


```




















# Extra code

test loop code 

```{r}

t <- da5 %>% filter(Taxa == "Morganella" | Taxa == "Anaerobiospirillum_A")

m_full <- gam(Abundance ~ age_s + temp_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = t, method = "REML", family = nb(link = "log")) 

 summary(m_full)
 
m_main <- gam(Abundance ~ age_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = t, method = "REML", family = nb(link = "log")) 

 # extract taxa coef
      coef <- round(summary(m_full)$p.coeff[3], 4)
    
      # extract ci's
      std <- round(summary(m_full)$se[3], 4)

      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)  
      
      t <- da5 %>% filter(Taxa == "Morganella")

m_full <- gam(Abundance ~ age_s + temp_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = t, method = "REML", family = nb(link = "log")) 

m_main <- gam(Abundance ~ age_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = t, method = "REML", family = nb(link = "log")) 

# extract temp coef
      coef <- round(summary(m_full)$p.coeff[3], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[3], 4)
      
      
      model_results_df <- data.frame(
      Taxa = "Morganella",
      Predictor = names(summary(m_full)$p.coeff[c(2:8)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:8)], 4),
      StdError = round(summary(m_full)$se[c(2:8)], 4),
      stringsAsFactors = FALSE)

```



Manual models of community features

```{r}

hist(m_model$PC1)
hist(m_model$PC2)
hist(m_model$PC3)
hist(m_model$PC4)
hist(m_model$shannon)
hist(m_model$observed)


s_base <- lmer(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)

summary(s_temp)

s_temp <- lmer(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)   

AICctab(s_base, s_temp, weights = TRUE)
AICtab(s_base, s_temp, weights = TRUE)


o_base <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(total_reads)), 
                family = Gamma(link = "log"),
                data = m_model)

o_temp <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID)+ offset(log(total_reads)), 
                family = Gamma(link = "log"),
                data = m_model)
summary(o_temp)

AICctab(o_base, o_temp, weights = TRUE)
AICtab(o_base, o_temp, weights = TRUE)
               

pc1_base <- lmer(PC1 ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)

pc1_temp <- lmer(PC1 ~ group + age_s + sex + fruit_s +  residuals_s +rain_s + temp_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)   

AICctab(pc1_base, pc1_temp, weights = TRUE)
AICtab(pc1_base, pc1_temp, weights = TRUE)

pc2_base <- lmer(PC2 ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)
summary(pc2_base)

pc2_temp <- lmer(PC2 ~ group + age_s + sex + fruit_s +  residuals_s + rain_s + temp_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)   
summary(pc2_temp)

AICctab(pc2_base, pc2_temp, weights = TRUE)
AICtab(pc2_base, pc2_temp, weights = TRUE)


pc3_base <- lmer(PC3 ~ group + age_s + sex + fruit_s + residuals_s + rain_s +  (1 | ID) + offset(log(total_reads)), 
               data = m_model)

pc3_temp <- lmer(PC3 ~ group + age_s + sex + fruit_s +  residuals_s + rain_s + temp_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)   

AICctab(pc3_base, pc3_temp, weights = TRUE)
AICtab(pc3_base, pc3_temp, weights = TRUE)


pc4_base <- lmer(PC4 ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)

pc4_temp <- lmer(PC4 ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(total_reads)), 
               data = m_model)   

AICctab(pc4_base, pc4_temp, weights = TRUE)
AICtab(pc4_base, pc4_temp, weights = TRUE)


simpson <- estimate_richness(physeq2, measures = "Simpson") 
simpson <- simpson %>%
  rownames_to_column(var = "sample") %>%
  rename(simpson = Simpson)


 Plot model weights
p2 <- ggplot(com_data_long, aes(y = order)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank(),
        axis.text.y = element_blank()) +
  annotate("text", x = -0.8, y = "PC1", label = "interaction model", color = "#bc80bd", size = 4, fontface = "bold") +
  annotate("text", x = 0.65, y = "PC1", label = "base model", color = "#fdb462", size = 4, fontface = "bold")
p2   

# add plots together  
combo <- ggarrange(p1, p2, ncol = 2, nrow = 1)


# Reshape data
aic_long <- filtered_aic %>%
  select(Taxa, model, weight) %>%
  spread(key = model, value = weight)
          
# Plot model weights
p2 <- ggplot(aic_long, aes(y = Taxa)) +
  # Plot m_full (on the left side)
  geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
  # Plot m_main (on the right side)
  geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("m_full" = "#fb8072", "m_main" = "#80b1d3")) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),   
    labels = c("1", "0.5", "0", "0.5", "1"),
    expand = c(0.05, 0.05)) +
  labs(x = "Model Weight", y= "", fill = "Model") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 10, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) +
  annotate("text", x = -0.35, y = "Shannon", label = "full model", color = "#fb8072", size = 4, fontface = "bold") +
  annotate("text", x = 0.93, y = "Shannon", label = "base model", color = "#80b1d3", size = 4, fontface = "bold")
  
p2
ggsave("scripts/compiled/gm_outcome/figures/taxa_temp_weight.pdf", p2, width = 8, height = 16)

```

## e. Generate linear preds

```{r}

# Initialize an empty list to store results for all taxa
lin_preds <- list()

# Define the range of temp_s values for prediction
temp_seq <- seq(from = min(lin_abund$temp_s, na.rm = TRUE), 
                to = max(lin_abund$temp_s, na.rm = TRUE), length.out = 100)

# Iterate over unique values in final_temp_abund$Taxa
for (i in unique(lin_abund$Taxa)) {
    
    # Subset your data based on current value of i
    data <- droplevels(subset(lin_abund, Taxa == i))
    
    # run model 
    gam_m <- gam(Abundance ~ age_s + temp_s + rain_s + fruit_s + group + sex + offset(log(filtered_reads)), 
                   data = data, method = "REML", family = nb(link = "log"))
    
    # Generate predictions over the specified range of temp_s
    pred_data <- data.frame(
      age_s = mean(data$age_s, na.rm = TRUE),
      temp_s = temp_seq,
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = unique(data$group)[1], # assuming group is a factor
      sex = unique(data$sex)[1], # assuming sex is a factor
      filtered_reads = mean(data$filtered_reads, na.rm = TRUE)
    )
    
    # Predict abundance values
    predicted_values <- predict(gam_m, newdata = pred_data)
    
    # Store the predicted values along with temperature sequence and Taxa label
    df <- data.frame(
      Taxa = i,
      temp_s = temp_seq,
      preds = predicted_values
    )
    
    # Store the results in the list
    lin_preds[[i]] <- df
    
}

# Combine all results into a single dataframe
lin_preds <- do.call(rbind, lin_preds)
row.names(lin_preds) <- NULL # clean up row names


lin_plot <- ggplot(lin_preds, aes(x = temp_s, y = preds, group = Taxa)) + 
  geom_line(color = "#80b1d3") +
  theme_minimal() +
  labs(y = "", x = "temperature (°C)") +
  #scale_x_continuous(
    #breaks = seq(min(lin_preds$temp_s), max(lin_preds$temp_s), length.out = 6), 
    #labels = inverse_scale) + 
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        text = element_text(size = 10, color = "black", family = "sans"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none",
        panel.background = element_blank()) 

lin_plot

combo <- ggarrange(gam_plot, lin_plot,
          ncol = 2, nrow = 1,
          labels = c("A", "B"))


###### Neg to Positve ##########

neg_pos_plot <- ggplot(neg_pos, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(x = "Temperature (scaled)", y = "Predicted Abundance") 


######## negative ########

neg_plot <- ggplot(neg, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(x = "Temperature (scaled)", y = "Predicted Abundance") 

```

tree code 

```{r}

tree_plot + 
  geom_tiplab(aes(color = taxa_temp_sensitive), size = 3) +
  geom_nodepoint(aes(color = node_temp_sensitive), size = 1.5) + 
  scale_color_manual(values = c("no" = "#80b1d3", "yes" = "#fb8072")) +  
  theme_tree()

```


```{r}


# 21 phylum

## add togther firmicutes ## 

firmicutes_values <- phylum_otu_table %>%
  rownames_to_column(var = "Phylum")  %>%
  filter(grepl("Firmicutes", Phylum)) %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(Phylum = "Firmicutes")  # Add taxon label

phylum_otu <- phylum_otu_table %>%
  rownames_to_column(var = "Phylum") %>%
  filter(!str_detect(Phylum, "Firmicutes")) 
  
phylum_otu2 <- rbind(phylum_otu, firmicutes_values)

```

