---
title: "Untitled"
author: "Lucore"
date: "2025-04-24"
output: html_document
--

####################### CURRENT CODE #########################
# Data prep

extract the necessary metadata and filter the phyloseq object for all genera included in the DA temp analysis. Then extract the tree.

```{r}

# get list of taxa with ne
direction <- filtered_coef %>% 
  mutate(effect_direction = ifelse(coef > 0, "positive", "negative")) %>%
  select(Taxa, effect_direction)

# add temp information to da table
da7 <- da6 %>%
  mutate(taxa_temp_sensitive = ifelse(Taxa %in% temp_filter, "sensitive", "nuetral")) %>%
  left_join(direction, by = "Taxa") %>%
  mutate(effect_direction = case_when(
    is.na(effect_direction) ~ "neutral",
    TRUE ~ effect_direction))
  
# get genus names
genus_names <- da7 %>%
  filter(level == "Genus") %>%
  pull(Taxa) %>%
  unique()

# aggregate by taxa
genus_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered, "Genus"))

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
temp_genus_phylo <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)
genus_tax <- as.data.frame(tax_table(temp_genus_phylo))

# prep final tree metadata 
genus_meta <- da7 %>%
  filter(level == "Genus") %>%
  select(Taxa, taxa_temp_sensitive, effect_direction) %>%
  distinct() %>%
  left_join(taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight), by = "Taxa")

# create phylum groups for clade labels
phylum_groups <- genus_tax %>%
  rownames_to_column("OTU") %>%
  mutate(Genus = as.character(Genus), Phylum = as.character(Phylum)) %>%
  filter(!is.na(Phylum), !is.na(Genus)) %>%
  group_by(Phylum) %>%
  summarise(genus_list = list(Genus)) %>%
  unnest_longer(genus_list, values_to = "Genus")

```


# assign tip labels

First assign the tip labels as genus names instead of otu numbers. Add the metadata for temp sensitivity and change the tip labels to the model weight varible

```{r}

# tip labels are still otu numbers
head(genus_tree$tip.label)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels
  
# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# add phylum node assignment to tree data
tree_data <- left_join(tree_data, phylum_nodes, by = "node")

# now make tip labels the model weight
genus_tree$tip.label <- tip_data$weight
head(genus_tree$tip.label)

```

# ancestral state reconstruction

reconstruction returns the maximum likelihood estimates of the model weights for each node, assumes brownian motion. Higher weights are more temperature sensitive than lower weights. 

```{r}

# run ace function
ace_result <- ace(tip_data$weight, genus_tree, type = "continuous")

# extract node probs
node_df <- data.frame(
  node = as.numeric(names(ace_result$ace)),
  ace_weight = ace_result$ace) # ignore warning will fix in next step

# assign node probs to node
node_df$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

tree_data <- left_join(tree_data, node_df, by = "node") %>%
  mutate(ace_weight = ifelse(isTip, weight, ace_weight)) # assign weights to tips for plotting branch color

```


# Plot tree

```{r}

tree <- ggtree(genus_tree, branch.length = "none") %<+% tree_data +
  geom_tree(aes(color = ace_weight), size = 1) +
  
  scale_x_continuous(
  expand = expansion(mult = c(0, 0)),
  limits = c(NA, 45)  # or whatever max x value you need
) +
  
  scale_color_gradient2(low = "#7570b3", mid = "grey90", high = "#d95f02", midpoint = 0.5, name = "Model weight") +
  
  ggnewscale::new_scale_color() + # add a new color scale for the tip circles
  
  geom_tiplab(aes(label = Taxa), size = 3.5, offset = 0.5) +
  
  geom_point2(aes(subset = isTip, color = effect_direction), shape = 19, size = 4, stroke = 0, show.legend = FALSE) +

  scale_color_manual(values = c("positive" = "#1f78b4", "negative" = "#e41a1c", "neutral"  = "#666666"), name = "Effect direction") +
  
  geom_text2(aes(subset = isTip, label = case_when(effect_direction == "positive" ~ "+", effect_direction == "negative" ~ "–")),
  size = 3.6, vjust = 0.4, hjust = 0.5, color = "white") +
  
  theme_tree2() + 
  
  theme(legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()) +
  
  # Direction effect legend
  annotate("text", x = 1.2, y = 82, label = "Effect direction", size = 3.7, hjust = 0) +
  
  annotate("point", x = 1.5, y = 80.2, shape = 19, size = 4, color = "#1f78b4") +
  annotate("text", x = 2.1, y = 80.25, label = "positive", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 80.3, label = "+", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 78.4, shape = 19, size = 4, color = "#e41a1c") +
  annotate("text", x = 2.1, y = 78.45, label = "negative", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 78.5, label = "–", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 76.6, shape = 19, size = 4, color = "#666666") +
  annotate("text", x = 2.1, y = 76.65, label = "no effect", size = 3.3, hjust = 0) +
  
  # Bacteroidota
  annotate("segment", x = 32.5, xend = 32.5, y = 31, yend = 43, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 31, yend = 31, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 43, yend = 43, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 37, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 31, y = 1, yend = 1, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 1, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  # Proteobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 7, yend = 30, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 7, yend = 7, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 30, yend = 30, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 18, label = "Proteobacteria", size = 5.5, hjust = 0) +
  
  # Cyanobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 2, yend = 5, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 2, yend = 2, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 5, yend = 5, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 3.5, label = "Cyanobacteria", size = 5.5, hjust = 0) +
  
  #Fusobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 50, yend = 52, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 50, yend = 50, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 52, yend = 52, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 51, label = "Fusobacteriota", size = 5.5, hjust = 0) +
  
  #Actinobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 47, yend = 49, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 47, yend = 47, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 49, yend = 49, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 48, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 67, yend = 65, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 67, yend = 67, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 65, yend = 65, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 66, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  #Firmicutes
  annotate("segment", x = 32.5, xend = 32.5, y = 68, yend = 97, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 68, yend = 68, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 97, yend = 97, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 82.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 53, yend = 64, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 53, yend = 53, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 64, yend = 64, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 59.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  #Desulfobacterota
  annotate("segment", x = 32.5, xend = 32.5, y = 45, yend = 46, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 45, yend = 45, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 46, yend = 46, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 45.5, label = "Desulfobacterota", size = 5.5, hjust = 0) + 
  
  #Spirochaetota
  annotate("segment", x = 32.5, xend = 31, y = 44, yend = 44, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 44, label = "Spirochaetota", size = 5.5, hjust = 0)  +
  
  #Campylobacterota
  annotate("segment", x = 32.5, xend = 31, y = 6, yend = 6, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 6, label = "Campylobacterota", size = 5.5, hjust = 0) 
  
tree

ggsave("figures/tree.pdf", tree, width =8, height = 12.5)


```




################ OLD CODE ###############
### i. data prep

extract the necessary metadata and filter the phyloseq object for all genera included in the DA temp analysis. Then extract the tree. 

```{r}

# get list of taxa with ne
direction <- filtered_coef %>% 
  mutate(effect_direction = ifelse(coef > 0, "positive", "negative")) %>%
  select(Taxa, effect_direction)

# add temp information to da table
da7 <- da6 %>%
  mutate(taxa_temp_sensitive = ifelse(Taxa %in% temp_filter, "sensitive", "nuetral")) %>%
  left_join(direction, by = "Taxa") %>%
  mutate(effect_direction = case_when(
    is.na(effect_direction) ~ "neutral",
    TRUE ~ effect_direction))
  
# get genus names
genus_names <- da7 %>%
  filter(level == "Genus") %>%
  pull(Taxa) %>%
  unique()

# aggregate by taxa
genus_aggregated <- suppressMessages(phyloseq::tax_glom(phylo_filtered, "Genus"))

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
temp_genus_phylo <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)
genus_tax <- as.data.frame(tax_table(temp_genus_phylo))

# prep final tree metadata 
genus_meta <- da7 %>%
  filter(level == "Genus") %>%
  select(Taxa, taxa_temp_sensitive, effect_direction) %>%
  distinct() %>%
  left_join(taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight), by = "Taxa")

```


### ii. compile tree info

# assign tip labels

First assign the tip labels as genus names instead of otu numbers. Add the metadata for temp sensitivity and change the tip labels to the model weight varible

```{r}

# tip labels are still otu numbers
head(genus_tree$tip.label)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels

# create phylum groups for node labels
phylum_groups <- genus_tax %>%
  rownames_to_column("OTU") %>%
  mutate(Genus = as.character(Genus), Phylum = as.character(Phylum)) %>%
  filter(!is.na(Phylum), !is.na(Genus)) %>%
  group_by(Phylum) %>%
  summarise(genus_list = list(Genus)) %>%
  unnest_longer(genus_list, values_to = "Genus")

# group together firmicutes 
phylum_groups <- phylum_groups %>%
  mutate(Phylum_Grouped = case_when(
    grepl("Firmicutes", Phylum) ~ "Firmicutes",
    TRUE ~ Phylum
  )) %>%
  group_by(Phylum_Grouped) %>%
  summarise(genus_list = list(unique(unlist(genus_list)))) 
  #unnest_longer(genus_list, values_to = "Genus") # two phyla Campylobacterota and Spirochaetota only have one genus so there is no node assignment 
  
# calcultae phylum node
phylum_nodes <- phylum_groups %>%
  mutate(node = purrr::map(genus_list, ~ getMRCA(genus_tree, .x))) %>%
  filter(Phylum != "Campylobacterota" & Phylum != "Spirochaetota") %>%
  mutate(node = as.integer(node)) %>%
  select(Phylum, node)
  
# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# add phylum node assignment to tree data
tree_data <- left_join(tree_data, phylum_nodes, by = "node")

# now make tip labels the model weight
genus_tree$tip.label <- tip_data$weight
head(genus_tree$tip.label)

```

# ancestral state reconstruction

reconstruction returns the maximum likelihood estimates of the model weights for each node, assumes brownian motion. Higher weights are more temperature sensitive than lower weights. 

```{r}

# run ace function
ace_result <- ace(tip_data$weight, genus_tree, type = "continuous")

# extract node probs
node_df <- data.frame(
  node = as.numeric(names(ace_result$ace)),
  ace_weight = ace_result$ace) # ignore warning will fix in next step

# assign node probs to node
node_df$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

tree_data <- left_join(tree_data, node_df, by = "node")


```

# plot tree

```{r}

library(ggnewscale)

tree_data <- tree_data %>%
  mutate(ace_weight = ifelse(isTip, weight, ace_weight))

```



```{r}

tree <- ggtree(genus_tree, branch.length = "none") %<+% tree_data +
  geom_tree(aes(color = ace_weight), size = 1) +
  
  scale_color_gradient(low = "#bc80bd", high = "#fdb462", name = "Model weight") +
  
  ggnewscale::new_scale_color() + # add a new color scale for the tip circles
  
  geom_tiplab(aes(label = Taxa), size = 3.5, offset = 0.5) +
  
  geom_point2(aes(subset = isTip, color = effect_direction), shape = 19, size = 4, stroke = 0, show.legend = FALSE) +

  scale_color_manual(values = c("positive" = "#80b1d3", "negative" = "#fb8072", "neutral"  = "grey60"), name = "Effect direction") +
  
  geom_text2(aes(subset = isTip, label = case_when(effect_direction == "positive" ~ "+", effect_direction == "negative" ~ "–")),
  size = 3.6, vjust = 0.4, hjust = 0.5, color = "white") +
  
  theme_tree2() +
  
  theme(legend.position = c(0, 1),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()) +
  
  # Bacteroidota
  annotate("segment", x = 39, xend = 39, y = 31, yend = 43, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 31, yend = 31, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 43, yend = 43, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 37, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 39, xend = 37.5, y = 1, yend = 1, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 1, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  # Proteobacteria
  annotate("segment", x = 39, xend = 39, y = 7, yend = 30, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 7, yend = 7, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 30, yend = 30, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 18, label = "Proteobacteria", size = 5.5, hjust = 0) +
  
  # Cyanobacteria
  annotate("segment", x = 39, xend = 39, y = 2, yend = 5, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 2, yend = 2, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 5, yend = 5, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 3.5, label = "Cyanobacteria", size = 5.5, hjust = 0) +
  
  #Fusobacteriota
  annotate("segment", x = 39, xend = 39, y = 50, yend = 52, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 50, yend = 50, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 52, yend = 52, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 51, label = "Fusobacteriota", size = 5.5, hjust = 0) +
  
  #Actinobacteriota
  annotate("segment", x = 39, xend = 39, y = 47, yend = 49, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 47, yend = 47, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 49, yend = 49, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 48, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 39, xend = 39, y = 67, yend = 65, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 67, yend = 67, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 65, yend = 65, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 66, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  #Firmicutes
  annotate("segment", x = 39, xend = 39, y = 68, yend = 97, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 68, yend = 68, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 97, yend = 97, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 82.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 39, xend = 39, y = 53, yend = 64, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 53, yend = 53, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 64, yend = 64, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 59.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  #Desulfobacterota
  annotate("segment", x = 39, xend = 39, y = 45, yend = 46, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 45, yend = 45, color = "black", size = 0.6) +
  annotate("segment", x = 39, xend = 37.5, y = 46, yend = 46, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 45.5, label = "Desulfobacterota", size = 5.5, hjust = 0) + 
  
  #Spirochaetota
  annotate("segment", x = 39, xend = 37.5, y = 44, yend = 44, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 44, label = "Spirochaetota", size = 5.5, hjust = 0)  +
  
  #Campylobacterota
  annotate("segment", x = 39, xend = 37.5, y = 6, yend = 6, color = "black", size = 0.6) +
  annotate("text", x = 40, y = 6, label = "Campylobacterota", size = 5.5, hjust = 0) 
  
tree

ggsave("figures/tree.pdf", tree, width =9, height = 12.5)

```















color branch color by avg reconstructed model weight of parent and child nodes. I am averaging so that I don't have to choose a node color for the branch. Alternative would be to color branch by the parent node value

```{r}

edge_df <- tree_data %>%
  select(Taxa, isTip, node, x, y, parent, ace_weight, effect_direction, Phylum) %>%
  rename(child_node = node, child_x = x, child_y = y, child_weight = ace_weight) %>%
  left_join(
    tree_data %>%
      select(node, x, y, ace_weight) %>%
      rename(parent = node, parent_x = x, parent_y = y, parent_weight = ace_weight),
    by = "parent"
  ) %>%
  mutate(mid_weight = rowMeans(cbind(parent_weight, child_weight), na.rm = TRUE))

```



```{r}


plot <- ggplot(edge_df) +
  geom_segment(aes(x = parent_x, y = parent_y,
                   xend = child_x, yend = child_y,
                   color = mid_weight), size = 1.2) +
  scale_color_gradient(low = "grey60", high = "red", name = "Model weight") +
  
  geom_point(
    data = edge_df %>% filter(isTip == TRUE),
    aes(x = child_x, y = child_y, fill = effect_direction),
    size = 2.5, shape = 21, stroke = 0.3, color = "black") +
    
  geom_text(data = edge_df %>% filter(!is.na(Phylum)), 
            aes(x = child_x, y = child_y, label = Phylum),
            size = 2.5, hjust = -0.05) + 
  
  geom_text(data = edge_df %>% filter(isTip == TRUE),
    aes(x = child_x, y = child_y, label = Taxa),
    size = 2.5, hjust = -0.2, color = "black") +
  
  scale_fill_manual(
    values = c(
      "positive" = "forestgreen",
      "negative" = "firebrick",
      "neutral"  = "grey60"
    ),
    name = "Effect direction"
  ) +
  theme_tree2() +
  coord_cartesian(clip = "off")


plot
```






# uncertainty histogram

deepest nodes have widest confidence intervals

```{r}

# Width of the 95% confidence interval per node
ci_width <- ace_result$CI95[, 2] - ace_result$CI95[, 1]

# Create a dataframe
ci_df <- data.frame(
  node = (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree)),
  ci_lower = ace_result$CI95[, 1],
  ci_upper = ace_result$CI95[, 2],
  ci_width = ci_width
)

ggplot(ci_df, aes(x = ci_width)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "white", alpha = 0.9) +
  theme_minimal(base_size = 14) +
  labs(
    title = "95% Confidence Interval Width for Ancestral Nodes",
    x = "CI Width",
    y = "Number of Internal Nodes") 

```




















### ii. compile tree info

Here we compile the data for the tree and run the ancestral state reconstruction. First assign the tip labels as genus names instead of otu numbers. Then, add the metadata for temp sensitivity and change the tip labels to the temp sensitivity varible. Run ace() to do ancestral reconstruction and add node probability to data. Set node as "temp sensitive" is "yes" probability is greater than "no" probability. Branches are colored by node probabiliy and branch tips are colors by taxa sensitivity. 

```{r}

library(ggtree)
library(tidytree)
library(ape)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels

# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# now make tip labels the temp variable
genus_tree$tip.label <- tip_data$taxa_temp_sensitive
head(genus_tree$tip.label)

# run ace function
ace_result <- ace(tip_data$taxa_temp_sensitive, genus_tree, type = "discrete")

# extract node probs
node_probs <- as.data.frame(ace_result$lik.anc) 

# assign node probs to node
node_probs$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

# add the node probabilities
tree_data <- left_join(tree_data, node_probs, by = "node") %>%
  mutate(node_temp_sensitive = ifelse(yes > no, "yes", "no"),
         branch_colors = case_when(
           is.na(node_temp_sensitive) ~ taxa_temp_sensitive,
           TRUE ~ node_temp_sensitive))

# update tree plot with new data
tree_plot$data <- tree_data

```

### iii. plot tree

```{r}
tree_plot + 
  geom_tree(aes(color = branch_colors), size = 1) +
  geom_tiplab(size = 3) +
  scale_color_manual(values = c("no" = "#80b1d3", "yes" = "#fb8072")) +  
  theme_tree()

```









# data prep

```{r}

# compile data

tree_coef <- taxa_coef2 %>%
  select(Taxa, coef) %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         status = if_else(Taxa %in% temp_filter, "sensitive", "neutral"))

tree_aic <- taxa_aic2 %>%
  filter(model == "m_full") %>%
  select(Taxa, weight)

tree_data <- tree_aic %>% left_join(tree_coef, by = "Taxa")


# filter and extract tree

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
pruned_tree <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)


```


