---
title: "2025-02-17_shotgun_phyloseq_objects"
author: "Lucore"
date: "2025-02-17"
output: html_document
---

NOTES: This sheet compiled the physloseq objects for the shotgun data. There are four different objects: 1. species abundnace data generated by bracken/kraken2, 2. pathway abundances generated by humann3, 3. KEGG abundances generated by humann3, and 4. EGGNOG group abundnaces generated by humann3. All the phyloseq objects contain the UNSTRATIFIED DATA, so no taxa information in the functional phyloseq objects. 

2025-02-27 addition: have changed this to make the uniref 90 and the uniref 50 phyloseq objects to be able to compared mapped and unmapped reads between the two databases

# 1. load environment

```{r}

library(tidyverse)
library(phyloseq)


m <- read_csv("data/metadata/2025-02-17_shotgun_compiled_metadata.csv")

species <- read_table("/Users/Jordan/Desktop/research_repos/tables_shotgun/bracken_abundance_species_mpa.txt")

### uniref_90 ###

KO_uniref90 <-  read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_90/humann_genefamilies_KO.tsv")

nog_uniref90 <-  read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_90/humann_genefamilies_eggnog.tsv")

path_uniref90 <- read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_90/humann_pathabundance_unstratified.tsv")

### uniref_50 ###

KO_uniref50 <-  read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_50/humann_genefamilies_KO.tsv")

nog_uniref50 <-  read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_50/humann_genefamilies_eggnog.tsv")

path_uniref50 <- read_tsv("/Users/Jordan/Desktop/research_repos/tables_shotgun/uniref_50/humann_pathabundance_unstratified.tsv")

```

# 2. Make phyloseq objects

## a. make species phyloseq

```{r}

# clean up sample names 
species <- species %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(species = `#Classification`) %>%
  column_to_rownames(var = "species")

# create phylo table
phylo_table = otu_table(species , taxa_are_rows = T) 

# Metadata, make row names sample names to make phyloseq object
m <- m %>% column_to_rownames(var = "sample")
phylo_meta = sample_data(m)

# Check sample names are equal
sample_names(phylo_meta) == sample_names(phylo_table)
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 

source("functions/parse_taxonomy.R")

species_tax_table <- species %>%
  rownames_to_column(var = "taxa_id") %>%
  select(taxa_id) %>%
  mutate(taxa = taxa_id) %>%
  column_to_rownames("taxa_id") %>%
  mutate(parsed_taxonomy = map_df(taxa, parse_taxonomy)) %>%
  unnest_wider(parsed_taxonomy) %>%
  rename_with(~ sub("^parsed_taxonomy$", "", .x)) %>% # Remove prefix from column names
  column_to_rownames("taxa") # readd column names

phylo_tax = tax_table(as.matrix(species_tax_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
species_phylo  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
species_phylo 

```

## b. make pathway phyloseq

### i. uniref 90

```{r}

# clean up sample names 
path <- path_uniref90 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(pathway = `# Pathway`) %>%
  column_to_rownames(var = "pathway") 

path <- path[, 1:95] # removes the negative (sample)

# create phylo table
phylo_table = otu_table(path, taxa_are_rows = T) 

# Check sample names are equal
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
path_table <- path %>%
  rownames_to_column(var = "pathway") %>%
  mutate(pathway_id = pathway) %>%
  select(pathway, pathway_id) %>%
  column_to_rownames("pathway_id") 

phylo_tax = tax_table(as.matrix(path_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
pathway_phylo_uniref90  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
pathway_phylo_uniref90 

```

### ii. uniref 50

```{r}

# clean up sample names 
path <- path_uniref50 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(pathway = `# Pathway`) %>%
  column_to_rownames(var = "pathway") 

path <- path[, 1:95] # removes the negative (sample)

# create phylo table
phylo_table = otu_table(path, taxa_are_rows = T) 

# Check sample names are equal
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
path_table <- path %>%
  rownames_to_column(var = "pathway") %>%
  mutate(pathway_id = pathway) %>%
  select(pathway, pathway_id) %>%
  column_to_rownames("pathway_id") 

phylo_tax = tax_table(as.matrix(path_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
pathway_phylo_uniref50  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
pathway_phylo_uniref50 

```

## c. make KO phyloseq

### i. uniref 90

```{r}

# clean up sample names 
KO <- KO_uniref90 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(gene_ko = `# Gene Family`) 

KO <- KO[, 1:96] # removes the negative (sample)

# filter for KO groups 
ko_groups <- KO %>%
  filter(!str_detect(gene_ko, "\\|")) %>%
  column_to_rownames(var = "gene_ko")
  
# filter out gene KO groups for stratified table
ko_stratified <- KO %>%
  filter(str_detect(gene_ko, "\\|")) %>% 
  filter(str_detect(gene_ko, "\\|")) %>% 
  column_to_rownames(var = "gene_ko")

# create phylo table
phylo_table = otu_table(ko_groups, taxa_are_rows = T) 

# Check sample names are equal
sample_names(phylo_meta) == sample_names(phylo_table)
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
ko_groups_table <- ko_groups %>%
  rownames_to_column(var = "gene_ko") %>%
  mutate(ko_family = gene_ko) %>%
  select(gene_ko, ko_family) %>%
  column_to_rownames("ko_family") 

phylo_tax = tax_table(as.matrix(ko_groups_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
ko_phylo_uniref90  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
ko_phylo_uniref90 

```

### ii. uniref 50

```{r}

# clean up sample names 
KO <- KO_uniref50 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(gene_ko = `# Gene Family`) 

KO <- KO[, 1:96] # removes the negative (sample)

# filter for KO groups 
ko_groups <- KO %>%
  filter(!str_detect(gene_ko, "\\|")) %>%
  column_to_rownames(var = "gene_ko")
  
# filter out gene KO groups for stratified table
ko_stratified <- KO %>%
  filter(str_detect(gene_ko, "\\|")) %>% 
  filter(str_detect(gene_ko, "\\|")) %>% 
  column_to_rownames(var = "gene_ko")

# create phylo table
phylo_table = otu_table(ko_groups, taxa_are_rows = T) 

# Check sample names are equal
sample_names(phylo_meta) == sample_names(phylo_table)
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
ko_groups_table <- ko_groups %>%
  rownames_to_column(var = "gene_ko") %>%
  mutate(ko_family = gene_ko) %>%
  select(gene_ko, ko_family) %>%
  column_to_rownames("ko_family") 

phylo_tax = tax_table(as.matrix(ko_groups_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
ko_phylo_uniref50  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
ko_phylo_uniref50 

```


## d. make eggnog phyloseq

### i. uniref 90

```{r}

# clean up sample names 
nog <- nog_uniref90 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(gene_ko = `# Gene Family`) 

nog <- nog[, 1:96] # removes the negative (sample)

# filter for KO groups 
nog_groups <- nog %>%
  filter(!str_detect(gene_ko, "\\|")) %>%
  column_to_rownames(var = "gene_ko")
  
# filter out gene KO groups for stratified table
nog_stratified <- nog %>%
  filter(str_detect(gene_ko, "\\|")) %>% 
  filter(str_detect(gene_ko, "\\|")) %>% 
  column_to_rownames(var = "gene_ko")

# create phylo table
phylo_table = otu_table(nog_groups, taxa_are_rows = T) 

# Check sample names are equal
sample_names(phylo_meta) == sample_names(phylo_table)
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
nog_groups_table <- nog_groups %>%
  rownames_to_column(var = "gene_ko") %>%
  mutate(ko_family = gene_ko) %>%
  select(gene_ko, ko_family) %>%
  column_to_rownames("ko_family") 

phylo_tax = tax_table(as.matrix(nog_groups_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
eggnog_phylo_uniref90  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
eggnog_phylo_uniref90 




```

### ii. uniref 50

```{r}

# clean up sample names 
nog <- nog_uniref50 %>%
  rename_with(~ str_replace(., "(JL).*", "JL"), everything()) %>%
  rename(gene_ko = `# Gene Family`) 

nog <- nog[, 1:96] # removes the negative (sample)

# filter for KO groups 
nog_groups <- nog %>%
  filter(!str_detect(gene_ko, "\\|")) %>%
  column_to_rownames(var = "gene_ko")
  
# filter out gene KO groups for stratified table
nog_stratified <- nog %>%
  filter(str_detect(gene_ko, "\\|")) %>% 
  filter(str_detect(gene_ko, "\\|")) %>% 
  column_to_rownames(var = "gene_ko")

# create phylo table
phylo_table = otu_table(nog_groups, taxa_are_rows = T) 

# Check sample names are equal
sample_names(phylo_meta) == sample_names(phylo_table)
identical(sample_names(phylo_meta), sample_names(phylo_table))

# Crate taxonomy table from otu table 
nog_groups_table <- nog_groups %>%
  rownames_to_column(var = "gene_ko") %>%
  mutate(ko_family = gene_ko) %>%
  select(gene_ko, ko_family) %>%
  column_to_rownames("ko_family") 

phylo_tax = tax_table(as.matrix(nog_groups_table)) #taxonomy

# Check sample taxa names are equal
identical(taxa_names(phylo_tax), taxa_names(phylo_table)) 

# Make phyloseq object
eggnog_phylo_uniref50  <- phyloseq(phylo_table, phylo_meta, phylo_tax) 
eggnog_phylo_uniref50 




```

# 3. export phyloseq objects

```{r}

saveRDS(eggnog_phylo_uniref50, file = "phyloseq_objects/eggnog_phylo_u50.rds")
saveRDS(ko_phylo_uniref50, file = "phyloseq_objects/ko_phylo_u50.rds")
saveRDS(pathway_phylo_uniref50, file = "phyloseq_objects/pathway_phylo_u50.rds")

saveRDS(eggnog_phylo_uniref90, file = "phyloseq_objects/eggnog_phylo_u90.rds")
saveRDS(ko_phylo_uniref90, file = "phyloseq_objects/ko_phylo_u90.rds")
saveRDS(pathway_phylo_uniref90, file = "phyloseq_objects/pathway_phylo_u90.rds")

saveRDS(species_phylo, file = "phyloseq_objects/species_phylo.rds")

```

