---
title: "Untitled"
author: "Lucore"
date: "2025-05-15"
output: html_document
---

Notes: This is the compiled analysis for both the 16s and shotgun paper with the code for the final figures. 

It does not include the GAM analysis. 

# 1. Load environment

```{r}

# packages
library(tidyverse)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(ggnewscale)
library(zCompositions)
library(ggtree)
library(tidytree)
library(ape)
library(jsonlite)
library(file2meco)
library(phyloseq)
library(flextable)

#physeq objects
tax_physeq <- readRDS("phyloseq_objects/2024-12-18_phyloseq_16s.rds")
path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

# functions
source("functions/my_scale.R")
source("functions/inverse_scale.R")
source("functions/kegg_parse_json.R")

# conflicts
conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)


```


# 2. Data set up

## A. Taxonomy (16s seqs)

### i. Metadata filtering

using 277 samples that have feeding and neopterin metadata, neopterin value is neopterin averaged 3 days before and after GM sample

```{r}

tax_m <- data.frame(sample_data(tax_physeq)) %>% 
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals, neo)

tax_m_filtered <- tax_m[!is.na(tax_m$fruit), ] # remove samples with no diet data
tax_m_filtered <- tax_m_filtered[!is.na(tax_m_filtered$neo), ] # remove samples with no neo, 277 samples left

tax_physeq2 <- prune_samples(rownames(tax_m_filtered), tax_physeq) # filter physeq object

tax_physeq2 <- prune_taxa(taxa_sums(tax_physeq2) > 0, tax_physeq2) # remove ASV's with zero abundance

tax_physeq2

```

### ii. Prevalence filtering 

Identifies taxa that are not present in at least 0.5% of the total samples (at least two samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth.  

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(tax_physeq2),
                MARGIN = ifelse(taxa_are_rows(tax_physeq2), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(tax_physeq2),
                      tax_table(tax_physeq2))

# define prevalence threshold
prevalenceThreshold = 0.005 * nsamples(tax_physeq2)  
prevalenceThreshold # must appear in at least two samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

ggsave("figures/supp/figS_prev.pdf", prev_fig, width = 6, height = 4)

# summary stats
asv_onesample <- nrow(prevdf[prevdf$Prevalence == 1,]) # number of ASV's present only in one sample
total_asv <- nrow(prevdf)
percent_onesample <- (asv_onesample/total_asv) * 100 # percent ASV's in only one sample

asv_nophyla <- nrow(prevdf[is.na(prevdf$Phylum),]) # number of ASV's with no assigned phylum 
percent_nophyla <- (asv_nophyla/total_asv) * 100 # percent number of ASV's with no phylum

## execute prevalence filter ##
tax_phylo_filtered <- prune_taxa((prev0 > prevalenceThreshold), tax_physeq2) 
tax_phylo_filtered

## remove taxa with unnasigned phyla ## 
tax_phylo_filtered2 <- subset_taxa(tax_phylo_filtered, !is.na(Phylum)) # remove 102 taxa 
tax_phylo_filtered2

# calc reads per sample
reads_filtered <- as.data.frame(sample_sums(tax_phylo_filtered2)) %>% 
  rownames_to_column(var = "sample") %>%
  rename(filtered_reads = 'sample_sums(tax_phylo_filtered2)')
  
# add reads per sample to metadata
tax_m_filtered2 <- tax_m_filtered %>%
  left_join(reads_filtered, by = "sample") 

# % assignment for remaining taxonomy levels

asign_class <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Class"])))/1159)*100)

asign_order <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Order"])))/1159)*100)

asign_family <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Family"])))/1159)*100)

asign_genus <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Genus"])))/1159)*100)

######## evaluate which phlya have greater than 1% rel abund ############

phylum_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Phylum"))
  
phylum_otu_table <- as.data.frame(otu_table(phylum_aggregated))
phylum_tax_table <- as.data.frame(tax_table(phylum_aggregated))

phylum_otu_ids <- rownames(phylum_otu_table)
phylum_tax_ids <- rownames(phylum_tax_table)

phylum_names <- phylum_tax_table[phylum_otu_ids, "Phylum"]

rownames(phylum_otu_table) <- phylum_names

# across samples
df_total <- phylum_otu_table2 %>%
  mutate(Phylum_grouped = case_when(
    str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
    TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarise(TotalCount = sum(Count)) %>%
  ungroup() %>%
  mutate(RelativeAbundance = TotalCount / sum(TotalCount)) %>%
  mutate(Phylum_grouped = case_when(RelativeAbundance > 0.01 ~ Phylum_grouped, 
                                    TRUE ~ "Less than 1% rel abund")) %>%
  group_by(Phylum_grouped) %>%
  summarise(RA = sum(RelativeAbundance)) %>%
  ungroup() %>%
  mutate(RA = 100*RA,
         Phylum_grouped = fct_reorder(Phylum_grouped, RA))
                                    

ggplot(df_total, aes(x = Phylum_grouped, y = RA, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Relative Abundance (%)", fill = "Phylum") +
  theme_minimal() +
  coord_flip() 

```

###  iii. Generate alpha diversity metrics

```{r}

shannon <- estimate_richness(tax_phylo_filtered2, measures = "Shannon") # calc shannon
shannon <- shannon %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

hist(shannon$shannon) # normal dist

simpson <- estimate_richness(tax_phylo_filtered2, measures = "Simpson") # calc Simpson
simpson <- simpson %>%
  rownames_to_column(var = "sample") %>%
  rename(simpson = Simpson) 

hist(simpson$simpson) # normal dist

observed <- estimate_richness(tax_phylo_filtered2, measures = "Observed") # calc observed
observed <- observed %>%
  rownames_to_column(var = "sample") %>%
  rename(observed = Observed) %>%
  mutate( sample = str_replace_all(sample, "^X", ""),  
          sample = str_replace_all(sample, "\\.", "-"))

hist(observed$observed) # gamma dist


tax_m_filtered3 <- tax_m_filtered2 %>% # add to metadata
  left_join(shannon, by = "sample") %>%
  left_join(observed, by = "sample") %>%
  left_join(simpson, by = "sample")

```

### iv. CLR transformationand PCA

The rows need to be samples when doing the PCA because results show how samples cluster based on ASV's instead of the samples themselves

Use the CLR transformation and PCA analysis because of Gloor et al., 2017, Baniel publications, Gloor and Ried 2016, and this tutorial https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot reccomend that normalizing for sequencing depth is best done with a CLR transformation (as opposed to rarefaction) and that sample distances are best estimated using Aitchisons distances as opposed to Bray Curtis or Unifrac because they do not reflect the compositional nature of the data. Aitchison distance is the euclidean distances between the samples after the CLR transformation.

Retain PCA axes 1-3 because they explain >5% of the variance

```{r}

######### transformation and PCA #########

otu_table <- as.matrix(t(otu_table(tax_phylo_filtered)))
otu_table[1:5,1:5] # rows are samples 

set.seed(123)
czm <- cmultRepl(otu_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
czm[1:5,1:5] # rows are samples 

clr <- t(apply(czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
clr[1:5,1:5] # rows are samples 

pca <- prcomp(clr) # calc pca
plot(pca$x[,1],pca$x[,2]) # plot 

d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="") # PC1 14.4%
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 8.1%

######### make scree plot #########

eigenvalues <- pca$sdev^2  # square the standard deviations to get eigenvalues
percent_explained <- 100 * eigenvalues / sum(eigenvalues) 

# scree plot
pe_plot <- tibble(pe = percent_explained, axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  geom_point() +
  labs(x = "Principal component axis", y = "Percent variance explained") +
  scale_y_continuous(breaks = seq(0, max(percent_explained), by = 2.5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        axis.ticks.y = element_line(color = "black", linewidth = 0.3),
        axis.ticks.x = element_line(color = "black", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig1S.pdf", pe_plot, width = 6, height = 6)

######### Add PCA values to metadata #########

df<-data.frame(pca$x)
df$sample<-rownames(df)
df_small <- df[, c(1:3,278)]

# add to metadata
tax_m_filtered4 <- tax_m_filtered3 %>%
  left_join(df_small, by = "sample") 

```

### v. prep data for modeling

scale variables and make categorical variables factors. Make all palmas sampels tenori, only 6 palmas samples not enough to estimate effect and they have overlapping home ranges.

```{r}

# scale variables
tax_model <- tax_m_filtered4 %>%
  mutate(across(c(age, temp, rain, fruit, residuals, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)),
         ID = as.factor(ID))  

```

### vi. Suppelemenatary food plots

show that temp and fruit are correlated, to be able to evaluate temp, need to control for fruit.  

```{r}

# fruit ~ temp
fruit_plot <- ggplot(tax_model, aes(x = temp, y = fruit)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") +
  labs(y = "Count of fruit eaten", x = "Temperature (°C)") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig2S.pdf", fruit_plot, width = 6, height = 6)

```


## B. Pathways (shotgun seqs)

### i. Metadata filtering

using 71 samples that have feeding and neotperin metadata

```{r}

path_m <- data.frame(sample_data(path_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

path_m_filtered <- path_m[!is.na(path_m$fruit), ] # remove samples with no diet data, drops 8 samples

path_m_filtered <- path_m_filtered[!is.na(path_m_filtered$neo), ] # remove samples with no neopterin data, down to 71 samples


# filter phyloseq objects to remove samples with no diet data 
path_physeq2 <- prune_samples(path_m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

### ii. Prevalence filtering

Remove unmapped and unintegrated values. Unmapped values represent the total abundance of reads that failed to align to any known gene sequences and unintegrated values represent genes that did not belong to a known pathway.

```{r}

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

### iii. Generate alpha diversity metrics

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

Ignore the warning, `estimate_richness()` prefers count data but it gives you exactly the same values as if you calculated it with vegan. 

```{r}

shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

simpson <- estimate_richness(path_filtered, measures = "Simpson") %>%
  rownames_to_column(var = "sample") %>%
  rename(simpson = Simpson) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(shannon, by = "sample") %>%
  full_join(simpson, by = "sample")

hist(path_alpha$shannon) 
hist(path_alpha$observed)
hist(path_alpha$simpson)

path_m1 <- path_m_filtered %>% # add to metadata
  left_join(path_alpha, by = "sample")

```

### iv. CLR transformation and PCA 

```{r}

######### transformation and PCA #########

path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 


path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 21.5%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 13.1%

######### Add PCA values to metadata #########

path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:3,72))

path_m2 <- path_m1 %>%
  left_join(path_df_small, by = "sample")

```

### v. Prep data to model

scale variables and make categorical variables factors

```{r}

path_model <- path_m2 %>%
  mutate(across(c(age, temp, rain, fruit, neo, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```


# 3. Relatinship wth temp

## A. Community-level modeling

### i. Taxonomy

Compare m_main with base-level predictors group, age, sex, fruit, diet diversity, rain, and ID as predictors, to m_full which adds a temperature predictor. Both model include an offset of reads to control for seq depth. Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed. 

When you run the loop there is a singular fit error, this is from PC2, but the output looks okay and it is only for the m_main model not the m_full model. We are using the coefficient from m_full. 

```{r}

com_coef <- list() 
com_aic <- list()

com_features <- colnames(tax_model)[c(15, 17:20)] # list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      
      m_main <- lmer(formula_main, data = tax_model)
      m_full <- lmer(formula_full, data = tax_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
tax_com_aic <- do.call(rbind, com_aic)
tax_com_coef <- do.call(rbind, com_coef)


#### model observed ######

o_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

o_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = tax_model)

AICc <- as.data.frame(AICctab(o_full, o_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(o_full)$coeff[8,1], 4)
std <- round(summary(o_full)$coeff[8,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

tax_com_coef <- rbind(tax_com_coef, results_df)
tax_com_aic <- rbind(tax_com_aic, aic_df)

```

### ii. Pathways

normally distributed enough to model with Gaussian (PC1, shannon, and simpson are most concerning)

```{r}

####### histogranms ########

hist(path_model$shannon) 
hist(path_model$simpson)
hist(path_model$observed) 
hist(path_model$PC1)
hist(path_model$PC2)
hist(path_model$PC3)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(path_model)[c(15:20)] # list of features to model

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = path_model)
      m_full <- lmer(formula_full, data = path_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
path_com_aic <- do.call(rbind, com_aic)
path_com_coef <- do.call(rbind, com_coef)

```

### iii. Plot

#### a. functions

```{r}

make_coef_plot <- function(data) {
  ggplot(data, aes(x = order, y = coef, color = direction)) +
    geom_hline(yintercept = 0, linetype = "dashed", size = 0.5, color = "grey40") +
    geom_point(size = 3) +
    geom_linerange(aes(ymin = low_ci_95, ymax = high_ci_95), linewidth = 1) +
    geom_linerange(aes(ymin = low_ci_50, ymax = high_ci_50), linewidth = 1.5) +
    scale_color_manual(values = c("positive" = "#80b1d3", "negative" = "#fb8072"),
                       breaks = c("positive", "negative")) +
    coord_flip() +
    labs(y = "", x = "") +
    theme(axis.line = element_line(color = "black", linewidth = 0.28),
          axis.text = element_text(size = 12, color = "black"),
          text = element_text(size = 10, color = "black", family = "sans"),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          legend.position = "none",
          panel.background = element_blank()
    )
}

make_model_weight_plot <- function(data) {
  ggplot(data, aes(y = order)) +
    geom_bar(stat = "identity", aes(x = -m_full, fill = "m_full"), position = "identity", width = 0.4) +
    geom_bar(stat = "identity", aes(x = m_main, fill = "m_main"), position = "identity", width = 0.4) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    scale_fill_manual(values = c("m_full" = "#bc80bd", "m_main" = "#fdb462")) +
    xlim(-1, 1) +
    labs(x = "", y = "") +
    theme(
      axis.line = element_line(color = "black", linewidth = 0.28),
      axis.text.y = element_blank(),
      axis.text = element_text(size = 12, color = "black"),
      text = element_text(size = 10, color = "black", family = "sans"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      legend.position = "none",
      panel.background = element_blank()
    )
}

```

#### b. data prep

```{r}

## taxa data prep ##

plot_com_coef <- tax_com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_com_model <- tax_com_aic %>%
  select(metric, model, weight) %>%
  mutate(model = case_when(model == "o_full" ~ "m_full",
                           model == "o_main" ~ "m_main",
                           TRUE ~ model)) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)

## path data prep ##

plot_path_observed <- path_com_coef %>%
  filter(metric == "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_coef <- path_com_coef %>%
  filter(metric != "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_model_obs <- path_com_aic %>%
  filter(metric == "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%
   mutate(order = factor("observed"))

path_order <- c("PC2", "simpson", "shannon", "PC3", "PC1")

plot_path_model <- path_com_aic %>%
  filter(metric != "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%  
  mutate(order = factor(metric, levels = levels(plot_path_coef$order))) %>%
  arrange(order)

levels <- levels(plot_com_model$order)
which(levels == "PC3")


```

#### c. combined plot

add x lables, fuss with ticks

```{r}

p1 <- make_coef_plot(plot_com_coef) +
  ylim(-10, 10) +
  ggtitle("A. Taxonomy coefficients") 

p2 <- make_coef_plot(plot_path_coef) +
  ylim(-21, 21) +
  ggtitle("C. Pathway coefficients")

p3 <- make_coef_plot(plot_path_observed) + 
  ylim(-100, 100) 

p4 <- make_model_weight_plot(plot_com_model) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 6.45, label = "temp model", color = "#bc80bd", size = 5) +
  annotate("text", x = 0.5, y = 6.45, label = "base model", color = "#fdb462", size = 5)

p5 <- make_model_weight_plot(plot_path_model) +
  ggtitle("D. Pathway model weights") 

p6 <- make_model_weight_plot(plot_path_model_obs)


ggarrange(p1, p4, p2, p5, p3, p6, ncol = 2, nrow = 3, 
          align = "v",
          heights = c(1, 1, 0.3))

```


## B. Taxonomic differential abundance testing

### i. Preprocessing

clean up taxa names in phyloseq object, use `tax_glom()` to combine taxa at the genus level and at the family level (i.e. sum abundances of the ASV's that are grouped together into genus or family). Change row names from ASV ID to taxa name. 

```{r}

# get rid of all the weird, unnassigned names and turn them all into NA's
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("Unassigned*",NA,x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("uncultured*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("unidentified*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("gut metagenome",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("metagenome",NA, x)))

# Tax glom 
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))
family_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Family"))

# extract tables
genus_otu_table <- as.data.frame(otu_table(genus_aggregated))
genus_tax_table <- as.data.frame(tax_table(genus_aggregated))
genus_metadata <- sample_data(genus_aggregated)

family_otu_table <- as.data.frame(otu_table(family_aggregated))
family_tax_table <- as.data.frame(tax_table(family_aggregated))
family_metadata <- sample_data(family_aggregated)

#change OTU rownames to ASV family names
# Ensure that the OTU IDs are the same in both tables (rows in otu_df and tax_df)
genus_otu_ids <- rownames(genus_otu_table)
genus_tax_ids <- rownames(genus_tax_table)

family_otu_ids <- rownames(family_otu_table)
family_tax_ids <- rownames(family_tax_table)

# Extract the taxa names from the taxonomy table
genus_names <- genus_tax_table[genus_otu_ids, "Genus"]
family_names <- family_tax_table[family_otu_ids, "Family"]

# Set the row names of the OTU table to the family names
rownames(genus_otu_table) <- genus_names
rownames(family_otu_table) <- family_names

```

### ii. Calc rel abundance 

Calculate the relative abundance by dividing the raw abundance counts by the total reads per sample. Using the "filtered_reads" column because we are using the phyloseq object `tax_phylo_filtered2` that has already been filtered at a 0.5% threshold. 

```{r}

# extract table for DA testing
genus_da <-psmelt(genus_aggregated)
family_da <-psmelt(family_aggregated)

# Extract variables of interest 
genus_da1 <- genus_da %>% 
  mutate(Taxa = Genus) %>% # make genus the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>% 
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Genus") %>%
  mutate(RA = (Abundance/filtered_reads)*100)

family_da1 <- family_da %>% 
  mutate(Taxa = Family) %>% # make family the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>%
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Family") %>%
  mutate(RA = (Abundance/filtered_reads)*100) # calculates relative abundance

# n = 277
length(unique(genus_da1$sample))  
length(unique(family_da1$sample))

```

### iii. Filter out low abundance taxa

Calculate the average relative abundance for each taxa across all of the samples and then keep the taxa that have at lease 0.2% relative abundance across all samples. Left with 134 genera and 79 families. 

```{r}

#Calculate average relative abundance of each taxa across all samples
genus_a1<-sapply(unique(genus_da1$Taxa),function(x) mean(genus_da1$RA[genus_da1$Taxa==x],na.rm=TRUE))
genus_a2<-data.frame(cbind(Taxa=as.character(unique(genus_da1$Taxa)),RA=genus_a1))
genus_a2$RA<-as.numeric(as.character(genus_a2$RA))

family_a1<-sapply(unique(family_da1$Taxa),function(x) mean(family_da1$RA[family_da1$Taxa==x],na.rm=TRUE))
family_a2<-data.frame(cbind(Taxa=as.character(unique(family_da1$Taxa)),RA=family_a1))
family_a2$RA<-as.numeric(as.character(family_a2$RA))


#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

genus_da2 <-droplevels(subset(genus_da1, genus_da1$Taxa %!in% NA)) # remove asv NAs
genus_da3 <-droplevels(subset(genus_da2, genus_da2$Taxa %in% unique(genus_a2$Taxa[genus_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(genus_da2$Taxa))
length(unique(genus_da3$Taxa)) # 134 genera

family_da2 <-droplevels(subset(family_da1, family_da1$Taxa %!in% NA)) # remove asv NAs
family_da3 <-droplevels(subset(family_da2, family_da2$Taxa %in% unique(family_a2$Taxa[family_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(family_da2$Taxa))
length(unique(family_da3$Taxa)) # 79 families

# Combine family and genus data frames 
da4 <- rbind(genus_da3, family_da3) %>%
  filter(!(Taxa == "CAG-274" & level == "Family")) # Remove CAG-274 as family, listed as both family and genus

```

### iv. Scale variables

scale variables for modeling, combine tenori and palmas, 211 total taxa for data 

```{r}

# Scale variables
da5 <- da4 %>%
   mutate(across(c(temp, rain, age, fruit, neo), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID),
         sex = as.factor(sex)) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)))

length(unique(da5$Taxa)) # 211 taxa total

```

### v. AIC comparison

this loops takes a while to run. 

Compare m_main with all predictors except temp and m_full which includes temp for all taxa. Not including diet diversity in these models to help simplify the models. 

```{r}

# create lists
taxa_coef <- list()
taxa_aic <- list()
model_results <- list()

# loop
for (i in unique(da5$Taxa)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(da5, Taxa == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- gam(Abundance ~ age_s + temp_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 

      m_main <- gam(Abundance ~ age_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 
      
    }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
     # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      # extract temp coef
      coef <- round(summary(m_full)$p.coeff[3], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[3], 4)
    
      # clac CI'
      ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
      ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
      # bind data
      df <- cbind(ci_95, ci_50)
      
      #create data frames
    aic_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
        Taxa = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3),
        stringsAsFactors = FALSE)

    options(scipen = 999)  

    model_results_df <- data.frame(
      Taxa = i,
      Predictor = names(summary(m_full)$p.coeff[c(2:7)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:7)], 4),
      StdError = round(summary(m_full)$se[c(2:7)], 4),
      stringsAsFactors = FALSE)
    
    
      # save results
    taxa_aic[[i]] <- aic_df
    taxa_coef[[i]] <- results_df
    model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
taxa_aic <- do.call(rbind, taxa_aic)
taxa_coef <- do.call(rbind, taxa_coef)
model_results <- do.call(rbind, model_results)



```

### vi. taxa selection

First, we remove all taxa with explicit convergence warnings, then we remove all taxa that look like they didn't converge based on SE in the model results. 48 taxa did not converge. After removing the taxa that don't converge we tested 163 taxa in total and 93 of those have m_full with a model weight  > 0.5

Then, we select all taxa where m_full (the model with the temp predictor) fits the best. 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(taxa_aic$Taxa[taxa_aic$Warning == "Convergence Warning"]) # 22 do not converge

# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 80) %>%
  pull(Taxa) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 48 do not converge

# remove error taxa
da6 <- da5 %>% filter(!Taxa %in% errors)
length(unique(da6$Taxa)) # 163 remaining taxa

taxa_coef2 <- taxa_coef %>% filter(!Taxa %in% errors)
length(taxa_coef2$Taxa)

taxa_aic2 <- taxa_aic %>% filter(!Taxa %in% errors)
length(taxa_aic2$Taxa) # 324/2 = 162

# filter for taxa where full model fits best
temp_filter <- taxa_aic2 %>%
  group_by(Taxa) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa) # 93 taxa
length(unique(temp_filter))

filtered_coef <- taxa_coef2 %>%filter(Taxa %in% temp_filter)
length(filtered_coef$Taxa)

filtered_aic <- taxa_aic2 %>%filter(Taxa %in% temp_filter)
length(filtered_aic$Taxa)

```

### vii summary stats

create final table of model results for all taxa that were tested, include coefficients and model weights, use for supplemental table
calculate some descriptive statistics for the results

```{r}

# final table of model results for all tested taxa
levels <- da6 %>% select(Taxa, level, Phylum) %>%
  distinct()

weights <- taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight)

summary <- taxa_coef2 %>%
  left_join(levels, by = "Taxa") %>%
  left_join(weights, by = "Taxa") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         sensitive = ifelse(weight > 0.5, "temp sensitive", "nuetral")) %>%
  rename(model_weight = weight,
         temp_coef = coef,
         std_error = std)

supp_table1 <- summary %>%
  select(Taxa, temp_coef, std_error, level, model_weight, sensitive) %>%
  flextable() %>%
    autofit()
print(supp_table1, preview = "docx")

# descriptive stats
table(summary$sensitive) # 93 temp sensitive
table(summary$level) # 97 genera, 66 families 

temp_sensitive <- summary %>% filter(sensitive == "temp sensitive")
table(temp_sensitive$level) # 37 families, 56 genera
length(unique(temp_sensitive$Phylum)) # 10 phyla
table(temp_sensitive$direction)# 82 negative, 11 positive

phyla_results <- summary %>% filter(level == "Genus")
table(phyla_results$Phylum, phyla_results$sensitive)

```


### viii. Ancestral state rescontruction

Create cladogram that shows the results of the temp DAA for genera with an ancestral state reconstruction using the model weights for the temp model

#### a. data prep 

extract the necessary metadata and filter the phyloseq object for all genera included in the DA temp analysis. Then extract the tree. 

```{r}

# get list of taxa with ne
direction <- filtered_coef %>% 
  mutate(effect_direction = ifelse(coef > 0, "positive", "negative")) %>%
  select(Taxa, effect_direction)

# add temp information to da table
da7 <- da6 %>%
  mutate(taxa_temp_sensitive = ifelse(Taxa %in% temp_filter, "sensitive", "nuetral")) %>%
  left_join(direction, by = "Taxa") %>%
  mutate(effect_direction = case_when(
    is.na(effect_direction) ~ "neutral",
    TRUE ~ effect_direction))
  
# get genus names
genus_names <- da7 %>%
  filter(level == "Genus") %>%
  pull(Taxa) %>%
  unique()

# aggregate by taxa
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
temp_genus_phylo <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)
genus_tax <- as.data.frame(tax_table(temp_genus_phylo))

# prep final tree metadata 
genus_meta <- da7 %>%
  filter(level == "Genus") %>%
  select(Taxa, taxa_temp_sensitive, effect_direction) %>%
  distinct() %>%
  left_join(taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight), by = "Taxa")

# create phylum groups for clade labels
phylum_groups <- genus_tax %>%
  rownames_to_column("OTU") %>%
  mutate(Genus = as.character(Genus), Phylum = as.character(Phylum)) %>%
  filter(!is.na(Phylum), !is.na(Genus)) %>%
  group_by(Phylum) %>%
  summarise(genus_list = list(Genus)) %>%
  unnest_longer(genus_list, values_to = "Genus")

```

#### b. assign tip labels

First assign the tip labels as genus names instead of otu numbers. Add the metadata for temp sensitivity and change the tip labels to the model weight varible

```{r}

# tip labels are still otu numbers
head(genus_tree$tip.label)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels
  
# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# now make tip labels the model weight
genus_tree$tip.label <- tip_data$weight
head(genus_tree$tip.label)

```

####  c. ancestral state reconstruction

reconstruction returns the maximum likelihood estimates of the model weights for each node, assumes brownian motion. Higher weights are more temperature sensitive than lower weights. 

```{r}

# run ace function
ace_result <- ace(tip_data$weight, genus_tree, type = "continuous")

# extract node probs
node_df <- data.frame(
  node = as.numeric(names(ace_result$ace)),
  ace_weight = ace_result$ace) # ignore warning will fix in next step

# assign node probs to node
node_df$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

tree_data <- left_join(tree_data, node_df, by = "node") %>%
  mutate(ace_weight = ifelse(isTip, weight, ace_weight)) # assign weights to tips for plotting branch color

# summary stats - add this to tree to get node labels to inspect weights of ancestral nodes 
#geom_text2(
  #aes(subset = !isTip, label = node),
  #size = 3,
 # hjust = -0.3,
  #color = "black"
#) 

```

#### d. plot tree

labeled phyla have > 1% rel abundance 

```{r}

tree <- ggtree(genus_tree, branch.length = "none") %<+% tree_data +
  geom_tree(aes(color = ace_weight), size = 1) +
  
  scale_x_continuous(
  expand = expansion(mult = c(0, 0)),
  limits = c(NA, 45)  # or whatever max x value you need
) +
  
  scale_color_gradient(high = "#7570b3", low = "#d95f02", name = "Model weight") +
  
  ggnewscale::new_scale_color() + # add a new color scale for the tip circles
  
  geom_tiplab(aes(label = Taxa), size = 3.5, offset = 0.5) +
  
  geom_point2(aes(subset = isTip, color = effect_direction), shape = 19, size = 4, stroke = 0, show.legend = FALSE) +

  scale_color_manual(values = c("positive" = "#1f78b4", "negative" = "#e41a1c", "neutral"  = "#666666"), name = "Effect direction") +
  
  geom_text2(aes(subset = isTip, label = case_when(effect_direction == "positive" ~ "+", effect_direction == "negative" ~ "–")),
  size = 3.6, vjust = 0.4, hjust = 0.5, color = "white") +
  
  theme_tree2() +
  
  theme(legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()) +
  
  # Direction effect legend
  annotate("text", x = 1.2, y = 82, label = "Effect direction", size = 3.7, hjust = 0) +
  
  annotate("point", x = 1.5, y = 80.2, shape = 19, size = 4, color = "#1f78b4") +
  annotate("text", x = 2.1, y = 80.25, label = "positive", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 80.3, label = "+", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 78.4, shape = 19, size = 4, color = "#e41a1c") +
  annotate("text", x = 2.1, y = 78.45, label = "negative", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 78.5, label = "–", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 76.6, shape = 19, size = 4, color = "#666666") +
  annotate("text", x = 2.1, y = 76.65, label = "no effect", size = 3.3, hjust = 0) +
  
  # Bacteroidota
  annotate("segment", x = 32.5, xend = 32.5, y = 31, yend = 43, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 31, yend = 31, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 43, yend = 43, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 37, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 31, y = 1, yend = 1, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 1, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  # Proteobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 7, yend = 30, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 7, yend = 7, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 30, yend = 30, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 18, label = "Proteobacteria", size = 5.5, hjust = 0) +
  
  #Actinobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 47, yend = 49, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 47, yend = 47, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 49, yend = 49, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 48, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 67, yend = 65, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 67, yend = 67, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 65, yend = 65, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 66, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  #Firmicutes
  annotate("segment", x = 32.5, xend = 32.5, y = 68, yend = 97, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 68, yend = 68, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 97, yend = 97, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 82.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 53, yend = 64, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 53, yend = 53, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 64, yend = 64, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 59.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  #Campylobacterota
  annotate("segment", x = 32.5, xend = 31, y = 6, yend = 6, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 6, label = "Campylobacterota", size = 5.5, hjust = 0) 
  
tree

ggsave("figures/tree.pdf", tree, width =8, height = 12.5)


```


## C. Pathway differential abundance testing


