---
title: "Untitled"
author: "Lucore"
date: "2025-05-15"
output: html_document
---

Notes: This is the compiled analysis for both the 16s and shotgun with the code for the final figures. 

# 1. Load environment

```{r}

# packages
library(tidyverse)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(ggnewscale)
library(zCompositions)
library(ggtree)
library(tidytree)
library(ape)
library(jsonlite)
library(file2meco)
library(phyloseq)
library(flextable)

#physeq objects
tax_physeq <- readRDS("phyloseq_objects/2024-12-18_phyloseq_16s.rds")
path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

# functions
source("functions/my_scale.R")
source("functions/inverse_scale.R")
source("functions/kegg_parse_json.R")
source("functions/make_coef_plot.R")
source("functions/model_weight_plot.R")
source("functions/model_weight_plot.R")
source("functions/generate_shades.R")

# conflicts
conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)
conflicts_prefer(cowplot::get_legend)


```


# 2. Data set up

## A. Taxonomy (16s seqs)

### i. Metadata filtering

using 277 samples that have feeding and neopterin metadata, neopterin value is neopterin averaged 3 days before and after GM sample

```{r}

tax_m <- data.frame(sample_data(tax_physeq)) %>% 
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals, neo)

tax_m_filtered <- tax_m[!is.na(tax_m$fruit), ] # remove samples with no diet data
tax_m_filtered <- tax_m_filtered[!is.na(tax_m_filtered$neo), ] # remove samples with no neo, 277 samples left

tax_physeq2 <- prune_samples(rownames(tax_m_filtered), tax_physeq) # filter physeq object

tax_physeq2 <- prune_taxa(taxa_sums(tax_physeq2) > 0, tax_physeq2) # remove ASV's with zero abundance

tax_physeq2

```

### ii. Prevalence filtering 

Identifies taxa that are not present in at least 0.5% of the total samples (at least two samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth.  

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(tax_physeq2),
                MARGIN = ifelse(taxa_are_rows(tax_physeq2), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(tax_physeq2),
                      tax_table(tax_physeq2))

# define prevalence threshold
prevalenceThreshold = 0.005 * nsamples(tax_physeq2)  
prevalenceThreshold # must appear in at least two samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

ggsave("figures/supp/figS_prev.pdf", prev_fig, width = 6, height = 4)

# summary stats
asv_onesample <- nrow(prevdf[prevdf$Prevalence == 1,]) # number of ASV's present only in one sample
total_asv <- nrow(prevdf)
percent_onesample <- (asv_onesample/total_asv) * 100 # percent ASV's in only one sample

asv_nophyla <- nrow(prevdf[is.na(prevdf$Phylum),]) # number of ASV's with no assigned phylum 
percent_nophyla <- (asv_nophyla/total_asv) * 100 # percent number of ASV's with no phylum

## execute prevalence filter ##
tax_phylo_filtered <- prune_taxa((prev0 > prevalenceThreshold), tax_physeq2) 
tax_phylo_filtered

## remove taxa with unnasigned phyla ## 
tax_phylo_filtered2 <- subset_taxa(tax_phylo_filtered, !is.na(Phylum)) # remove 102 taxa 
tax_phylo_filtered2

# calc reads per sample
reads_filtered <- as.data.frame(sample_sums(tax_phylo_filtered2)) %>% 
  rownames_to_column(var = "sample") %>%
  rename(filtered_reads = 'sample_sums(tax_phylo_filtered2)')
  
# add reads per sample to metadata
tax_m_filtered2 <- tax_m_filtered %>%
  left_join(reads_filtered, by = "sample") 

# % assignment for remaining taxonomy levels

asign_class <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Class"])))/1159)*100)

asign_order <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Order"])))/1159)*100)

asign_family <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Family"])))/1159)*100)

asign_genus <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Genus"])))/1159)*100)

######## evaluate which phlya have greater than 1% rel abund ############

phylum_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Phylum"))
  
phylum_otu_table <- as.data.frame(otu_table(phylum_aggregated))
phylum_tax_table <- as.data.frame(tax_table(phylum_aggregated))

phylum_otu_ids <- rownames(phylum_otu_table)
phylum_tax_ids <- rownames(phylum_tax_table)

phylum_names <- phylum_tax_table[phylum_otu_ids, "Phylum"]

rownames(phylum_otu_table) <- phylum_names

# across samples
phylum_otu_table2 <- phylum_otu_table %>%
  rownames_to_column("Phylum") %>%
  pivot_longer(-Phylum, names_to = "sample", values_to = "Count") %>%
  left_join(reads_filtered, by = "sample") %>%
  group_by(sample) %>%
  mutate(RelativeAbundance = Count / sum(Count),
          Phylum_grouped = case_when(
           str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
           RelativeAbundance > 0.01 ~ Phylum,  
           TRUE ~ "Less than 1% rel abund"  
         )) %>%
  ungroup()

df_total <- phylum_otu_table2 %>%
  mutate(Phylum_grouped = case_when(
    str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
    TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarise(TotalCount = sum(Count)) %>%
  ungroup() %>%
  mutate(RelativeAbundance = TotalCount / sum(TotalCount)) %>%
  mutate(Phylum_grouped = case_when(RelativeAbundance > 0.01 ~ Phylum_grouped, 
                                    TRUE ~ "Less than 1% rel abund")) %>%
  group_by(Phylum_grouped) %>%
  summarise(RA = sum(RelativeAbundance)) %>%
  ungroup() %>%
  mutate(RA = 100*RA,
         Phylum_grouped = fct_reorder(Phylum_grouped, RA))

ggplot(df_total, aes(x = Phylum_grouped, y = RA, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Relative Abundance (%)", fill = "Phylum") +
  theme_minimal() +
  coord_flip() 

```

###  iii. Generate alpha diversity metrics

```{r}

shannon <- estimate_richness(tax_phylo_filtered2, measures = "Shannon") # calc shannon
shannon <- shannon %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

hist(shannon$shannon) # normal dist

observed <- estimate_richness(tax_phylo_filtered2, measures = "Observed") # calc observed
observed <- observed %>%
  rownames_to_column(var = "sample") %>%
  rename(observed = Observed) %>%
  mutate( sample = str_replace_all(sample, "^X", ""),  
          sample = str_replace_all(sample, "\\.", "-"))

hist(observed$observed) # gamma dist


tax_m_filtered3 <- tax_m_filtered2 %>% # add to metadata
  left_join(shannon, by = "sample") %>%
  left_join(observed, by = "sample") 

```

### iv. CLR transformationand PCA

The rows need to be samples when doing the PCA because results show how samples cluster based on ASV's instead of the samples themselves

Use the CLR transformation and PCA analysis because of Gloor et al., 2017, Baniel publications, Gloor and Ried 2016, and this tutorial https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot reccomend that normalizing for sequencing depth is best done with a CLR transformation (as opposed to rarefaction) and that sample distances are best estimated using Aitchisons distances as opposed to Bray Curtis or Unifrac because they do not reflect the compositional nature of the data. Aitchison distance is the euclidean distances between the samples after the CLR transformation.

Retain PCA axes 1-3 because they explain >5% of the variance

```{r}

######### transformation and PCA #########

otu_table <- as.matrix(t(otu_table(tax_phylo_filtered)))
otu_table[1:5,1:5] # rows are samples 

set.seed(123)
czm <- cmultRepl(otu_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
czm[1:5,1:5] # rows are samples 

clr <- t(apply(czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
clr[1:5,1:5] # rows are samples 

pca <- prcomp(clr) # calc pca
plot(pca$x[,1],pca$x[,2]) # plot 

d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="") # PC1 14.4%
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 8.1%

######### make scree plot #########

eigenvalues <- pca$sdev^2  # square the standard deviations to get eigenvalues
percent_explained <- 100 * eigenvalues / sum(eigenvalues) 

# scree plot
pe_plot <- tibble(pe = percent_explained, axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  geom_point() +
  labs(x = "Principal component axis", y = "Percent variance explained") +
  scale_y_continuous(breaks = seq(0, max(percent_explained), by = 2.5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        axis.ticks.y = element_line(color = "black", linewidth = 0.3),
        axis.ticks.x = element_line(color = "black", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig1S.pdf", pe_plot, width = 6, height = 6)

######### Add PCA values to metadata #########

df<-data.frame(pca$x)
df$sample<-rownames(df)
df_small <- df[, c(1:3,278)]

# add to metadata
tax_m_filtered4 <- tax_m_filtered3 %>%
  left_join(df_small, by = "sample") 

```

### v. prep data for modeling

scale variables and make categorical variables factors. Make all palmas sampels tenori, only 6 palmas samples not enough to estimate effect and they have overlapping home ranges.

```{r}

# scale variables
tax_model <- tax_m_filtered4 %>%
  mutate(across(c(age, temp, rain, fruit, residuals, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)),
         ID = as.factor(ID))  

```

### vi. Suppelemenatary food plots

show that temp and fruit are correlated, to be able to evaluate temp, need to control for fruit.  

```{r}

# fruit ~ temp
fruit_plot <- ggplot(tax_model, aes(x = temp, y = fruit)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") +
  labs(y = "Count of fruit eaten", x = "Temperature (°C)") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/fig2S.pdf", fruit_plot, width = 6, height = 6)

```


## B. Pathways (shotgun seqs)

### i. Metadata filtering

using 71 samples that have feeding and neotperin metadata

```{r}

path_m <- data.frame(sample_data(path_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

path_m_filtered <- path_m[!is.na(path_m$fruit), ] # remove samples with no diet data, drops 8 samples

path_m_filtered <- path_m_filtered[!is.na(path_m_filtered$neo), ] # remove samples with no neopterin data, down to 71 samples


# filter phyloseq objects to remove samples with no diet data 
path_physeq2 <- prune_samples(path_m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

### ii. Prevalence filtering

Remove unmapped and unintegrated values. Unmapped values represent the total abundance of reads that failed to align to any known gene sequences and unintegrated values represent genes that did not belong to a known pathway.

```{r}

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

### iii. Generate alpha diversity metrics

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

Ignore the warning, `estimate_richness()` prefers count data but it gives you exactly the same values as if you calculated it with vegan. 

```{r}

shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(shannon, by = "sample")

hist(path_alpha$shannon) 
hist(path_alpha$observed)

path_m1 <- path_m_filtered %>% # add to metadata
  left_join(path_alpha, by = "sample")

```

### iv. CLR transformation and PCA 

```{r}

######### transformation and PCA #########

path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 


path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 21.5%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 13.1%

######### Add PCA values to metadata #########

path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:3,72))

path_m2 <- path_m1 %>%
  left_join(path_df_small, by = "sample")

```

### v. Prep data to model

scale variables and make categorical variables factors

```{r}

path_model <- path_m2 %>%
  mutate(across(c(age, temp, rain, fruit, neo, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```


# 3. Temperature analysis

## A. Community-level modeling

### i. Taxonomy

Compare m_main with base-level predictors group, age, sex, fruit, diet diversity, rain, and ID as predictors, to m_full which adds a temperature predictor. Both model include an offset of reads to control for seq depth. Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed. 

When you run the loop there is a singular fit error, this is from PC2, but the output looks okay and it is only for the m_main model not the m_full model. We are using the coefficient from m_full. 

```{r}

com_coef <- list() 
com_aic <- list()

com_features <- colnames(tax_model)[c(15, 17:19)] # list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      
      m_main <- lmer(formula_main, data = tax_model)
      m_full <- lmer(formula_full, data = tax_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
tax_com_aic <- do.call(rbind, com_aic)
tax_com_coef <- do.call(rbind, com_coef)


#### model observed ######

o_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

o_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = tax_model)

AICc <- as.data.frame(AICctab(o_full, o_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(o_full)$coeff[8,1], 4)
std <- round(summary(o_full)$coeff[8,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

tax_com_coef <- rbind(tax_com_coef, results_df)
tax_com_aic <- rbind(tax_com_aic, aic_df)

```

### ii. Pathways

normally distributed enough to model with Gaussian (PC1, shannon, and simpson are most concerning)

```{r}

####### histogranms ########

hist(path_model$shannon) 
hist(path_model$observed) 
hist(path_model$PC1)
hist(path_model$PC2)
hist(path_model$PC3)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(path_model)[c(15:19)] # list of features to model

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = path_model)
      m_full <- lmer(formula_full, data = path_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
path_com_aic <- do.call(rbind, com_aic)
path_com_coef <- do.call(rbind, com_coef)

```

### iii. Plot model results

#### a. data prep

```{r}

## taxa data prep ##

plot_com_coef <- tax_com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_com_model <- tax_com_aic %>%
  select(metric, model, weight) %>%
  mutate(model = case_when(model == "o_full" ~ "m_full",
                           model == "o_main" ~ "m_main",
                           TRUE ~ model)) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)

## path data prep ##

plot_path_observed <- path_com_coef %>%
  filter(metric == "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_coef <- path_com_coef %>%
  filter(metric != "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_model_obs <- path_com_aic %>%
  filter(metric == "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%
   mutate(order = factor("observed"))

path_order <- c("PC2", "shannon", "PC3", "PC1")

plot_path_model <- path_com_aic %>%
  filter(metric != "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%  
  mutate(order = factor(metric, levels = levels(plot_path_coef$order))) %>%
  arrange(order)

levels <- levels(plot_com_model$order)
which(levels == "PC3") # height for plot label


```

#### b. combined plot

```{r}

p1 <- make_coef_plot(plot_com_coef) +
  ylim(-10, 10) +
  ggtitle("A. Taxonomy temperature coefficients") +
  ylab("")

p2 <- make_coef_plot(plot_path_coef) +
  ylim(-24, 24) +
  ggtitle("C. Metabolic pathway temperature coefficients") +
  ylab("")

p3 <- make_coef_plot(plot_path_observed) + 
  ylim(-100, 100) +
  ylab("Effect size")

p4 <- model_weight_plot(plot_com_model) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 5.45, label = "temp model", color = "#fee08b", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 5.45, label = "base model", color = "#66bd63", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p5 <- model_weight_plot(plot_path_model) +
  ggtitle("D. Metabolic pathway model weights") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p6 <- model_weight_plot(plot_path_model_obs) +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
  labels = c("100", "50", "0", "50", "100"),
  limits = c(-1, 1)) +
  xlab("model weight")

com_temp_plot <- ggarrange(p1, p4, p2, p5, p3, p6, ncol = 2, nrow = 3, 
          widths = c(1, 1),
          align = "v",
          heights = c(1, 1, 0.4))

ggsave("figures/fig1.pdf", com_temp_plot, width = 12, height = 8)

```

### iv. Plot alpha diversity results

plot model predictions for alpha diversity over raw data

```{r}

# set up predictions df
temp_seq <- seq(from = min(tax_model$temp_s, na.rm = TRUE), 
                to = max(tax_model$temp_s, na.rm = TRUE), length.out = 100)

pred_data <- data.frame(
  group = unique(tax_model$group)[1], 
  age_s = mean(tax_model$age_s, na.rm = TRUE),
  sex = unique(tax_model$sex)[1], 
  fruit_s = mean(tax_model$fruit_s, na.rm = TRUE),
  residuals_s = mean(tax_model$residuals_s, na.rm = TRUE),
  rain_s = mean(tax_model$rain_s, na.rm = TRUE),
  temp_s = temp_seq,
  filtered_reads = mean(tax_model$filtered_reads, na.rm = TRUE))

########## shannon ##########
s_temp <- lm(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + offset(log(filtered_reads)), 
               data = tax_model)
s_preds <- predict(s_temp,  newdata = pred_data, interval = "confidence", level = 0.95 ) %>%
  data.frame() %>%
  rename(s_preds = fit,
         s_lower = lwr,
         s_upper = upr)
pred_data <- cbind(pred_data, s_preds) 

# to rescales x axis
mean_temp <- mean(tax_model$temp)
sd_temp <- sd(tax_model$temp) 

s_plot <- ggplot(pred_data, aes(x = temp_s, y = s_preds)) +
  geom_point(data = tax_model, aes(x = temp_s, y = shannon),alpha = 0.5, size = 4, color = "#fee08b") +
  geom_smooth(method = "lm", lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = s_lower, ymax = s_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  ggtitle("A. Shannon's diversity") +
  labs(y = "", x = "") +
  scale_x_continuous(
    breaks = seq(min(tax_model$temp_s), max(tax_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        text = element_text(size = 10, color = "black", family = "sans"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

########## observed ##########
o_temp <- glm(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s +
                offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

o_preds <- data.frame(predict(o_temp,  newdata = pred_data, type = "response", se.fit = TRUE)) %>%
  select(-residual.scale) %>%
  rename(o_preds = fit,
         o_se = se.fit)

pred_data <- cbind(pred_data, o_preds) %>%
  mutate(o_lower = o_preds - 0.97 * o_se, # calc CI's 
         o_upper = o_preds + 0.97 * o_se)

o_plot <- ggplot(pred_data, aes(x = temp_s, y = o_preds)) +
  geom_point(data = tax_model, aes(x = temp_s, y = observed),alpha = 0.5, size = 4, color = "#fee08b") +
  geom_line(lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = o_lower, ymax = o_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  labs(y = "", x = "") +
  ggtitle("B. Observed ASV's") +
  scale_x_continuous(
    breaks = seq(min(tax_model$temp_s), max(tax_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        text = element_text(size = 10, color = "black", family = "sans"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

# add plots together  

alpha_plots <- ggarrange(s_plot, o_plot, ncol = 2, nrow = 1)

alpha_plots <- annotate_figure(alpha_plots, bottom = text_grob("temperature (°C)", size = 14))

ggsave("figures/fig2.pdf", alpha_plots, width =8, height = 4)

```

## B. Taxonomic differential abundance testing

### i. Preprocessing

clean up taxa names in phyloseq object, use `tax_glom()` to combine taxa at the genus level and at the family level (i.e. sum abundances of the ASV's that are grouped together into genus or family). Change row names from ASV ID to taxa name. 

```{r}

# get rid of all the weird, unnassigned names and turn them all into NA's
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("Unassigned*",NA,x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("uncultured*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("unidentified*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("gut metagenome",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("metagenome",NA, x)))

# Tax glom 
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))
family_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Family"))

# extract tables
genus_otu_table <- as.data.frame(otu_table(genus_aggregated))
genus_tax_table <- as.data.frame(tax_table(genus_aggregated))
genus_metadata <- sample_data(genus_aggregated)

family_otu_table <- as.data.frame(otu_table(family_aggregated))
family_tax_table <- as.data.frame(tax_table(family_aggregated))
family_metadata <- sample_data(family_aggregated)

#change OTU rownames to ASV family names
# Ensure that the OTU IDs are the same in both tables (rows in otu_df and tax_df)
genus_otu_ids <- rownames(genus_otu_table)
genus_tax_ids <- rownames(genus_tax_table)

family_otu_ids <- rownames(family_otu_table)
family_tax_ids <- rownames(family_tax_table)

# Extract the taxa names from the taxonomy table
genus_names <- genus_tax_table[genus_otu_ids, "Genus"]
family_names <- family_tax_table[family_otu_ids, "Family"]

# Set the row names of the OTU table to the family names
rownames(genus_otu_table) <- genus_names
rownames(family_otu_table) <- family_names

```

### ii. Calc rel abundance 

Calculate the relative abundance by dividing the raw abundance counts by the total reads per sample. Using the "filtered_reads" column because we are using the phyloseq object `tax_phylo_filtered2` that has already been filtered at a 0.5% threshold. 

```{r}

# extract table for DA testing
genus_da <-psmelt(genus_aggregated)
family_da <-psmelt(family_aggregated)

# Extract variables of interest 
genus_da1 <- genus_da %>% 
  mutate(Taxa = Genus) %>% # make genus the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>% 
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Genus") %>%
  mutate(RA = (Abundance/filtered_reads)*100)

family_da1 <- family_da %>% 
  mutate(Taxa = Family) %>% # make family the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, total_3_day, Phylum) %>%
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Family") %>%
  mutate(RA = (Abundance/filtered_reads)*100) # calculates relative abundance

# n = 277
length(unique(genus_da1$sample))  
length(unique(family_da1$sample))

```

### iii. Filter out low abundance taxa

Calculate the average relative abundance for each taxa across all of the samples and then keep the taxa that have at lease 0.2% relative abundance across all samples. Left with 134 genera and 79 families. 

```{r}

#Calculate average relative abundance of each taxa across all samples
genus_a1<-sapply(unique(genus_da1$Taxa),function(x) mean(genus_da1$RA[genus_da1$Taxa==x],na.rm=TRUE))
genus_a2<-data.frame(cbind(Taxa=as.character(unique(genus_da1$Taxa)),RA=genus_a1))
genus_a2$RA<-as.numeric(as.character(genus_a2$RA))

family_a1<-sapply(unique(family_da1$Taxa),function(x) mean(family_da1$RA[family_da1$Taxa==x],na.rm=TRUE))
family_a2<-data.frame(cbind(Taxa=as.character(unique(family_da1$Taxa)),RA=family_a1))
family_a2$RA<-as.numeric(as.character(family_a2$RA))


#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

genus_da2 <-droplevels(subset(genus_da1, genus_da1$Taxa %!in% NA)) # remove asv NAs
genus_da3 <-droplevels(subset(genus_da2, genus_da2$Taxa %in% unique(genus_a2$Taxa[genus_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(genus_da2$Taxa))
length(unique(genus_da3$Taxa)) # 134 genera

family_da2 <-droplevels(subset(family_da1, family_da1$Taxa %!in% NA)) # remove asv NAs
family_da3 <-droplevels(subset(family_da2, family_da2$Taxa %in% unique(family_a2$Taxa[family_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(family_da2$Taxa))
length(unique(family_da3$Taxa)) # 79 families

# Combine family and genus data frames 
da4 <- rbind(genus_da3, family_da3) %>%
  filter(!(Taxa == "CAG-274" & level == "Family")) # Remove CAG-274 as family, listed as both family and genus

```

### iv. Scale variables

scale variables for modeling, combine tenori and palmas, 211 total taxa for data 

```{r}

# Scale variables
da5 <- da4 %>%
   mutate(across(c(temp, rain, age, fruit, neo), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID),
         sex = as.factor(sex)) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)))

length(unique(da5$Taxa)) # 211 taxa total

```

### v. AIC comparison

this loops takes a while to run. 

Compare m_main that includes all predictors except temp and m_full which includes temp for all taxa. Same model comparison as the community-level analysis except not including diet diversity in these models to help simplify the models. 

```{r}

# create lists
taxa_coef <- list()
taxa_aic <- list()
model_results <- list()

# loop
for (i in unique(da5$Taxa)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(da5, Taxa == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- gam(Abundance ~ age_s + temp_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 

      m_main <- gam(Abundance ~ age_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 
      
    }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
     # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      # extract temp coef
      coef <- round(summary(m_full)$p.coeff[3], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[3], 4)
    
      # clac CI'
      ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
      ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
      # bind data
      df <- cbind(ci_95, ci_50)
      
      #create data frames
    aic_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
        Taxa = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3),
        stringsAsFactors = FALSE)

    options(scipen = 999)  

    model_results_df <- data.frame(
      Taxa = i,
      Predictor = names(summary(m_full)$p.coeff[c(2:7)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:7)], 4),
      StdError = round(summary(m_full)$se[c(2:7)], 4),
      stringsAsFactors = FALSE)
    
    
      # save results
    taxa_aic[[i]] <- aic_df
    taxa_coef[[i]] <- results_df
    model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
taxa_aic <- do.call(rbind, taxa_aic)
taxa_coef <- do.call(rbind, taxa_coef)
model_results <- do.call(rbind, model_results)



```

### vi. taxa selection

First, we remove all taxa with explicit convergence warnings, then we remove all taxa that look like they didn't converge based on SE in the model results. 48 taxa did not converge. After removing the taxa that don't converge we tested 163 taxa in total and 93 of those have m_full with a model weight  > 0.5

Then, we select all taxa where m_full (the model with the temp predictor) fits the best. 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(taxa_aic$Taxa[taxa_aic$Warning == "Convergence Warning"]) # 22 do not converge

# remove all taxa where one or more predictor have huge SE
convergence_errors <- model_results %>% 
  filter(StdError > 80) %>%
  pull(Taxa) %>%
  unique() 

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 48 do not converge

# remove error taxa
da6 <- da5 %>% filter(!Taxa %in% errors)
length(unique(da6$Taxa)) # 163 remaining taxa

taxa_coef2 <- taxa_coef %>% filter(!Taxa %in% errors)
length(taxa_coef2$Taxa)

taxa_aic2 <- taxa_aic %>% filter(!Taxa %in% errors)
length(taxa_aic2$Taxa) # 324/2 = 162

# filter for taxa where full model fits best
tax_temp_filter <- taxa_aic2 %>%
  group_by(Taxa) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa) # 93 taxa
length(unique(tax_temp_filter))

filtered_coef <- taxa_coef2 %>%filter(Taxa %in% tax_temp_filter)
length(filtered_coef$Taxa)

filtered_aic <- taxa_aic2 %>%filter(Taxa %in% tax_temp_filter)
length(filtered_aic$Taxa)

```

### vii summary stats

create final table of model results for all taxa that were tested, include coefficients and model weights, use for supplemental table
calculate some descriptive statistics for the results

```{r}

# final table of model results for all tested taxa
levels <- da6 %>% select(Taxa, level, Phylum) %>%
  distinct()

weights <- taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight)

summary <- taxa_coef2 %>%
  left_join(levels, by = "Taxa") %>%
  left_join(weights, by = "Taxa") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         sensitive = ifelse(weight > 0.5, "temp sensitive", "nuetral")) %>%
  rename(model_weight = weight,
         temp_coef = coef,
         std_error = std)

write.csv(summary, "figures/supp/taxonomy_temp_DAA_results.csv", row.names = FALSE)

# descriptive stats
table(summary$sensitive) # 93 temp sensitive
table(summary$level) # 97 genera, 66 families 

temp_sensitive <- summary %>% filter(sensitive == "temp sensitive")
table(temp_sensitive$level) # 37 families, 56 genera
length(unique(temp_sensitive$Phylum)) # 10 phyla
table(temp_sensitive$direction)# 82 negative, 11 positive

phyla_results <- summary %>% filter(level == "Genus")
table(phyla_results$Phylum, phyla_results$sensitive)

```


### viii. Ancestral state rescontruction

Create cladogram that shows the results of the temp DAA for genera with an ancestral state reconstruction using the model weights for the temp model

#### a. data prep 

extract the necessary metadata and filter the phyloseq object for all genera included in the DA temp analysis. Then extract the tree. 

```{r}

# get list of taxa with ne
direction <- filtered_coef %>% 
  mutate(effect_direction = ifelse(coef > 0, "positive", "negative")) %>%
  select(Taxa, effect_direction)

# add temp information to da table
da7 <- da6 %>%
  mutate(taxa_temp_sensitive = ifelse(Taxa %in% tax_temp_filter, "sensitive", "nuetral")) %>%
  left_join(direction, by = "Taxa") %>%
  mutate(effect_direction = case_when(
    is.na(effect_direction) ~ "neutral",
    TRUE ~ effect_direction))
  
# get genus names
genus_names <- da7 %>%
  filter(level == "Genus") %>%
  pull(Taxa) %>%
  unique()

# aggregate by taxa
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))

# filter phyloseq object by taxa 
tree_taxa <- taxa_names(genus_aggregated)[tax_table(genus_aggregated)[, "Genus"] %in% genus_names]
temp_genus_phylo <- prune_taxa(tree_taxa, genus_aggregated)

# extract the tree
genus_tree <- phy_tree(temp_genus_phylo)
genus_tax <- as.data.frame(tax_table(temp_genus_phylo))

# prep final tree metadata 
genus_meta <- da7 %>%
  filter(level == "Genus") %>%
  select(Taxa, taxa_temp_sensitive, effect_direction) %>%
  distinct() %>%
  left_join(taxa_aic2 %>% filter(model == "m_full") %>% select(Taxa, weight), by = "Taxa")

# create phylum groups for clade labels
phylum_groups <- genus_tax %>%
  rownames_to_column("OTU") %>%
  mutate(Genus = as.character(Genus), Phylum = as.character(Phylum)) %>%
  filter(!is.na(Phylum), !is.na(Genus)) %>%
  group_by(Phylum) %>%
  summarise(genus_list = list(Genus)) %>%
  unnest_longer(genus_list, values_to = "Genus")

```

#### b. assign tip labels

First assign the tip labels as genus names instead of otu numbers. Add the metadata for temp sensitivity and change the tip labels to the model weight varible

```{r}

# tip labels are still otu numbers
head(genus_tree$tip.label)

# get genus names for tip labels
tip_labels <- as.character(tax_table(temp_genus_phylo)[, "Genus"])

# assign labels 
genus_tree$tip.label <- tip_labels
head(genus_tree$tip.label) # check tip labels
  
# look a tree data 
tree_plot <- ggtree(genus_tree)
tree_data <- tree_plot$data
tree_data <- tree_data %>% mutate(Taxa = label)

# check that Taxa and tip labels match
all(genus_meta$Taxa %in% genus_tree$tip.label) 

# add metadata to tree data
tree_data <- tree_data %>% left_join(genus_meta, by = "Taxa") 
tip_data <- tree_data %>% filter(isTip == "TRUE")

# now make tip labels the model weight
genus_tree$tip.label <- tip_data$weight
head(genus_tree$tip.label)

```

####  c. ancestral state reconstruction

reconstruction returns the maximum likelihood estimates of the model weights for each node, assumes brownian motion. Higher weights are more temperature sensitive than lower weights. 

```{r}

# run ace function
ace_result <- ace(tip_data$weight, genus_tree, type = "continuous")

# extract node probs
node_df <- data.frame(
  node = as.numeric(names(ace_result$ace)),
  ace_weight = ace_result$ace) # ignore warning will fix in next step

# assign node probs to node
node_df$node <- (Ntip(genus_tree) + 1):(Ntip(genus_tree) + Nnode(genus_tree))  # Node numbers

tree_data <- left_join(tree_data, node_df, by = "node") %>%
  mutate(ace_weight = ifelse(isTip, weight, ace_weight)) # assign weights to tips for plotting branch color

# summary stats - add this to tree to get node labels to inspect weights of ancestral nodes 
#geom_text2(
  #aes(subset = !isTip, label = node),
  #size = 3,
 # hjust = -0.3,
  #color = "black"
#) 

```

#### d. plot tree

labeled phyla have > 1% rel abundance 

```{r}

tree <- ggtree(genus_tree, branch.length = "none") %<+% tree_data +
  geom_tree(aes(color = ace_weight), size = 1) +
  
  scale_x_continuous(
  expand = expansion(mult = c(0, 0)),
  limits = c(NA, 45)  # or whatever max x value you need
) +
  
  scale_color_gradient(low = "#66bd63", high = "#fee08b", name = "Model weight") +
  
  ggnewscale::new_scale_color() + # add a new color scale for the tip circles
  
  geom_tiplab(aes(label = Taxa), size = 3.5, offset = 0.5) +
  
  geom_point2(aes(subset = isTip, color = effect_direction), shape = 19, size = 4, stroke = 0, show.legend = FALSE) +

  scale_color_manual(values = c("positive" = "#80b1d3", "negative" = "#fb8072", "neutral"  = "grey60"), name = "Effect direction") +
  
  geom_text2(aes(subset = isTip, label = case_when(effect_direction == "positive" ~ "+", effect_direction == "negative" ~ "–")),
  size = 3.6, vjust = 0.4, hjust = 0.5, color = "white") +
  
  theme_tree2() +
  
  theme(legend.position = c(0.02, 0.98),
        legend.justification = c(0, 1),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()) +
  
  # Direction effect legend
  annotate("text", x = 1.2, y = 82, label = "Effect direction", size = 3.7, hjust = 0) +
  
  annotate("point", x = 1.5, y = 80.2, shape = 19, size = 4, color = "#80b1d3") +
  annotate("text", x = 2.1, y = 80.25, label = "positive", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 80.3, label = "+", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 78.4, shape = 19, size = 4, color = "#fb8072") +
  annotate("text", x = 2.1, y = 78.45, label = "negative", size = 3.3, hjust = 0) +
  annotate("text", x = 1.5, y = 78.5, label = "–", size = 4, color = "white") +
  
  annotate("point", x = 1.5, y = 76.6, shape = 19, size = 4, color = "grey60") +
  annotate("text", x = 2.1, y = 76.65, label = "no effect", size = 3.3, hjust = 0) +
  
  # Bacteroidota
  annotate("segment", x = 32.5, xend = 32.5, y = 31, yend = 43, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 31, yend = 31, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 43, yend = 43, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 37, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 31, y = 1, yend = 1, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 1, label = "Bacteroidota", size = 5.5, hjust = 0) +
  
  # Proteobacteria
  annotate("segment", x = 32.5, xend = 32.5, y = 7, yend = 30, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 7, yend = 7, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 30, yend = 30, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 18, label = "Proteobacteria", size = 5.5, hjust = 0) +
  
  #Actinobacteriota
  annotate("segment", x = 32.5, xend = 32.5, y = 47, yend = 49, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 47, yend = 47, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 49, yend = 49, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 48, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 67, yend = 65, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 67, yend = 67, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 65, yend = 65, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 66, label = "Actinobacteriota", size = 5.5, hjust = 0) +
  
  #Firmicutes
  annotate("segment", x = 32.5, xend = 32.5, y = 68, yend = 97, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 68, yend = 68, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 97, yend = 97, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 82.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  annotate("segment", x = 32.5, xend = 32.5, y = 53, yend = 64, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 53, yend = 53, color = "black", size = 0.6) +
  annotate("segment", x = 32.5, xend = 31, y = 64, yend = 64, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 59.5, label = "Firmicutes", size = 5.5, hjust = 0) +
  
  #Campylobacterota
  annotate("segment", x = 32.5, xend = 31, y = 6, yend = 6, color = "black", size = 0.6) +
  annotate("text", x = 33, y = 6, label = "Campylobacterota", size = 5.5, hjust = 0) 
  
tree

ggsave("figures/fig3.pdf", tree, width =8, height = 12.5)


```


## C. Pathway differential abundance testing

### i. Prep data

starting with 547 pathways

```{r}

# wrangle pathway levels
pathway_levels <- as.data.frame(MetaCyc_pathway_map) %>%
  rownames_to_column("path_id") %>%
   select(pathway, path_id)

# wrangle clr abundances
path_abund <- path_clr %>%
  data.frame() %>%
  rownames_to_column("sample") %>%
  gather(key = "pathway", value = "clr_abundance", -sample) %>%
  left_join(path_m_filtered, by = "sample") %>%
  mutate(pathway = str_remove(pathway, "^X")) %>% 
  separate(pathway, into = c("path_id", "pathway_name"), sep = "\\.\\.", extra = "merge") %>%
  mutate(path_id = str_replace_all(path_id, "\\.", "-")) %>%
  left_join(pathway_levels, by = "path_id") %>%
  rename(path_name = pathway) %>%
  select(sample, path_id, path_name, clr_abundance:reads) %>%
  filter(path_id != "ALL-CHORISMATE-PWY") # remove path with no lvl 2 assignment 
  
length(unique(path_abund$path_id)) # 547 pathways
table(path_abund$sample) # each sample has 547 paths

# scale variables
path_abund <- path_abund %>%
  mutate(across(c(age, temp, rain, fruit, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         sex = as.factor(sex),
         ID = as.factor(ID))

```

### ii. AIC comparison

```{r}

path_coef <- list()
path_aic <- list()
path_model_results <- list()

for (i in unique(path_abund$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df
    path_model_results[[i]] <- model_results_df

  })
}

# combine results
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)
path_model_results <- do.call(rbind, path_model_results)


```

### iii. select taxa

164 do not converge so we tested 383 total (started with 547)

Temp model fits best for 169 pathways which means that it doesn't fit best for 214

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic$path_id[path_aic$Warning == "Convergence Warning"]) # 164 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no huge SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 164 do not converge

# remove error taxa
path_abund2 <- path_abund %>% filter(!path_id %in% errors)
length(unique(path_abund2$path_id)) # 383 remaining 

path_coef2 <- path_coef %>% filter(!path_id %in% errors)
length(path_coef2$path_id)

path_aic2 <- path_aic %>% filter(!path_id %in% errors)
length(path_aic2$path_id) 

# filter for taxa where full model fits best
temp_filter <- path_aic2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(temp_filter)) # temp model fits best for 169 pathways

# filter df's for where temp model fits best 
path_abund3 <- path_abund2 %>% filter(path_id %in% temp_filter)
length(unique(path_abund3$path_id))  

path_coef3 <- path_coef2 %>% filter(path_id %in% temp_filter)
length(path_coef3$path_id)

path_aic3 <- path_aic2 %>% filter(path_id %in% temp_filter)
length(path_aic3$path_id)

```

### iv. supplemental table

table with modeling information for all pathways for the supplement

```{r}

# final table of model results for all tested pathways
levels <- path_abund2 %>% select(path_name, path_id) %>%
  distinct()

weights <- path_aic2 %>% filter(model == "m_full") %>% select(path_id, weight)

groupings <- read.csv("data/path_group_assignment.csv")

path_summary <- path_coef2 %>%
  left_join(levels, by = "path_id") %>%
  left_join(weights, by = "path_id") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         sensitive = ifelse(weight > 0.5, "temp sensitive", "nuetral")) %>%
  rename(model_weight = weight,
         temp_coef = coef,
         std_error = std,
         pathway = path_name) %>%
  left_join(groupings, by = "pathway") %>%
  mutate(group = case_when(sensitive == "temp sensitive" & is.na(group) ~ "no group assignment",
         TRUE ~ group))

write.csv(path_summary, "figures/supp/pathway_temp_DAA_results.csv", row.names = FALSE)

```

### v. plot figure

#### a. data prep

```{r}

temp_path_info <- path_coef3 %>%
  left_join(pathway_levels, by = "path_id") %>%
  select(path_id, pathway, coef:high_ci_50) %>% 
  left_join(groupings, by = "pathway") %>%
  filter(!is.na(group)) %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative"),
         path_id = reorder(path_id, -coef),
         pathway = str_remove_all(pathway, " \\(.*?\\)"),
         pathway = str_replace_all(pathway, "&beta;", "beta"),
         pathway = case_when(pathway == "pyruvate fermentation to acetate and-lactate I" ~ "pyruvate fermentation to acetate and (S)-lactate I",
                             pathway == "superpathway of2-lipid A biosynthesis" ~ "superpathway of (Kdo)2-lipid A biosynthesis",
                             pathway == "fatty acid elongation -- saturated" ~ "fatty acid elongation",
                             pathway == "palmitoleate biosynthesis I-dodec-5-enoate)" ~ "palmitoleate biosynthesis I",
                             TRUE ~pathway))

# make functional groups
lipid <- temp_path_info %>% filter(group == "lipid catabolism and remodeling") %>% arrange(desc(coef)) %>%
  mutate(order = factor(pathway, levels = pathway))

redox <- temp_path_info %>% filter(group == "redox maintenance") %>% arrange(desc(coef)) %>%
  mutate(order = factor(pathway, levels = pathway))

nucleotide <- temp_path_info %>% filter(group == "nucleotide mobilization") %>% arrange(desc(coef)) %>%
  mutate(order = factor(pathway, levels = pathway))

immune <- temp_path_info %>% filter(group == "immune regulation") %>% arrange(desc(coef)) %>%
  mutate(order = factor(pathway, levels = pathway))


# order for oxygen groups
oxygen <- temp_path_info %>% filter(group == "oxygen-driven metabolism and carbohydrate fermentation")

subgroup_order <- c("starch degradation III", "D-arabinose degradation I", "D-glucarate degradation I", "superpathway of D-glucarate and D-galactarate degradation")

all_subgroups <- unique(oxygen$subgroup)

custom_order <- c(subgroup_order, setdiff(all_subgroups, subgroup_order))

oxygen <- oxygen %>%
  mutate(subgroup = factor(subgroup, levels = custom_order)) %>%
  arrange(subgroup, desc(coef)) %>%
  mutate(order = factor(pathway, levels = pathway))


```

#### b. coefficient plots

warnings when plotting are from special characters

```{r}

########## lipid metabolism #############

lipid_p <- make_coef_plot(lipid) +
  ggtitle("A. Lipid metabolism") +
  ylab("") +
  ylim(-5, 5) +
  theme(axis.text.x = element_blank(),
         plot.margin = margin(t = 5, r = 120, b = 5, l = 5)) +
  annotate("segment", x = "superpathway of glyoxylate cycle and fatty acid degradation",  xend = "9-cis, 11-trans-octadecadienoyl-CoA degradation", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "fatty acid salvage", y = 4.85, label = "decreased catabolism\nand recycling", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "stearate biosynthesis II",  xend = "petroselinate biosynthesis", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "gondoate biosynthesis", y = 4.85, label = "increased synthesis\nand remodeling", hjust = 0, size = 5, color = "#80b1d3") 
lipid_p

########## oxygen metabolism #############

oxygen_p <- make_coef_plot(oxygen) +
  ggtitle("B. Oxygen-driven metabolism and fermentation") +
  ylab("") +
  ylim(-5, 5) +
  theme(axis.text.x = element_blank(),
        plot.margin = margin(t = 5, r = 120, b = 5, l = 5)) +
  annotate("segment", x = "pyruvate fermentation to acetate and (S)-lactate I",  xend = "formaldehyde assimilation II", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "glycogen degradation II", y = 4.85, label = "increased anerobic\nmetabolism", hjust = 0, size = 5, color = "#80b1d3") +
  annotate("segment", x = "superpathway of glycolysis, pyruvate dehydrogenase, TCA, and glyoxylate bypass",  xend = "glyoxylate cycle", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "superpathway of glycol metabolism and degradation", y = 4.85, label = "decreased aerobic\nmetabolism", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "superpathway of D-glucarate and D-galactarate degradation",  xend = "starch degradation III", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "D-glucarate degradation I", y = 4.85, label = "decreased complex\ncarbohydrate fermentation", hjust = 0, size = 5, color = "#fb8072") 

oxygen_p

########## Redox maintenance #############

redox_p <- make_coef_plot(redox) +
  ggtitle("C. Redox maintenance") +
  ylab("") +
  ylim(-5, 5) +
  theme(plot.margin = margin(t = 5, r = 120, b = 5, l = 5)) +
  annotate("segment", x = "assimilatory sulfate reduction IV",  xend = "L-methionine biosynthesis IV", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "superpathway of methylglyoxal degradation", y = 4.85, label = "reduced redox\nmaintenance", hjust = 0, size = 5, color = "#fb8072") +
  annotate("segment", x = "tetrapyrrole biosynthesis I",  xend = "L-ascorbate degradation I", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "pentose phosphate pathway I", y = 4.85, label = "increased redox\nmaintenance", hjust = 0, size = 5, color = "#80b1d3") 

redox_p

########## nucleotide mobilization #############

nucleotide_p <- make_coef_plot(nucleotide) +
  ggtitle("D. Nucleotide mobilization") +
  ylab("") +
  ylim(-5, 5) +
  theme(plot.margin = margin(t = 5, r = 120, b = 5, l = 5)) +
  annotate("segment", x = "pyrimidine deoxyribonucleosides salvage",  xend = "adenosine nucleotides degradation I", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = "UMP biosynthesis III", y = 4.85, label = "increased synthesis\nand salvage", hjust = 0, size = 5, color = "#80b1d3")

nucleotide_p 

########## immune regulation #############

immune_p <- make_coef_plot(immune) +
  ggtitle("E. Immune regulation") +
  ylab("") +
  ylim(-5, 5) +
  theme(plot.margin = margin(t = 5, r = 120, b = 5, l = 5)) +
  annotate("segment", x = "acetyl-CoA fermentation to butanoate II",  xend = "L-glutamate degradation XI", y = 4.7, yend = 4.7, color = "#fb8072", size = 0.6) +
  annotate("text", x = "L-glutamate degradation VII", y = 4.85, label = "decreased\nanti-inflammatory\nsignals", hjust = 0, size = 5, color = "#fb8072") + 
  annotate("segment", x = "(5Z)-dodecenoate biosynthesis I",  xend = "L-tryptophan biosynthesis", y = 4.7, yend = 4.7,
           color = "#80b1d3", size = 0.6) +
  annotate("text", x = 1.5, y = 4.85, label = "increased\nanti-inflammatory\nsignals", hjust = 0, size = 5, color = "#80b1d3")

immune_p

#### combine plots #####

combined <- ggarrange(lipid_p, oxygen_p, redox_p, nucleotide_p, immune_p, ncol = 3, nrow = 2, align = "v")

ggsave("figures/fig4.pdf", combined, width =32, height = 13)


# find midpoint betwen two discrete values 
#levels <- levels(immune$pathway)

# Get numeric positions of those two
#a <- which(levels == "enterobacterial common antigen biosynthesis")
#b <- which(levels == "superpathway of (Kdo)2-lipid A biosynthesis")

# Find midpoint
#mid <- (a + b) / 2

```



# 4. Neopterin * Temperature analysis

## A. Community-level modeling

Compare m_main that has the base predictors plus the temp predictor and m_full which has an interaction between neopterin and temperature. Both model include an offset of total reads to control for seq depth.

### i. Taxonomy

Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed.

```{r}

# create empty lists
tax_coef_int <- list() 
tax_aic_int <- list()

com_features <- colnames(tax_model)[c(15, 17:19)] # list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(filtered_reads))"))
      
    m_main <- lmer(formula_main, data = tax_model)
    m_full <- lmer(formula_full, data = tax_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[10,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[10,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    tax_aic_int[[i]] <- aic_df
    tax_coef_int[[i]] <- results_df

}

# combine results 
tax_aic_int <- do.call(rbind, tax_aic_int)
tax_coef_int <- do.call(rbind, tax_coef_int)


#### model observed ######

m_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

m_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = tax_model)

AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(m_full)$coeff[10,1], 4)
std <- round(summary(m_full)$coeff[10,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

tax_coef_int <- rbind(tax_coef_int, results_df)

tax_aic_int <- rbind(tax_aic_int, aic_df) 

```

### ii. Pathways

```{r}
path_coef_int <- list() 
path_aic_int <- list()

com_features <- colnames(tax_model)[c(15:19)] # list of features to model

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s * neo_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = path_model)
      m_full <- lmer(formula_full, data = path_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[10,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[10,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    path_aic_int[[i]] <- aic_df
    path_coef_int[[i]] <- results_df

}

# combine results 
path_aic_int <- do.call(rbind, path_aic_int)
path_coef_int <- do.call(rbind, path_coef_int)
```


### iii. Plot model results

#### a. data prep

```{r}

## taxa data prep ##

plot_coef_int <- tax_coef_int %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_aic_int <- tax_aic_int %>%
  select(metric, model, weight) %>%
  mutate(model = case_when(model == "o_full" ~ "m_full",
                           model == "o_main" ~ "m_main",
                           TRUE ~ model)) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_coef_int$order))) %>%
  arrange(order)

## path data prep ##

plot_path_coef_int_observed <- path_coef_int %>%
  filter(metric == "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_coef_int <- path_coef_int %>%
  filter(metric != "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))


plot_path_aic_int_observed <- path_aic_int %>%
  filter(metric == "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%    
  mutate(order = factor("observed"))
  
path_order <- c("PC3", "PC2", "shannon", "PC1")  

plot_path_aic_int <- path_aic_int %>%
  filter(metric != "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%  
  mutate(order = factor(metric, levels = levels(plot_path_coef_int$order))) %>%
  arrange(order)


levels <- levels(plot_aic_int$order)
which(levels == "PC1") # height for plot label


```

#### b. combined plot

```{r}

p1 <- make_coef_plot(plot_coef_int) +
  ylim(-10, 10) +
  ggtitle("A. Taxonomy interaction coefficients") +
  ylab("")

p2 <- make_coef_plot(plot_path_coef_int) +
  ylim(-36, 36) +
  ggtitle("C. Metabolic pathway interaction coefficients") +
  ylab("")

p3 <- make_coef_plot(plot_path_coef_int_observed) + 
  ylim(-165, 165) +
  ylab("Effect size")

p4 <- model_weight_plot(plot_aic_int) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 5.45, label = "interaction model", color = "#fee08b", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 5.45, label = "temp model", color = "#66bd63", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p5 <- model_weight_plot(plot_path_aic_int) +
  ggtitle("D. Metabolic pathway model weights") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p6 <- model_weight_plot(plot_path_aic_int_observed) +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
  labels = c("100", "50", "0", "50", "100"),
  limits = c(-1, 1)) +
  xlab("model weight")

int_com_temp_plot <- ggarrange(p1, p4, p2, p5, p3, p6, ncol = 2, nrow = 3, 
          widths = c(1, 1),
          align = "v",
          heights = c(1, 1, 0.4))

ggsave("figures/fig5.pdf", int_com_temp_plot, width = 12, height = 8)

```

## B. Taxonomic Differential abundance testing

### i. AIC comparison

Take all temp taxa (taxa that had a relationship with temp) and fit with m_main model with all the base predictors plus temp and m_full model with base predictors and a temp * neopterin interaction. 

```{r}

# filter abundance for temp taxa
temp_abund <- da6 %>% filter(Taxa %in% tax_temp_filter)
length(unique(temp_abund$Taxa)) # 93

# create lists
tax_neo_coef <- list()
tax_neo_aic <- list()
tax_neo_model_results <- list()

for (i in unique(temp_abund$Taxa)) {
  tryCatch({
    
    # subset based on taxa
    data <- droplevels(subset(temp_abund, Taxa == i))
    
    # convergence warning variable
    warning_msg <- "No Warning"
    
    withCallingHandlers({

      m_full <- gam(Abundance ~ age_s + temp_s * neo_s + rain_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 

      m_main <- gam(Abundance ~ age_s + rain_s + temp_s + sex + group + fruit_s + s(ID, bs = "re") + offset(log(filtered_reads)), data = data, method = "REML", family = nb(link = "log")) 
      
    }, warning = function(w) {
   
      warning_msg <- conditionMessage(w)
  
      if (grepl("Fitting terminated with step failure", warning_msg)) {
      warning_msg <<- "Convergence Warning"  # update warning variable
      } 
    })
    
     # compare with AICctab
      AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
      AICc$model <- rownames(AICc)
      
      # extract temp coef
      coef <- round(summary(m_full)$p.coeff[9], 4)
    
      # extract temp ci's
      std <- round(summary(m_full)$se[9], 4)
    
      # clac CI'
      ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
      ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
      # bind data
      df <- cbind(ci_95, ci_50)
      
      #create data frames
    aic_df <- data.frame(
        Taxa = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3),
        Warning = warning_msg,
        stringsAsFactors = FALSE)
    
    results_df <- data.frame(
        Taxa = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3),
        stringsAsFactors = FALSE)

    options(scipen = 999)  

    model_results_df <- data.frame(
      Taxa = i,
      Predictor = names(summary(m_full)$p.coeff[c(2:9)]),
      Coefficient = round(summary(m_full)$p.coeff[c(2:9)], 4),
      StdError = round(summary(m_full)$se[c(2:9)], 4),
      stringsAsFactors = FALSE)
    
    
      # save results
    tax_neo_aic[[i]] <- aic_df
    tax_neo_coef[[i]] <- results_df
    tax_neo_model_results[[i]] <- model_results_df
      
  }, error = function(e) {

    message(paste("Error with Taxa:", i, "- Skipping."))  # skip to next taxa if error
    
  })
}

# combine results
tax_neo_aic <- do.call(rbind, tax_neo_aic)
tax_neo_coef <- do.call(rbind, tax_neo_coef)
tax_neo_model_results <- do.call(rbind, tax_neo_model_results)

```

### ii. taxa selection

same as previous AIC taxa filtered. Remove convergence warnings and any taxa that look like they didn't converge from SE. Filter for taxa where m_full (model with temp * neo interaction) have > .50 model weight. 20 taxa left. 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(tax_neo_aic$Taxa[tax_neo_aic$Warning == "Convergence Warning"]) # 2 warnings

# remove all taxa where one or more predictos have huge SE
convergence_errors <- tax_neo_model_results %>%
  filter(StdError > 100) %>%
  pull(Taxa) %>%
  unique()

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 2 do not converge

# remove error taxa
temp_abund2 <- temp_abund %>% filter(!Taxa %in% errors)
length(unique(temp_abund2$Taxa)) # 91 remaining taxa

tax_neo_coef2 <- tax_neo_coef %>% filter(!Taxa %in% errors)
length(tax_neo_coef2$Taxa)

tax_neo_aic2 <- tax_neo_aic %>% filter(!Taxa %in% errors)
length(tax_neo_aic2$Taxa) # 182/2 = 91

# filter for taxa where full model fits best
tax_neo_filter <- tax_neo_aic2 %>%
  group_by(Taxa) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(Taxa) 
length(tax_neo_filter) # 20 taxa

# filter abundances for temp taxa
tax_neo_abund <- temp_abund2 %>% filter(Taxa %in% tax_neo_filter)
length(unique(tax_neo_abund$Taxa)) 

tax_filtered_neo_coef <- tax_neo_coef2 %>% filter(Taxa %in% tax_neo_filter)
length(tax_filtered_neo_coef$Taxa)

tax_filtered_neo_aic <- tax_neo_aic2 %>% filter(Taxa %in% tax_neo_filter)
length(tax_filtered_neo_aic$Taxa) 

```

### iii. get model predictions

```{r}

# Create neo bins
tax_neo_abund$neo_bin <- cut(tax_neo_abund$neo_s, breaks = 3, labels = c("Bin1", "Bin2", "Bin3"), include.lowest = TRUE)

bin_limits <- levels(cut(tax_neo_abund$neo_s, breaks = 3, include.lowest = TRUE)) # look at bin range
bin_avg <- tapply(tax_neo_abund$neo_s, tax_neo_abund$neo_bin, mean, na.rm = TRUE) # use for co plot

# Create temp seq
temp_seq <- seq(min(tax_neo_abund$temp_s, na.rm = TRUE), max(tax_neo_abund$temp_s, na.rm = TRUE), length.out = 100)

# list
neo_preds <- list()

# loop
for (i in unique(tax_neo_abund$Taxa)) {
  
  data <- droplevels(subset(tax_neo_abund, Taxa == i))
  
  m <- gam(Abundance ~ age_s + temp_s * neo_s + rain_s + rain_s + sex + group + fruit_s + offset(log(filtered_reads)), 
           data = data, 
           method = "REML", 
           family = nb(link = "log"))
  
  coef <- round(summary(m)$p.coeff[9], 4)
  
  # neo_s values and bins
  neo_values <- c(-0.1001032,  1.0280093,  2.5125214)
  bins <- paste0("neopterin_avg", 1:3)
  
  # Generate predictions for each neo_s value
  preds <- lapply(seq_along(neo_values), function(j) {
    neo_s <- neo_values[j]
    bin <- bins[j]
    
    # preds dataset
    pred <- expand.grid(
      temp_s = temp_seq,
      neo_s = neo_s,
      age_s = mean(data$age_s, na.rm = TRUE),
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = unique(data$group)[1],
      sex = unique(data$sex)[1],
      filtered_reads = mean(data$filtered_reads, na.rm = TRUE)
    )
    
    # get model predictions
    preds_model <- predict(m, newdata = pred, type = "link", se.fit = TRUE)
    
    # store preds and CI's 
    df <- data.frame(
      temp_s = pred$temp_s,  
      neo_s = neo_s,
      fit = preds_model$fit,
      lower = preds_model$fit - 1.96 * preds_model$se.fit,
      upper = preds_model$fit + 1.96 * preds_model$se.fit,
      bin = bin,
      coef = coef
    )
    
    return(df)
  })
  
  # combine predictions for all bins 
  preds_df <- do.call(rbind, preds)
  preds_df$Taxa <- i 
  
  # store preds
  neo_preds[[i]] <- preds_df
}

# Combine all predictions into a single dataframe
tax_neo_preds <- do.call(rbind, neo_preds)
row.names(neo_preds) <- NULL # clean up row names

```

### vi. plot by Taxa

Plot with the CI's for each taxa for the supplement

```{r}

pdf("figures/supp/taxa_interaction_plottedpreds.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(tax_neo_preds$Taxa)) {
  
  data <- subset(tax_neo_preds, Taxa == i)

  p <- ggplot(data, aes(x = temp_s, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "grey") +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(
      title = paste("predicted Abundance for", i),
      x = "temperature (scaled)",
      y = "predicted log abundance") 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```

### v. Plot by effect direction

#### a. data prep
```{r}

########### Prep data ################

mean_neo <- mean(tax_neo_abund$neo)
sd_neo <- sd(tax_neo_abund$neo) 

neo_reverse <- function(x) {
  return(round(x * 2 * sd_neo + mean_neo))  # reverse scale in figure axes
}

bin_limits <- levels(cut(tax_neo_abund$neo_s, breaks = 3, include.lowest = TRUE)) # look at bin range
neo_reverse(3.26)

tax_effect_type <- tax_neo_preds %>%
  group_by(Taxa, bin) %>%
  do({m <- lm(fit ~ temp_s, data = .)
    slope <- coef(m)["temp_s"]
    data.frame(slope = slope)}) %>%
  ungroup() %>%
  pivot_wider(names_from = bin, values_from = slope,
                     names_prefix = "slope_") %>%
  mutate(effect_type = case_when(slope_neopterin_avg1 > 0 & slope_neopterin_avg3 > 0 ~ "positive_amplifying",
                                 slope_neopterin_avg1 < 0 & slope_neopterin_avg3 < 0 ~ "negative_amplifying",
                                 slope_neopterin_avg1 < 0 & slope_neopterin_avg3 > 0 ~ "positive_inverse",
                                 TRUE ~ "other")) %>%
  select(Taxa, effect_type)


tax_neo_preds2 <- tax_neo_preds %>% 
  mutate(bin = case_when(bin == "neopterin_avg1" ~ "neopterin: 108-1915 (nmol/L)",
                         bin == "neopterin_avg2" ~ "neopterin: 1915-3711 (nmol/L)",
                         bin == "neopterin_avg3" ~ "neopterin: 3711-5525 (nmol/L)")) %>%
  left_join(tax_effect_type, by = "Taxa")


pos <- tax_neo_preds2 %>% filter(effect_type == "positive_amplifying")
neg <- tax_neo_preds2 %>% filter(effect_type == "negative_amplifying")
inverse <- tax_neo_preds2 %>% filter(effect_type == "positive_inverse")

length(unique(pos$Taxa))
length(unique(neg$Taxa))
length(unique(inverse$Taxa))


```

#### b. plot

```{r}

# bin values
unique_bins <- unique(tax_neo_preds2$bin)

# to rescale x axis
mean_temp <- mean(tax_neo_abund$temp)
sd_temp <- sd(tax_neo_abund$temp) 

################### NEG PLOT ######################

y_min <- min(neg$fit)
y_max <- max(neg$fit)

neg_taxa <- unique(neg$Taxa) # unique taxa
neg_colors <- setNames(generate_shades("#fb8072", length(neg_taxa)), neg_taxa) # color assignment

neg_list <- list()

# loop
for (i in unique_bins) {
  
  neg_filtered <- neg[neg$bin == i, ]
  
    y_tick_labels <- if (i == "neopterin: 108-1915 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(neg_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = "", x = "", title = paste(i)) +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    scale_x_continuous(breaks = seq(min(tax_neo_preds2$temp_s), max(tax_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) +
    scale_color_manual(values = neg_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  neg_list[[as.character(i)]] <- bin_plot
  
}

# legend
neg_taxa <- paste(c("<b>Included taxa:</b>", unique(neg$Taxa)), collapse = "<br>")

neg_labels <- richtext_grob(neg_taxa,
  x = 0, y = 0.85, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 10),
  box_gp = gpar(col = NA))  # no box around the text

neg_plot <- ggarrange(do.call(ggarrange, c(neg_list, nrow = 1)),
  as_ggplot(neg_labels), 
  ncol = 2,
  widths = c(3, 0.6))

neg_plot

################## positive amplifying###########################

y_min <- min(pos$fit)
y_max <- max(pos$fit)

pos_taxa <- unique(pos$Taxa) # unique taxa
pos_colors <- setNames(generate_shades("#66bd63", length(pos_taxa)), pos_taxa) # color assignment

pos_list <- list()

# loop
for (i in unique_bins) {
  
  pos_filtered <- pos[pos$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 108-1915 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(pos_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = "", x = "") +
    scale_y_continuous(limits = c(y_min, y_max),
                       labels = function(x) sprintf("  %s", x)) + # a two spaces in front of ticks to line up axes with other two plots with neg scales
    scale_x_continuous(breaks = seq(min(tax_neo_preds2$temp_s), max(tax_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) +
    scale_color_manual(values = pos_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5)) 
  
  pos_list[[as.character(i)]] <- bin_plot
  
}

# legend
pos_taxa <- paste(c("<b>Included taxa:</b>", unique(pos$Taxa)), collapse = "<br>")

pos_labels <- richtext_grob(pos_taxa,
  x = 0, y = 0.85, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 10),
  box_gp = gpar(col = NA))  # no box around the text

pos_plot <- ggarrange(do.call(ggarrange, c(pos_list, nrow = 1)),
  as_ggplot(pos_labels), 
  ncol = 2,
  widths = c(3, 0.6))

pos_plot

################### positive inverse ######################

y_min <- min(inverse$fit)
y_max <- max(inverse$fit)

inverse_taxa <- unique(inverse$Taxa) # unique taxa
inverse_colors <- setNames(generate_shades("#80b1d3", length(inverse_taxa)), inverse_taxa) # color assignment

inverse_list <- list()

# loop
for (i in unique_bins) {
  
  inverse_filtered <- inverse[inverse$bin == i, ]
  
    y_tick_labels <- if (i == "neopterin: 108-1915 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(inverse_filtered, aes(x = temp_s, y = fit, color = Taxa)) +
    geom_line() +
    labs(y = "", x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) + 
    scale_x_continuous(breaks = seq(min(tax_neo_preds2$temp_s), max(tax_neo_preds2$temp_s), length.out = 6), 
      labels = inverse_scale) +
    scale_color_manual(values = inverse_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  inverse_list[[as.character(i)]] <- bin_plot
  
}

# legend
inverse_taxa <- paste(c("<b>Included taxa:</b>", unique(inverse$Taxa)), collapse = "<br>")

inverse_labels <- richtext_grob(inverse_taxa,
  x = 0, y = 0.85, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 10),
  box_gp = gpar(col = NA))  # no box around the text

inverse_plot <- ggarrange(do.call(ggarrange, c(inverse_list, nrow = 1)),
  as_ggplot(inverse_labels), 
  ncol = 2,
  widths = c(3, 0.6))

inverse_plot<- annotate_figure(inverse_plot, bottom = text_grob("temperature (°C)", size = 14)) 

inverse_plot

######## combine ##########

pos_slope <- ggarrange(pos_plot, inverse_plot, 
                           nrow = 2, ncol = 1, align = "v",
                           labels = c("B", "C"))
pos_slope <- annotate_figure(pos_slope, left = text_grob("predicted\nlog abundance", size = 12))  

neg_slope <- ggarrange(neg_plot, 
                       nrow = 1, ncol = 1, 
                       labels = "A")
neg_slope <- annotate_figure(neg_slope, left = text_grob("predicted\nlog abundance", size = 12)) 

spacer <- ggplot() + theme_void()

combined_plot <- ggarrange(neg_slope, spacer, pos_slope, 
                           nrow = 3, ncol = 1,
                           heights = c(1, 0.3, 2))

ggsave("figures/fig6.pdf", combined_plot, width =13, height = 9)

```




### vii. summary table

add interaction type to summary predictions and also do this for all the temp taxa tested not just the ones were the interaction effect has >50% of model weight

```{r}

# final table of model results for all tested taxa
levels <- tax_neo_abund %>% select(Taxa, level, Phylum) %>%
  distinct()

weights <- tax_filtered_neo_aic %>% filter(model == "m_full") %>% select(Taxa, weight)

neo_int_summary <- tax_filtered_neo_coef %>%
  left_join(levels, by = "Taxa") %>%
  left_join(weights, by = "Taxa") %>%
  rename(model_weight = weight,
         temp_coef = coef,
         std_error = std)

write.csv(summary, "figures/supp/taxonomy_interaction_DAA_results.csv", row.names = FALSE)


```


## C. Pathway differential abundance testing
### i. AIC comparions

```{r}

path_coef_int <- list()
path_aic_int <- list()
path_model_results_int <- list()

for (i in unique(path_abund3$path_id)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, path_id == i))

    # Default warning message
    warning_msg <- "No Warning"

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s + (1 | ID),
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s + (1 | ID),
                   data = data)

    # Check for convergence warning
    if (!is.null(m_full@optinfo$conv$lme4$messages)) {
      warning_msg <- "Convergence Warning"
    }

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 1], 6)
    std_value <- round(summary(m_full)$coefficients["temp_s:neo_s", 2], 6)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic_int[[i]] <- aic_df
    path_coef_int[[i]] <- results_df
    path_model_results_int[[i]] <- model_results_df

  })
}

# combine results
path_aic_int <- do.call(rbind, path_aic_int)
path_coef_int <- do.call(rbind, path_coef_int)
path_model_results_int <- do.call(rbind, path_model_results_int)

```

### ii. patheay selection

tested 169, 53 of those didn't converge so tested 116, out of 116 the interaction model first best for 65 of them 

```{r}

# Remove taxa with convergence warnings 
convergence_warnings <- unique(path_aic_int$path_id[path_aic_int$Warning == "Convergence Warning"]) # 53 do not converge
length(convergence_warnings)

# remove all taxa where one or more predictor have huge SE
convergence_errors <- path_model_results_int %>% 
  filter(StdError > 100) %>%
  pull(path_id) %>%
  unique() # no big SE's

errors <- c(convergence_warnings, convergence_errors)
length(unique(errors)) # 53 do not converge

# remove error taxa
path_abund4 <- path_abund3 %>% filter(!path_id %in% errors)
length(unique(path_abund4$path_id)) # 116 remaining 

path_coef_int2 <- path_coef_int %>% filter(!path_id%in% errors)
length(path_coef_int2$path_id)

path_aic_int2 <- path_aic_int %>% filter(!path_id %in% errors)
length(path_aic_int2$path_id) 

# filter for taxa where full model fits best
interaction_filter <- path_aic_int2 %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(interaction_filter)) # 65 interaction model fits best for all

# filter df's for where temp model fits best 
path_abund5 <- path_abund4 %>% filter(path_id %in% interaction_filter)
length(unique(path_abund5$path_id))  

path_coef_int3 <- path_coef_int2 %>% filter(path_id %in% interaction_filter)
length(path_coef_int3$path_id)

path_aic_int2 <- path_aic_int2 %>% filter(path_id %in% interaction_filter)
length(path_aic_int2$path_id)

```

### iii. summary table

```{r}


# final table of model results for all tested taxa
levels <- path_abund5 %>% select(path_id, path_name) %>%
  distinct()

weights <- path_aic_int2 %>% filter(model == "m_full") %>% select(path_id, weight)

path_int_summary <- path_coef_int3 %>%
  left_join(levels, by = "path_id") %>%
  left_join(weights, by = "path_id")


write.csv(path_int_summary, "figures/supp/path_interaction_DAA_results.csv", row.names = FALSE)

```

### iv get model predictions

Only using two neopterin bins because we don't have a wide range of neopterin values for the shotgun data as we did for the taxonomy data

```{r}

# Create neo bins
path_abund5$neo_bin <- cut(path_abund5$neo_s, breaks = 2, labels = c("Bin1", "Bin2"), include.lowest = TRUE)
bin_limits <- levels(cut(path_abund5$neo_s, breaks = 2, include.lowest = TRUE)) # look at bin range
bin_avg <- tapply(path_abund5$neo_s, path_abund5$neo_bin, mean, na.rm = TRUE) # use for co plot

# Create temp seq
temp_seq <- seq(min(path_abund5$temp_s, na.rm = TRUE), max(path_abund5$temp_s, na.rm = TRUE), length.out = 100)

# list
neo_preds <- list()

# loop
for (i in unique(path_abund5$path_id)) {
  
  data <- droplevels(subset(path_abund5, path_id == i))
  
  m <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s * neo_s,
            data = data)
  
  coef <- round(summary(m)$coefficients["temp_s:neo_s", 1], 6)
  
  # neo_s values and bins
  neo_values <- c(-0.04380621, 3.06643502)
  bins <- paste0("neopterin_avg", 1:2)
  
  # Generate predictions for each neo_s value
  preds <- lapply(seq_along(neo_values), function(j) {
    neo_s <- neo_values[j]
    bin <- bins[j]
    
    # preds dataset
    pred <- expand.grid(
      temp_s = temp_seq,
      neo_s = neo_s,
      age_s = mean(data$age_s, na.rm = TRUE),
      rain_s = mean(data$rain_s, na.rm = TRUE),
      fruit_s = mean(data$fruit_s, na.rm = TRUE),
      group = "Tenori",
      sex = "M"
    )
    
    # get model predictions
    preds_model <- predict(m, newdata = pred, se.fit = TRUE)
    
    # store preds and CI's 
    df <- data.frame(
      temp_s = pred$temp_s,  
      neo_s = neo_s,
      fit = preds_model$fit,
      lower = preds_model$fit - 1.96 * preds_model$se.fit,
      upper = preds_model$fit + 1.96 * preds_model$se.fit,
      bin = bin,
      coef = coef
    )
    
    return(df)
  })
  
  # combine predictions for all bins 
  preds_df <- do.call(rbind, preds)
  preds_df$path_id <- i 
  
  # store preds
  neo_preds[[i]] <- preds_df
}

# Combine all predictions into a single dataframe
path_neo_preds <- do.call(rbind, neo_preds)
row.names(path_neo_preds) <- NULL # clean up row names

path_name <- path_abund5 %>% select(path_id, path_name) %>%
  distinct()

path_neo_preds <- path_neo_preds %>% left_join(path_name, by = "path_id")

```

### v. plot by pathway

pdf for supplement with plotted model predictions for each pathway plus confidence intervals 

```{r}

pdf("figures/supp/pathway_interaction.pdf", width = 8, height = 11)  

plot_list <- list()

# loop
for (i in unique(path_neo_preds$path_name)) {
  
  data <- subset(path_neo_preds, path_name == i)

  p <- ggplot(data, aes(x = temp_s, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "grey") +
    geom_line() +
    facet_wrap(~ bin, nrow = 1) +  
    theme_minimal() +
    labs(
      title = paste("predicted Abundance for", i),
      x = "temperature (scaled)",
      y = "predicted log abundance") 
  
  plot_list[[i]] <- p
}

# save plots in groups of 3 per page
for (i in seq(1, length(plot_list), by = 3)) {
  
  batch_plots <- plot_list[i:min(i + 2, length(plot_list))]
  
  grid.arrange(grobs = batch_plots, ncol = 1)
}

dev.off()

```


### vi. plot by effect direction

#### a. data prep

```{r}

# Define effect type

slope <- path_neo_preds %>%
  group_by(path_name, bin) %>%
  do({m <- lm(fit ~ temp_s, data = .)
    slope <- coef(m)["temp_s"]
    data.frame(slope = slope)}) %>%
  ungroup() %>%
  pivot_wider(names_from = bin, values_from = slope,
                     names_prefix = "slope_") %>%
  mutate(effect_type = case_when(
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 > 0 & slope_neopterin_avg2 > slope_neopterin_avg1 ~ "positive_amplifying",
      slope_neopterin_avg1 < 0 & slope_neopterin_avg2 < 0 & slope_neopterin_avg1 > slope_neopterin_avg2 ~ "negative_amplifying",
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 > 0 & slope_neopterin_avg2 < slope_neopterin_avg1 ~ "positive_dampening",
      slope_neopterin_avg1 > 0 & slope_neopterin_avg2 < 0 ~ "negative_reverse",
      slope_neopterin_avg1 < 0 & slope_neopterin_avg2 > 0 ~ "positive_reverse",
      TRUE ~ "other")) %>%
  select(path_name, effect_type)
      
      
# Prep data

mean_neo <- mean(path_abund5$neo)
sd_neo <- sd(path_abund5$neo) 

neo_reverse <- function(x) {
  return(round(x * 2 * sd_neo + mean_neo))  # reverse scale in figure axes
}

bin_limits <- levels(cut(path_abund5$neo_s, breaks = 2, include.lowest = TRUE)) # look at bin range
neo_reverse(1.24)

path_neo_preds2 <- path_neo_preds %>% mutate(bin = case_when(bin == "neopterin_avg1" ~ "neopterin: 107-2812 (nmol/L)",
                                                   bin == "neopterin_avg2" ~ "neopterin: 2812-5524 (nmol/L)")) %>%
  left_join(slope, by = "path_name") %>%
  rename(pathway = path_name) %>%
  mutate(pathway = str_remove_all(pathway, " \\(.*?\\)"),
         pathway = str_replace_all(pathway, "&beta;", "beta"),
         pathway = str_replace_all(pathway, "&alpha;", "alpha"),
         pathway = str_replace_all(pathway, "&delta;", "delta")) 
  
  
  
# Create df's for each effect                               
                
positive_amplifying <- path_neo_preds2 %>% filter(effect_type == "positive_amplifying")
negative_amplifying <- path_neo_preds2 %>% filter(effect_type == "negative_amplifying")
positive_dampening <- path_neo_preds2 %>% filter(effect_type == "positive_dampening")
negative_reverse <- path_neo_preds2 %>% filter(effect_type == "negative_reverse")
positive_reverse <- path_neo_preds2 %>% filter(effect_type == "positive_reverse")

# info for each interaction type
length(unique(positive_amplifying$path_id)) # 10
length(unique(negative_amplifying$path_id)) # 7
length(unique(positive_dampening$path_id)) # 3
length(unique(negative_reverse$path_id)) # 18
length(unique(positive_reverse$path_id)) # 27

info_positive_amplifying <- positive_amplifying %>%
  select(path_id, pathway, effect_type) %>%
  distinct()

info_negative_amplifying <- negative_amplifying %>%
  select(path_id, pathway, effect_type) %>%
  distinct()

info_positive_dampening <- positive_dampening %>%
  select(path_id, pathway, effect_type) %>%
  distinct()

info_negative_reverse <- negative_reverse %>%
  select(path_id, pathway, effect_type) %>%
  distinct()

info_positive_reverse <- positive_reverse %>%
  select(path_id, pathway, effect_type) %>%
  distinct()

# bin values
unique_bins <- unique(path_neo_preds2$bin)

# to rescale x axis
mean_temp <- mean(path_abund5$temp)
sd_temp <- sd(path_abund5$temp) 

```

#### b. plot

```{r}

################### ngeaitve_amplifying ######################

y_min <- min(negative_amplifying$fit)
y_max <- max(negative_amplifying$fit)

negamp_path <- unique(negative_amplifying$pathway) # unique paths
negamp_colors <- setNames(generate_shades("#fb8072", length(negamp_path)), negamp_path) # color assignment

negative_amp_list <- list()

# loop
for (i in unique_bins) {
  
  negative_amp_filtered <- negative_amplifying[negative_amplifying$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(negative_amp_filtered, aes(x = temp_s, y = fit, color = pathway)) +
    geom_line() +
    labs(y = "", x = "") +
     scale_y_continuous(limits = c(y_min, y_max)) + 
    scale_x_continuous(breaks = seq(min(path_neo_preds2$temp_s), max(path_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) + 
    scale_color_manual(values = negamp_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  negative_amp_list[[as.character(i)]] <- bin_plot
  
}


# legend
negamp_path <- paste(c("<b>Included pathways:</b>", unique(negative_amplifying$pathway)), collapse = "<br>")

negamp_labels <- richtext_grob(negamp_path,
  x = 0, y = 1, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 8),
  box_gp = gpar(col = NA))  # no box around the text

negative_amp_plot <- ggarrange(do.call(ggarrange, c(negative_amp_list, nrow = 1)),
  as_ggplot(negamp_labels), 
  ncol = 2,
  widths = c(3, 2.2))

negative_amp_plot

################## negative_reverse ###########################

y_min <- min(negative_reverse$fit)
y_max <- max(negative_reverse$fit)

negrev_path <- unique(negative_reverse$pathway) 
negrev_colors <- setNames(generate_shades("#fdae61", length(negrev_path)), negrev_path) 

negative_reverse_list <- list()

# loop
for (i in unique_bins) {
  
  negative_reverse_filtered <- negative_reverse[negative_reverse$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(negative_reverse_filtered, aes(x = temp_s, y = fit, color = pathway)) +
    geom_line() +
    labs(y = "", x = "") +
    scale_y_continuous(limits = c(y_min, y_max),
                       labels = function(x) sprintf("     %s", x)) +
    scale_x_continuous(breaks = seq(min(path_neo_preds2$temp_s), max(path_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) + 
    scale_color_manual(values = negrev_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  negative_reverse_list[[as.character(i)]] <- bin_plot
  
}

# legend
negrev_path <- paste(c("<b>Included pathways:</b>", unique(negative_reverse$pathway)), collapse = "<br>")

negrev_labels <- richtext_grob(negrev_path,
  x = 0, y = 1, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 8),
  box_gp = gpar(col = NA)) 

negative_reverse_plot <- ggarrange(do.call(ggarrange, c(negative_reverse_list, nrow = 1)),
  as_ggplot(negrev_labels), 
  ncol = 2,
  widths = c(3, 2.2))

negative_reverse_plot

################## positive_dampening ###########################

y_min <- min(positive_dampening$fit)
y_max <- max(positive_dampening$fit)

posdamp_path <- unique(positive_dampening$pathway) 
posdamp_colors <- setNames(generate_shades("#bc80bd", length(posdamp_path)), posdamp_path)

positive_dampening_list <- list()

# loop
for (i in unique_bins) {
  
  positive_dampening_filtered <- positive_dampening[positive_dampening$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(positive_dampening_filtered, aes(x = temp_s, y = fit, color = pathway)) +
    geom_line() +
    labs(y = "", x = "") +
    scale_y_continuous(limits = c(y_min, y_max),
                       labels = function(x) sprintf("  %s", x)) +
    scale_x_continuous(breaks = seq(min(path_neo_preds2$temp_s), max(path_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) + 
    scale_color_manual(values = posdamp_colors) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  positive_dampening_list[[as.character(i)]] <- bin_plot
  
}

# legend
posdamp_path <- paste(c("<b>Included pathways:</b>", unique(positive_dampening$pathway)), collapse = "<br>")

posdamp_labels <- richtext_grob(posdamp_path,
  x = 0, y = 0.4, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 8),
  box_gp = gpar(col = NA)) 

positive_dampening_plot <- ggarrange(do.call(ggarrange, c(positive_dampening_list, nrow = 1)),
  as_ggplot(posdamp_labels), 
  ncol = 2,
  widths = c(3, 2.2))

positive_dampening_plot

################## positive_amplifying ###########################
y_min <- min(positive_amplifying$fit)
y_max <- max(positive_amplifying$fit)

posamp_path <- unique(positive_amplifying$pathway) 
posamp_colors <- setNames(generate_shades("#80b1d3", length(posamp_path)), posamp_path)

positive_amp_list <- list()

# loop
for (i in unique_bins) {
  
  pos_amp_filtered <- positive_amplifying[positive_amplifying$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(pos_amp_filtered, aes(x = temp_s, y = fit, color = pathway)) +
    geom_line() +
    labs(y = "", x = "", title = paste(i)) +
    scale_y_continuous(limits = c(y_min, y_max),
                       labels = function(x) sprintf("  %s", x)) +
    scale_x_continuous(breaks = seq(min(path_neo_preds2$temp_s), max(path_neo_preds2$temp_s), length.out = 6),
                       labels = rep("", 6)) + 
    scale_color_manual(values = posamp_colors) + 
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text.x = element_blank(),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  positive_amp_list[[as.character(i)]] <- bin_plot
  
}

#legend
posamp_path <- paste(c("<b>Included pathways:</b>", unique(positive_amplifying$pathway)), collapse = "<br>")

posamp_labels <- richtext_grob(posamp_path,
  x = 0, y = .85, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 8),
  box_gp = gpar(col = NA)) 

positive_amp_plot <- ggarrange(do.call(ggarrange, c(positive_amp_list, nrow = 1)),
  as_ggplot(posamp_labels), 
  ncol = 2,
  widths = c(3, 2.2))

positive_amp_plot

################## positive_reverse ###########################

y_min <- min(positive_reverse$fit)
y_max <- max(positive_reverse$fit)

posrev_path <- unique(positive_reverse$pathway) 
posrev_colors <- setNames(generate_shades("#66bd63", length(posrev_path)), posrev_path)

positive_reverse_list <- list()

# loop
for (i in unique_bins) {
  
  positive_reverse_filtered <- positive_reverse[positive_reverse$bin == i, ]
  
  y_tick_labels <- if (i == "neopterin: 107-2812 (nmol/L)") {
    element_text(size = 10, color = "black")
  } else {
    element_blank()
  }
  
  bin_plot <- ggplot(positive_reverse_filtered, aes(x = temp_s, y = fit, color = pathway)) +
    geom_line() +
    labs(y = "", x = "") +
    scale_y_continuous(limits = c(y_min, y_max)) +
    scale_x_continuous(breaks = seq(min(path_neo_preds2$temp_s), max(path_neo_preds2$temp_s), length.out = 6),
                       labels = inverse_scale) + 
    scale_color_manual(values = posrev_colors) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks.x = element_line(size = 0.5),
      axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
      axis.text.y = y_tick_labels,
      axis.ticks.y = element_line(size = 0.5),
      axis.text = element_text(size = 10, color = "black"),
      legend.position = "none",
      plot.margin = margin(5, 5, 5, 5))
  
  positive_reverse_list[[as.character(i)]] <- bin_plot
  
}

#legend
posrev_path <- paste(c("<b>Included pathways:</b>", unique(positive_reverse$pathway)), collapse = "<br>")

posrev_labels <- richtext_grob(posrev_path,
  x = 0, y = 1, hjust = 0, vjust = 1,
  gp = gpar(fontsize = 8),
  box_gp = gpar(col = NA)) 

positive_reverse_plot <- ggarrange(do.call(ggarrange, c(positive_reverse_list, nrow = 1)),
  as_ggplot(posrev_labels), 
  ncol = 2,
  widths = c(3, 2.2))

positive_reverse_plot


######## combine ##########

pos_slope <- ggarrange(positive_amp_plot, positive_reverse_plot, 
                           nrow = 2, ncol = 1, align = "v",
                           labels = c("A", "B"))

pos_slope <- annotate_figure(pos_slope, left = text_grob("predicted\nclr-transformed\nabundance", size = 12)) 


neg_slope <- ggarrange(negative_amp_plot, negative_reverse_plot, positive_dampening_plot,
                       nrow = 3, ncol = 1, align = "v", 
                       labels = c("C", "D", "E"))

neg_slope <- annotate_figure(neg_slope, left = text_grob("predicted\nclr-transformed\nabundance", size = 12)) 

neg_slope <- annotate_figure(neg_slope, bottom = text_grob("temperature (°C)", size = 14, x = 0.42))







spacer <- ggplot() + theme_void()

combined_plot <- ggarrange(pos_slope, spacer, neg_slope, 
                           nrow = 3, ncol = 1, 
                           heights = c(2, 1.3, 3))



ggsave("figures/fig7.pdf", combined_plot, width =11, height = 12)


```
