---
title: "Untitled"
author: "Lucore"
date: "2026-04-14"
output: html_document
---

To do: 

finish citations for taxa ratios
check ontent of pathway ratios
figure out modeling for redox pathways 



Notes: This is the new analysis that does not include the neopterin data. We are going to run only the temp analysis with all of the 16s and shotgun samples to increase sample size and cut down on the amount of analysis in the temp paper. 

We are going to use the log-constrast model approach instead of modeling the individual features. If th log-constrat appraoch works for both the 16s and shotgun data, will use those results for the temperature analysis instead of the tree and individual feature coefficient plots. 

# 1. Load environment

```{r}

# packages
library(tidyverse)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(ggnewscale)
library(zCompositions)
library(ggtree)
library(tidytree)
library(ape)
library(jsonlite)
library(file2meco)
library(phyloseq)
library(flextable)
library(gridtext)
library(grid)

#physeq objects
tax_physeq <- readRDS("phyloseq_objects/2024-12-18_phyloseq_16s.rds")
path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

# functions
source("functions/my_scale.R")
source("functions/inverse_scale.R")
source("functions/kegg_parse_json.R")
source("functions/make_coef_plot.R")
source("functions/model_weight_plot.R")
source("functions/model_weight_plot.R")
source("functions/generate_shades.R")

# conflicts
conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)
conflicts_prefer(cowplot::get_legend)


```

# 2. Data set up

## A. Taxonomy (16s seqs)

### i. Metadata filtering

using 699 samples that have feeding

```{r}

tax_m <- data.frame(sample_data(tax_physeq)) %>% 
  rename(temp = max_t_days15,
         rain = rain_days4) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals)

tax_m_filtered <- tax_m[!is.na(tax_m$fruit), ] # remove samples with no diet data

tax_physeq2 <- prune_samples(rownames(tax_m_filtered), tax_physeq) # filter physeq object

tax_physeq2 <- prune_taxa(taxa_sums(tax_physeq2) > 0, tax_physeq2) # remove ASV's with zero abundance

tax_physeq2

```

### ii. Prevalence filtering 

Identifies taxa that are not present in at least 0.5% of the total samples (at least two samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth.  

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(tax_physeq2),
                MARGIN = ifelse(taxa_are_rows(tax_physeq2), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(tax_physeq2),
                      tax_table(tax_physeq2))

# define prevalence threshold
prevalenceThreshold = 0.005 * nsamples(tax_physeq2)  
prevalenceThreshold # must appear in at least 3.5 samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

ggsave("figures/abridged_temp/supp/supp_fig4.pdf", prev_fig, width = 6, height = 4)

# summary stats
asv_onesample <- nrow(prevdf[prevdf$Prevalence == 1,]) # number of ASV's present only in one sample
total_asv <- nrow(prevdf)
percent_onesample <- (asv_onesample/total_asv) * 100 # percent ASV's in only one sample

asv_nophyla <- nrow(prevdf[is.na(prevdf$Phylum),]) # number of ASV's with no assigned phylum 
percent_nophyla <- (asv_nophyla/total_asv) * 100 # percent number of ASV's with no phylum

## execute prevalence filter ##
tax_phylo_filtered <- prune_taxa((prev0 > prevalenceThreshold), tax_physeq2) 
tax_phylo_filtered

## remove taxa with unnasigned phyla ## 
tax_phylo_filtered2 <- subset_taxa(tax_phylo_filtered, !is.na(Phylum)) # remove 67 taxa 
tax_phylo_filtered2

# calc reads per sample
reads_filtered <- as.data.frame(sample_sums(tax_phylo_filtered2)) %>% 
  rownames_to_column(var = "sample") %>%
  rename(filtered_reads = 'sample_sums(tax_phylo_filtered2)')
  
# add reads per sample to metadata
tax_m_filtered2 <- tax_m_filtered %>%
  left_join(reads_filtered, by = "sample") 

# % assignment for remaining taxonomy levels

asign_class <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Class"])))/1159)*100)

asign_order <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Order"])))/1159)*100)

asign_family <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Family"])))/1159)*100)

asign_genus <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Genus"])))/1159)*100)

######## evaluate which phlya have greater than 1% rel abund ############

phylum_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Phylum"))
  
phylum_otu_table <- as.data.frame(otu_table(phylum_aggregated))
phylum_tax_table <- as.data.frame(tax_table(phylum_aggregated))

phylum_otu_ids <- rownames(phylum_otu_table)
phylum_tax_ids <- rownames(phylum_tax_table)

phylum_names <- phylum_tax_table[phylum_otu_ids, "Phylum"]

rownames(phylum_otu_table) <- phylum_names

# across samples
phylum_otu_table2 <- phylum_otu_table %>%
  rownames_to_column("Phylum") %>%
  pivot_longer(-Phylum, names_to = "sample", values_to = "Count") %>%
  left_join(reads_filtered, by = "sample") %>%
  group_by(sample) %>%
  mutate(RelativeAbundance = Count / sum(Count),
          Phylum_grouped = case_when(
           str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
           RelativeAbundance > 0.01 ~ Phylum,  
           TRUE ~ "Less than 1% rel abund"  
         )) %>%
  ungroup()

df_total <- phylum_otu_table2 %>%
  mutate(Phylum_grouped = case_when(
    str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
    TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarise(TotalCount = sum(Count)) %>%
  ungroup() %>%
  mutate(RelativeAbundance = TotalCount / sum(TotalCount)) %>%
  mutate(Phylum_grouped = case_when(RelativeAbundance > 0.01 ~ Phylum_grouped, 
                                    TRUE ~ "Less than 1% rel abund")) %>%
  group_by(Phylum_grouped) %>%
  summarise(RA = sum(RelativeAbundance)) %>%
  ungroup() %>%
  mutate(RA = 100*RA,
         Phylum_grouped = fct_reorder(Phylum_grouped, RA))

ggplot(df_total, aes(x = Phylum_grouped, y = RA, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Relative Abundance (%)", fill = "Phylum") +
  theme_minimal() +
  coord_flip() 

```

###  iii. Generate alpha diversity metrics

```{r}

shannon <- estimate_richness(tax_phylo_filtered2, measures = "Shannon") # calc shannon
shannon <- shannon %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

hist(shannon$shannon) # normal dist

observed <- estimate_richness(tax_phylo_filtered2, measures = "Observed") # calc observed
observed <- observed %>%
  rownames_to_column(var = "sample") %>%
  rename(observed = Observed) %>%
  mutate( sample = str_replace_all(sample, "^X", ""),  
          sample = str_replace_all(sample, "\\.", "-"))

hist(observed$observed) # gamma dist


tax_m_filtered3 <- tax_m_filtered2 %>% # add to metadata
  left_join(shannon, by = "sample") %>%
  left_join(observed, by = "sample") 

```

### iv. CLR transformationand PCA

The rows need to be samples when doing the PCA because results show how samples cluster based on ASV's instead of the samples themselves

Use the CLR transformation and PCA analysis because of Gloor et al., 2017, Baniel publications, Gloor and Ried 2016, and this tutorial https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot reccomend that normalizing for sequencing depth is best done with a CLR transformation (as opposed to rarefaction) and that sample distances are best estimated using Aitchisons distances as opposed to Bray Curtis or Unifrac because they do not reflect the compositional nature of the data. Aitchison distance is the euclidean distances between the samples after the CLR transformation.

*****Retain PCA axes 1-3 because they explain >4% of the variance***

```{r}

######### transformation and PCA #########

otu_table <- as.matrix(t(otu_table(tax_phylo_filtered)))
otu_table[1:5,1:5] # rows are samples 

set.seed(123)
czm <- cmultRepl(otu_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
czm[1:5,1:5] # rows are samples 

clr <- t(apply(czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
clr[1:5,1:5] # rows are samples 

pca <- prcomp(clr) # calc pca
plot(pca$x[,1],pca$x[,2]) # plot 

d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="") # PC1 12.7%
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 7.4%
PC3 <- paste("PC3 ","(",round(sum(pca$sdev[3]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 4.7%
PC4 <- paste("PC4 ","(",round(sum(pca$sdev[4]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 3.8%

######### make scree plot #########

eigenvalues <- pca$sdev^2  # square the standard deviations to get eigenvalues
percent_explained <- 100 * eigenvalues / sum(eigenvalues) 

# scree plot
pe_plot <- tibble(pe = percent_explained, axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  geom_point() +
  labs(x = "Principal component axis", y = "Percent variance explained") +
  scale_y_continuous(breaks = seq(0, max(percent_explained), by = 2.5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        axis.ticks.y = element_line(color = "black", linewidth = 0.3),
        axis.ticks.x = element_line(color = "black", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/abridged_temp/supp/supp_fig6.pdf", pe_plot, width = 6, height = 6)

######### Add PCA values to metadata #########

df<-data.frame(pca$x)
df$sample<-rownames(df)
df_small <- df[, c(1:3,700)]

# add to metadata
tax_m_filtered4 <- tax_m_filtered3 %>%
  left_join(df_small, by = "sample") 

```

### v. prep data for modeling

scale variables and make categorical variables factors. 

in previous analysis, made all palmas samples tenorie because there were only 6 palmas samples, now we have 86 so lets try keeping the groups separate 

```{r}

# scale variables
tax_model <- tax_m_filtered4 %>%
  mutate(across(c(age, temp, rain, fruit, residuals), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID)) 

  #mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           #TRUE ~ group)),
          

```

### vi. Suppelemenatary food plots

show that temp and fruit are correlated, to be able to evaluate temp, need to control for fruit.  

```{r}

# fruit ~ temp
fruit_plot <- ggplot(tax_model, aes(x = temp, y = fruit)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") +
  labs(y = "Count of fruit eaten", x = "Temperature (°C)") +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/abridged_temp/supp/supp_fig5.pdf", fruit_plot, width = 6, height = 6)

```

## B. Pathways (shotgun seqs)

### i. Metadata filtering

using 87 samples that have feeding and metadata

```{r}

path_m <- data.frame(sample_data(path_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals, reads)

path_m_filtered <- path_m[!is.na(path_m$fruit), ] # remove samples with no diet data, drops 8 samples

# filter phyloseq objects to remove samples with no diet data 
path_physeq2 <- prune_samples(path_m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

### ii. Prevalence filtering

Remove unmapped and unintegrated values. Unmapped values represent the total abundance of reads that failed to align to any known gene sequences and unintegrated values represent genes that did not belong to a known pathway.

```{r}

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

### iii. Generate alpha diversity metrics

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

Ignore the warning, `estimate_richness()` prefers count data but it gives you exactly the same values as if you calculated it with vegan. 

```{r}

shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(shannon, by = "sample")

hist(path_alpha$shannon) 
hist(path_alpha$observed)

path_m1 <- path_m_filtered %>% # add to metadata
  left_join(path_alpha, by = "sample")

```

### iv. CLR transformation and PCA 

```{r}

######### transformation and PCA #########

path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 


path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 22.6%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 11.8%

######### Add PCA values to metadata #########

path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:3,88))

path_m2 <- path_m1 %>%
  left_join(path_df_small, by = "sample")

```

### v. Prep data to model

scale variables and make categorical variables factors

```{r}

path_model <- path_m2 %>%
  mutate(across(c(age, temp, rain, fruit, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```

# 3. Temperature analysis

## A. Community-level modeling

### i. Taxonomy

Compare m_main with base-level predictors group, age, sex, fruit, diet diversity, rain, and ID as predictors, to m_full which adds a temperature predictor. Both model include an offset of reads to control for seq depth. Use lmer() for shannon and PCA axes because they are (generally) normally distributed. Use glmer() for observed because it needs a gamma distribution, data is left skewed. 

When you run the loop there is a singular fit error, this is from PC2, but the output looks okay and it is only for the m_main model not the m_full model. We are using the coefficient from m_full. 

```{r}

## test code ##
# to make sure the loop indexes for the temp effect #
test <- lmer(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)), data = tax_model)
summary(test)
std <- round(summary(test)$coeff[9,2], 4)
coef <- round(summary(test)$coeff[9,1], 4)

####### histogranms ########

hist(tax_model$shannon) 
hist(tax_model$observed) 
hist(tax_model$PC1)
hist(tax_model$PC2)
hist(tax_model$PC3)

#### loop prep ####

com_coef <- list() 
com_aic <- list()

com_features <- colnames(tax_model)[c(14, 16:18)] # list of features to model

############ loop  #################
for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads))"))
      
      m_main <- lmer(formula_main, data = tax_model)
      m_full <- lmer(formula_full, data = tax_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[9,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[9,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
tax_com_aic <- do.call(rbind, com_aic)
tax_com_coef <- do.call(rbind, com_coef)


#### model observed ######

o_main <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

o_full <- glmer(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(filtered_reads)),
                family = Gamma(link = "log"),
                data = tax_model)

AICc <- as.data.frame(AICctab(o_full, o_main, weights = TRUE))
AICc$model <- rownames(AICc)

coef <- round(summary(o_full)$coeff[9,1], 4)
std <- round(summary(o_full)$coeff[9,2], 4)
ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
df <- cbind(ci_95, ci_50) 

aic_df <- data.frame(
  metric = "observed",
  model = AICc$model,
  dAICc = round(AICc$dAICc,3),
  weight = round(AICc$weight, 3))
    
results_df <- data.frame(
  metric = "observed",
  coef = round(df$taxa_coef, 3),
  std = round(df$taxa_std, 3),
  low_ci_95 = round(df$Lower_95, 3),
  high_ci_95 = round(df$Upper_95, 3),
  low_ci_50 = round(df$Lower_50, 3),
  high_ci_50 = round(df$Upper_50, 3))

tax_com_coef <- rbind(tax_com_coef, results_df)
tax_com_aic <- rbind(tax_com_aic, aic_df)

```

### ii. Pathways

normally distributed enough to model with Gaussian (PC1 and shannon are most concerning)

```{r}

####### histogranms ########

hist(path_model$shannon) 
hist(path_model$observed) 
hist(path_model$PC1)
hist(path_model$PC2)
hist(path_model$PC3)

############ loop  #################

com_coef <- list() 
com_aic <- list()

com_features <- colnames(path_model)[c(14:18)] # list of features to model

for (i in com_features) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + (1 | ID) + offset(log(reads))"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + (1 | ID) + offset(log(reads))"))
      
      m_main <- lmer(formula_main, data = path_model)
      m_full <- lmer(formula_full, data = path_model)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    com_aic[[i]] <- aic_df
    com_coef[[i]] <- results_df

}

# combine results 
path_com_aic <- do.call(rbind, com_aic)
path_com_coef <- do.call(rbind, com_coef)

```

### iii. Plot model results

#### a. data prep

```{r}

## taxa data prep ##

plot_com_coef <- tax_com_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_com_model <- tax_com_aic %>%
  select(metric, model, weight) %>%
  mutate(model = case_when(model == "o_full" ~ "m_full",
                           model == "o_main" ~ "m_main",
                           TRUE ~ model)) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_com_coef$order))) %>%
  arrange(order)

## path data prep ##

plot_path_observed <- path_com_coef %>%
  filter(metric == "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_coef <- path_com_coef %>%
  filter(metric != "observed") %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_model_obs <- path_com_aic %>%
  filter(metric == "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%
   mutate(order = factor("observed"))

path_order <- c("PC2", "shannon", "PC3", "PC1")

plot_path_model <- path_com_aic %>%
  filter(metric != "observed") %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%  
  mutate(order = factor(metric, levels = levels(plot_path_coef$order))) %>%
  arrange(order)

levels <- levels(plot_com_model$order)
which(levels == "PC3") # height for plot label


```

#### b. combined plot

```{r}

p1 <- make_coef_plot(plot_com_coef) +
  ylim(-10, 10) +
  ggtitle("A. Taxonomy temperature coefficients") +
  ylab("") +
  theme(axis.text = element_text(size = 12, color = "black"),
      text = element_text(size = 10, color = "black", family = "sans"),
      axis.title = element_text(size = 12, color = "black"),
      plot.title = element_text(size = 14, face = "bold"))

p2 <- make_coef_plot(plot_path_coef) +
  ylim(-24, 24) +
  ggtitle("C. Metabolic pathway temperature coefficients") +
  ylab("") +
  theme(axis.text = element_text(size = 12, color = "black"),
      text = element_text(size = 10, color = "black", family = "sans"),
      axis.title = element_text(size = 12, color = "black"),
      plot.title = element_text(size = 14, face = "bold"))

p3 <- make_coef_plot(plot_path_observed) + 
  ylim(-100, 100) +
  ylab("Effect size") +
  theme(axis.text = element_text(size = 12, color = "black"),
      text = element_text(size = 10, color = "black", family = "sans"),
      axis.title = element_text(size = 12, color = "black"),
      plot.title = element_text(size = 14, face = "bold"))

p4 <- model_weight_plot(plot_com_model) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 5.45, label = "temp model", color = "#f4b942", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 5.45, label = "base model", color = "#3f7743", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p5 <- model_weight_plot(plot_path_model) +
  ggtitle("D. Metabolic pathway model weights") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p6 <- model_weight_plot(plot_path_model_obs) +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
  labels = c("100", "50", "0", "50", "100"),
  limits = c(-1, 1)) +
  xlab("model weight")

com_temp_plot <- ggarrange(p1, p4, p2, p5, p3, p6, ncol = 2, nrow = 3, 
          widths = c(1, 1),
          align = "v",
          heights = c(1, 1, 0.4))

ggsave("figures/abridged_temp/newfig1.pdf", com_temp_plot, width = 12, height = 8)

```

### iv. Plot alpha diversity results

plot model predictions for alpha diversity over raw data

```{r}

# set up predictions df
temp_seq <- seq(from = min(tax_model$temp_s, na.rm = TRUE), 
                to = max(tax_model$temp_s, na.rm = TRUE), length.out = 100)

pred_data <- data.frame(
  group = unique(tax_model$group)[1], 
  age_s = mean(tax_model$age_s, na.rm = TRUE),
  sex = unique(tax_model$sex)[1], 
  fruit_s = mean(tax_model$fruit_s, na.rm = TRUE),
  residuals_s = mean(tax_model$residuals_s, na.rm = TRUE),
  rain_s = mean(tax_model$rain_s, na.rm = TRUE),
  temp_s = temp_seq,
  filtered_reads = mean(tax_model$filtered_reads, na.rm = TRUE))

########## shannon ##########
s_temp <- lm(shannon ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s + offset(log(filtered_reads)), 
               data = tax_model)
s_preds <- predict(s_temp,  newdata = pred_data, interval = "confidence", level = 0.95 ) %>%
  data.frame() %>%
  rename(s_preds = fit,
         s_lower = lwr,
         s_upper = upr)
pred_data <- cbind(pred_data, s_preds) 

# to rescales x axis
mean_temp <- mean(tax_model$temp)
sd_temp <- sd(tax_model$temp) 

s_plot <- ggplot(pred_data, aes(x = temp_s, y = s_preds)) +
  geom_point(data = tax_model, aes(x = temp_s, y = shannon),alpha = 0.5, size = 4, color = "#fee08b") +
  geom_smooth(method = "lm", lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = s_lower, ymax = s_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  ggtitle("A. Shannon diversity") +
  labs(y = "", x = "") +
  scale_x_continuous(
    breaks = seq(min(tax_model$temp_s), max(tax_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        text = element_text(size = 10, color = "black", family = "sans"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

########## observed ##########
o_temp <- glm(observed ~ group + age_s + sex + fruit_s + residuals_s + rain_s + temp_s +
                offset(log(filtered_reads)), 
                family = Gamma(link = "log"),
                data = tax_model)

o_preds <- data.frame(predict(o_temp,  newdata = pred_data, type = "response", se.fit = TRUE)) %>%
  select(-residual.scale) %>%
  rename(o_preds = fit,
         o_se = se.fit)

pred_data <- cbind(pred_data, o_preds) %>%
  mutate(o_lower = o_preds - 0.97 * o_se, # calc CI's 
         o_upper = o_preds + 0.97 * o_se)

o_plot <- ggplot(pred_data, aes(x = temp_s, y = o_preds)) +
  geom_point(data = tax_model, aes(x = temp_s, y = observed),alpha = 0.5, size = 4, color = "#fee08b") +
  geom_line(lwd = 1, color = "black") +
  geom_ribbon(aes(ymin = o_lower, ymax = o_upper), 
              fill = "grey60", alpha = 0.5) +
  theme_minimal() +
  labs(y = "", x = "") +
  ggtitle("B. Observed ASV's") +
  scale_x_continuous(
    breaks = seq(min(tax_model$temp_s), max(tax_model$temp_s), length.out = 6), 
    labels = inverse_scale) +  # rescale axis labels
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        text = element_text(size = 10, color = "black", family = "sans"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

# add plots together  

alpha_plots <- ggarrange(s_plot, o_plot, ncol = 2, nrow = 1)

alpha_plots <- annotate_figure(alpha_plots, bottom = text_grob("temperature (°C)", size = 14))

ggsave("figures/abridged_temp/supp/supp_fig2.pdf", alpha_plots, width =8, height = 4)

```

## B. Taxa analysis

### i. Preprocessing

clean up taxa names in phyloseq object, use `tax_glom()` to combine taxa at the genus level and at the family level (i.e. sum abundances of the ASV's that are grouped together into genus or family). Change row names from ASV ID to taxa name. 

```{r}

# get rid of all the weird, unnassigned names and turn them all into NA's
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("Unassigned*",NA,x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("uncultured*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("unidentified*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("gut metagenome",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("metagenome",NA, x)))

# Tax glom 
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))

# extract tables
genus_otu_table <- as.data.frame(otu_table(genus_aggregated))
genus_tax_table <- as.data.frame(tax_table(genus_aggregated))
genus_metadata <- sample_data(genus_aggregated)

#change OTU rownames to ASV family names
# Ensure that the OTU IDs are the same in both tables (rows in otu_df and tax_df)
genus_otu_ids <- rownames(genus_otu_table)
genus_tax_ids <- rownames(genus_tax_table)

# Extract the taxa names from the taxonomy table
genus_names <- genus_tax_table[genus_otu_ids, "Genus"]

# Set the row names of the OTU table to the family names
rownames(genus_otu_table) <- genus_names

```

### ii. Calc rel abundance 

Calculate the relative abundance by dividing the raw abundance counts by the total reads per sample. Using the "filtered_reads" column because we are using the phyloseq object `tax_phylo_filtered2` that has already been filtered at a 0.5% threshold. 

```{r}

# extract table for DA testing
genus_da <-psmelt(genus_aggregated)

# Extract variables of interest 
genus_da1 <- genus_da %>% 
  mutate(Taxa = Genus) %>% # make genus the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, Phylum, Family, Class) %>% 
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Genus") %>%
  mutate(RA = (Abundance/filtered_reads)*100)

# n = 699
length(unique(genus_da1$sample))  

```

### iii. Filter out low abundance taxa

Calculate the average relative abundance for each taxa across all of the samples and then keep the taxa that have at lease 0.2% relative abundance across all samples. Left with 128 genera 

```{r}

#Calculate average relative abundance of each taxa across all samples
genus_a1<-sapply(unique(genus_da1$Taxa),function(x) mean(genus_da1$RA[genus_da1$Taxa==x],na.rm=TRUE))
genus_a2<-data.frame(cbind(Taxa=as.character(unique(genus_da1$Taxa)),RA=genus_a1))
genus_a2$RA<-as.numeric(as.character(genus_a2$RA))

#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

genus_da2 <-droplevels(subset(genus_da1, genus_da1$Taxa %!in% NA)) # remove asv NAs
genus_da3 <-droplevels(subset(genus_da2, genus_da2$Taxa %in% unique(genus_a2$Taxa[genus_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(genus_da2$Taxa))
length(unique(genus_da3$Taxa)) # 128 genera

```

### iv. Scale variables

scale variables for modeling, 128 total taxa for data 

not going to combine palmas and tenori

```{r}

# Scale variables
da4 <- genus_da3 %>%
   mutate(across(c(temp, rain, age, fruit), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID),
         sex = as.factor(sex)) 
  #mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           #TRUE ~ group)))

length(unique(da4$Taxa)) # 128 taxa total

#pull out genera classification to make ration assignments 
#genus <- da4 %>% select(Taxa, Phylum, Family, Class) %>% distinct()
#write.csv(genus, "data/genera.csv", row.names = FALSE)

```

### v. create ratios

We aggregated counts to the genus level used multiplicative zero replacement `cmultRepl()` to replace zeros. For each predefined, biologically informed contrast, we computed the log-contrast as the difference of mean log-abundances between the numerator and denominator genus sets, i.e., the log of the ratio of their geometric means. Because balances are log-ratios, they are controlled for sequencing depth and live in Aitchison log-ratio space. 

REMEMBER STILL NEED TO UPDATE SOURCE LIST FOR CELLS IN YELLOW ON 2026-04-14_genera_categories.csv

```{r}

# make data a long format and convert to matrix  
genera_long <- da4 %>%
  select(Taxa, Abundance, sample) %>%
  pivot_wider(names_from = Taxa, values_from = Abundance, values_fill = 0) %>%
  arrange(sample) %>% 
  column_to_rownames("sample") %>%
  as.matrix()

# add psuedo count 
genera_czm <- cmultRepl(genera_long, label=0, method="CZM", z.warning = 1, z.delete = 1)
rs <- rowSums(as.matrix(genera_czm)) # check that each sample equals 1

#### build groups ###

groups <- da4 %>% select(Taxa, Phylum) %>% distinct()

# pathbiont vs core group

core_genera <- da4 %>%
  group_by(Taxa, sample) %>%            
  summarise(present = any(RA > 0), .groups = "drop") %>% 
  group_by(Taxa) %>%
  summarise(
    prevalence = mean(present),            # proportion of samples where present
    n_samples = sum(present)
  ) %>%
  filter(prevalence >= 0.75) %>%           # ≥75% of samples
  arrange(desc(prevalence)) %>%
  mutate(Pathogenic = "core") %>%
  select(Taxa, Pathogenic) # Categorizing Anaerobiospirillum_A as pathobiont instead of core

groups2 <- groups %>% 
  left_join(core_genera, by = "Taxa")
  
taxa_groups <- read_csv("data/2026-04-14_genera_categories.csv") %>%
  select(Taxa, Oxygen_tolerance, Pathobiont)

groups3 <- groups2 %>%
  left_join(taxa_groups, by = "Taxa") %>%
  mutate(Pathogenic = case_when(Pathobiont == "pathobiont" ~ "pathobiont", 
                                TRUE ~ Pathogenic)) %>%
  select(-Pathobiont)

# phyla group 
groups4 <- groups3 %>%
  mutate(Phylum_level = case_when(Phylum == "Firmicutes_A" ~ "firmi",
                                  Phylum == "Firmicutes_C" ~ "firmi",
                                  Phylum == "Firmicutes_D" ~ "firmi",
                                  Phylum == "Proteobacteria" ~ "proteo"))

####### make ratios ########

num_oxy <- groups4 %>% filter(Oxygen_tolerance == "facultative_anaerobe") %>% pull(Taxa) %>% unique()
den_oxy <- groups4 %>% filter(Oxygen_tolerance == "obligate_anaerobe") %>% pull(Taxa) %>% unique()

num_path <- groups4 %>% filter(Pathogenic == "pathobiont") %>% pull(Taxa) %>% unique()
den_path <- groups4 %>% filter(Pathogenic == "core") %>% pull(Taxa) %>% unique()

num_phy <- groups4 %>% filter(Phylum_level == "proteo") %>% pull(Taxa) %>% unique()
den_phy <- groups4 %>% filter(Phylum_level == "firmi") %>% pull(Taxa) %>% unique()

# compute balances
source("functions/make_balance_from_matrix.R")

B_oxy <- make_balance_from_matrix(genera_czm, num_oxy, den_oxy)
B_path <- make_balance_from_matrix(genera_czm, num_path, den_path)
B_phy <- make_balance_from_matrix(genera_czm, num_phy, den_phy)

balances <- data.frame(
  sample = rownames(genera_czm),
  bal_oxygen   = as.numeric(B_oxy),
  bal_path = as.numeric(B_path),
  bal_phylum   = as.numeric(B_phy))

# add to mediation data

tax_model2 <- tax_model %>%
  left_join(balances, by = "sample")
  

```

### vi. model ratios

Was able to add in the random effect for ID because of the larger sample size

only using the phylum and oxygen ratios because the pathogen ratio doesn't work with the larger sample size 

```{r}

####### histogranms ########

hist(tax_model2$bal_phylum) 
hist(tax_model2$bal_oxygen) 

############ loop  #################

tax_coef <- list() 
tax_aic <- list()

tax_ratios <- colnames(tax_model2)[c(24,26)] # list of features to model

for (i in tax_ratios) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + rain_s + (1 | ID)"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID)"))
      
      m_main <- lmer(formula_main, data = tax_model2)
      m_full <- lmer(formula_full, data = tax_model2)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[8,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[8,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    tax_aic[[i]] <- aic_df
    tax_coef[[i]] <- results_df

}

# combine results 
tax_aic <- do.call(rbind, tax_aic)
tax_coef <- do.call(rbind, tax_coef)

```

### vii. plot tax ratios

```{r}

## data prep

plot_tax_coef <- tax_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange(desc(coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_tax_aic <- tax_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_tax_coef$order))) %>%
  arrange(order)

## plot ##

tax_coef_plot <- make_coef_plot(plot_tax_coef) +
  #ylim(-0.5, 2) +
  ggtitle("A. Taxonomic temperature coefficients") +
  ylab("effect size") +
  scale_x_discrete(
    labels = function(x) {
    x <- gsub(
      "bal_phylum",
      "frac(Proteobacteria, Firmicutes)",
      x
    )
    x <- gsub(
      "bal_oxygen",
      "frac(Oxygen~tolerant, Obligate~anaerobes)",
      x
    )
    parse(text = x)
  }
) 

tax_coef_plot

tax_aic_plot <- model_weight_plot(plot_tax_aic) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 2.25, label = "temp model", color = "#f4b942", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 2.25, label = "base model", color = "#3f7743", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))
  
tax_aic_plot

tax_plot <- ggarrange(tax_coef_plot, tax_aic_plot, ncol = 2, nrow = 1)
tax_plot

```

### viii. manual modeling

the manual models just for checking 

see here that pathogen ration doesn't work 

```{r}

# phylum
main <- lmer(bal_phylum ~ group + age_s + sex + fruit_s + rain_s + (1 | ID), data = tax_model2)
temp <- lmer(bal_phylum ~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID), data = tax_model2)
summary(temp)
AICctab(main, temp, weights = TRUE)

#oxygen
main <- lmer(bal_oxygen ~ group + age_s + sex + fruit_s + rain_s + (1 | ID), data = tax_model2)
temp <- lmer(bal_oxygen ~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID), data = tax_model2)
summary(temp)
AICctab(main, temp, weights = TRUE)

#pathogen
main <- lmer(bal_path ~ group + age_s + sex + fruit_s + rain_s + (1 | ID), data = tax_model2)
temp <- lmer(bal_path ~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID), data = tax_model2)
summary(temp)
AICctab(main, temp, weights = TRUE)

```

## C. Pathway analysis

### i. Filter by prevalence

Identifies pathways that are not present in at least 30% of the total samples (at least 27 samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth. 

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(path_filtered),
                MARGIN = ifelse(taxa_are_rows(path_filtered), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(path_filtered),
                      tax_table(path_filtered))

# define prevalence threshold
prevalenceThreshold = 0.3 * nsamples(path_filtered)  
prevalenceThreshold # must appear in at least 27 samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

## execute prevalence filter ##
path_phylo_filtered2 <- prune_taxa((prev0 > prevalenceThreshold), path_filtered) 
path_phylo_filtered2

```

### ii. Filter by rel abundance

```{r}

# extract data for DA testing

path_da <-psmelt(path_phylo_filtered2)

# Extract variables of interest 
path_da1 <- path_da %>% 
  select(OTU, Abundance, Sample, reads) %>% 
  rename(gm_sample = Sample,
         path = OTU) %>%
  mutate(RA = (Abundance/reads)*100)

# n = 87
length(unique(path_da1$gm_sample))  

#Calculate average relative abundance of each taxa across all samples
path_a1<-sapply(unique(path_da1$path),function(x) mean(path_da1$RA[path_da1$path==x],na.rm=TRUE))
path_a2<-data.frame(cbind(path=as.character(unique(path_da1$path)),RA=path_a1))
path_a2$RA<-as.numeric(as.character(path_a2$RA))

#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

path_da2 <-droplevels(subset(path_da1, path_da1$path %!in% NA)) # remove asv NAs
path_da3 <-droplevels(subset(path_da2, path_da2$path %in% unique(path_a2$path[path_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(path_da3$path))


paths <- path_da3 %>% select(path) %>% distinct()

nrow(paths) # 272

#list of paths for category assignment
#write.csv(paths, "data/paths.csv", row.names = FALSE)

```

```{r}

path_old_groups <- read_csv("data/expanded2026-04-20_path_info_with_categories FROM gut_immune_axis repo.csv")

anti_join(paths, path_old_groups, by = "path") # three in paths that are not in the old assignments 

anti_join(path_old_groups, paths, by = "path")

#PWY-6270: isoprene biosynthesis I, PWY-5464: superpathway of cytosolic glycolysis (plants), pyruvate dehydrogenase and TCA cycle, add to redox maintenance 

### since mostly the same, just add the path assignments from the old sheet

path_groups <- full_join(paths, path_old_groups, by = "path") 

```

### iii. Create log contrasts 

Nucleotide mobilization: Redox maintenance

Pathobiont/immune agonist : Antinflammatory/immune regulation

Facultative:Obligate anaerobes - this excludes respiration paths

Aerobic respiration:anaerobic metabolism - dstinct from facultative vs obligate, shows that anaerobic metabolism dominates the community at high temps, this is going in the supplement 

```{r}

# make data a long format and convert to matrix  
path_long <- path_da3 %>%
  select(path, Abundance, gm_sample) %>%
  pivot_wider(names_from = path, values_from = Abundance, values_fill = 0) %>%
  arrange(gm_sample) %>% 
  column_to_rownames("gm_sample") %>%
  as.matrix()

# add psuedo count 
path_czm <- cmultRepl(path_long, label=0, method="CZM", z.warning = 1, z.delete = 1)
rs <- rowSums(as.matrix(path_czm)) # check that each sample equals 1
  
#path_groups <- read_csv("metadata/path_info_with_categories.csv") %>%
  #mutate(path = paste(Pathway_ID, Pathway_Name, sep = ": ")) %>%
  #select(path, Redox_nucleotide, Immune_antagonist, Facul_obligate, Respiration)

####### make ratios ########

num_redox <- path_groups %>% filter(Redox_nucleotide == "nucleotide") %>% pull(path) %>% unique()
den_redox <- path_groups %>% filter(Redox_nucleotide == "redox") %>% pull(path) %>% unique()

num_imm <- path_groups %>% filter(Immune_antagonist == "antagonist") %>% pull(path) %>% unique()
den_imm <- path_groups %>% filter(Immune_antagonist == "regulatory") %>% pull(path) %>% unique()

num_facul <- path_groups %>% filter(Facul_obligate == "facultative") %>% pull(path) %>% unique()
den_facul <- path_groups %>% filter(Facul_obligate == "obligate") %>% pull(path) %>% unique()

num_resp <- path_groups %>% filter(Respiration == "aerobic") %>% pull(path) %>% unique()
den_resp <- path_groups %>% filter(Respiration == "anaerobic") %>% pull(path) %>% unique()

# compute balances

B_redox <- make_balance_from_matrix(path_czm, num_redox, den_redox)
B_imm <- make_balance_from_matrix(path_czm, num_imm, den_imm)
B_facul <- make_balance_from_matrix(path_czm, num_facul, den_facul)
B_resp <- make_balance_from_matrix(path_czm, num_resp, den_resp)

balances <- data.frame(
  sample = rownames(path_czm),
  bal_redox  = as.numeric(B_redox),
  bal_facul = as.numeric(B_facul),
  bal_imm = as.numeric(B_imm),
  bal_resp = as.numeric(B_resp))


# add to mediation data

path_model2 <- path_model %>%
  left_join(balances, by = "sample")

```

```{r}

####### histogranms ########

hist(path_model2$bal_facul) 
hist(path_model2$bal_imm) 

  ############ loop  #################

path_coef <- list() 
path_aic <- list()

path_ratios <- colnames(path_model2)[c(25,26)] # list of features to model

for (i in path_ratios) {
      
      formula_main <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + rain_s"))
      formula_full <- as.formula(paste(i, "~ group + age_s + sex + fruit_s + rain_s + temp_s"))
      
      m_main <- lm(formula_main, data = path_model2)
      m_full <- lm(formula_full, data = path_model2)
    
    # AICc
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)
    
    # extract taxa coef
    coef <- round(summary(m_full)$coeff[7,1], 4)
    
    # extract se
    std <- round(summary(m_full)$coeff[7,2], 4)
    
    # calc CI's
    ci_95 <- data.frame(taxa_coef = coef, taxa_std = std, Lower_95 = coef - 1.96 * std, Upper_95 = coef + 1.96 * std)
    
    ci_50 <- data.frame(Lower_50 = coef - 0.6745 * std, Upper_50 = coef + 0.6745 * std)
    
    # bind data
    df <- cbind(ci_95, ci_50) 
    
    #create data frames
    aic_df <- data.frame(
        metric = i, 
        model = AICc$model,
        dAICc = round(AICc$dAICc,3),
        weight = round(AICc$weight, 3))
    
    results_df <- data.frame(
        metric = i, 
        coef = round(df$taxa_coef, 3),
        std = round(df$taxa_std, 3),
        low_ci_95 = round(df$Lower_95, 3),
        high_ci_95 = round(df$Upper_95, 3),
        low_ci_50 = round(df$Lower_50, 3),
        high_ci_50 = round(df$Upper_50, 3))

    # store results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df

}

# combine results 
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)

```

### iv. plot path rations 

```{r}


## data prep
plot_path_coef <- path_coef %>%
  mutate(direction = ifelse(coef > 0, "positive", "negative")) %>%
  arrange((coef)) %>%
  mutate(order = factor(metric, levels = metric))

plot_path_aic <- path_aic %>%
  select(metric, model, weight) %>%
  spread(key = model, value = weight) %>%          
  mutate(order = factor(metric, levels = levels(plot_path_coef$order))) %>%
  arrange(order)

## plot ##

path_coef_plot <- make_coef_plot(plot_path_coef) +
  #ylim(-0.5, 2) +
  ggtitle("C. Pathway temperature coefficients") +
  ylab("effect size") +
  scale_x_discrete(
    labels = function(x) {
    x <- gsub(
      "bal_facul",
      "frac('Typical of oxygen-tolerant','Typical of obligate anaerobes')",
      x
    )
    x <- gsub(
      "bal_imm",
      "frac('Typical of pathogens','Typical of commensals')",
      x
    )
    parse(text = x)
  }
) 

path_coef_plot

path_aic_plot <- model_weight_plot(plot_path_aic) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 2.25, label = "temp model", color = "#f4b942", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 2.25, label = "base model", color = "#3f7743", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))
  
path_aic_plot

path_plot <- ggarrange(path_coef_plot, path_aic_plot, ncol = 2, nrow = 1)
path_plot

```

### v. manual modeling

only going to use the imm and facultative groupings because the redox grouping doesn't work well with the larger dataset 

```{r}

main <- lm(bal_redox ~ group + age_s + sex + fruit_s + rain_s, data = path_model2)
temp <- lm(bal_redox ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = path_model2)
summary(temp)
AICctab(main, temp, weights = TRUE)

main <- lm(bal_imm  ~ group + age_s + sex + fruit_s + rain_s, data = path_model2)
temp <- lm(bal_imm ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = path_model2)
summary(temp)
AICctab(main, temp, weights = TRUE)

round(summary(temp)$coeff[7,1], 4)

main <- lm(bal_facul ~ group + age_s + sex + fruit_s + rain_s, data = path_model2)
temp <- lm(bal_facul ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = path_model2)
summary(temp)

AICctab(main, temp, weights = TRUE)

```

### vi.  combine tax and path plots

```{r}

p1 <- make_coef_plot(plot_tax_coef) +
  ylim(-0.025, 2.25) +
  ggtitle("A. Taxonomic temperature coefficients") +
  ylab("") +
  scale_x_discrete(
    labels = function(x) {
    x <- gsub(
      "bal_phylum",
      "frac(Proteobacteria, Firmicutes)",
      x
    )
    x <- gsub(
      "bal_oxygen",
      "frac(Oxygen~tolerant, Obligate~anaerobes)",
      x
    )
    parse(text = x)
  }
) 


p2 <- make_coef_plot(plot_path_coef) +
  ylim(-0.025, 2.25) +
  ggtitle("C. Pathway temperature coefficients") +
  ylab("effect size") +
  scale_x_discrete(
    labels = function(x) {
    x <- gsub(
      "bal_facul",
      "frac('Typical of oxygen-tolerant','Typical of obligate anaerobes')",
      x
    )
    x <- gsub(
      "bal_imm",
      "frac('Typical of pathogens','Typical of commensals')",
      x
    )
    parse(text = x)
  }
) 


p3 <- model_weight_plot(plot_tax_aic) +
  ggtitle("B. Taxonomy model weights") +
  annotate("text", x = -0.5, y = 2.3, label = "temp model", color = "#f4b942", size = 5, fontface = "bold") +
  annotate("text", x = 0.5, y = 2.3, label = "base model", color = "#3f7743", size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = rep("", 5),
    limits = c(-1, 1))

p4 <- model_weight_plot(plot_path_aic) +
  ggtitle("D. Taxonomy model weights") +
  scale_x_continuous(breaks = seq(-1, 1, length.out = 5),
    labels = c("100", "50", "0", "50", "100"),
    limits = c(-1, 1)) +
  xlab("model weight")

coef_plots <- ggarrange(p1, p2, ncol = 1, nrow= 2, 
          widths = c(1, 1),
          align = "v",
          heights = c(1, 1))

aic_plots <- ggarrange(p3, p4, ncol = 1, nrow= 2, 
          widths = c(1, 1),
          align = "v",
          heights = c(1, 1))

combo_plot <- ggarrange(coef_plots, aic_plots, ncol = 2, nrow = 1,
          widths = c(1.5, 1),
          align = "v",
          heights = c(1, 1))
combo_plot

```

## D. Redox pathway analysis

model the CLR of the filtered paths, 272

then picks out the paths that are related to temp

rmeove the pathways that are already included in the ratios

then... look at the remaining pathways, should tell a story about lipid metabolism for membrane remodeling, redox relationships, stress pathways, etc. 

```{r}

paths <- path_da3 %>% 
  dplyr::pull(path) %>% 
  unique()

# filter phyloseq object for filtered paths

path_phylo_filtered3 <- prune_taxa(
  taxa_names(path_phylo_filtered2) %in% paths,
  path_phylo_filtered2
)

#clr transformation of filtered phyloseq 

path_table <- as.matrix(t(otu_table(path_phylo_filtered3)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 
dim(path_clr) # 272 filtered paths

# wrangle clr abundances
path_abund <- path_clr %>%
  as_tibble(rownames = "sample") %>%
  pivot_longer(
    cols = -sample,
    names_to = "pathway",
    values_to = "clr_abundance"
  ) %>%
  left_join(path_m_filtered, by = "sample")
  
length(unique(path_abund$pathway)) # 561 pathways
table(path_abund$sample) # each sample has 561 paths

# scale variables
path_abund <- path_abund %>%
  mutate(across(c(age, temp, rain, fruit), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         sex = as.factor(sex),
         ID = as.factor(ID))

```

label with redox status 

```{r}

nucleotide <- path_groups %>% filter(Redox_nucleotide == "nucleotide") %>% pull(path) %>% unique()
length(nucleotide)

redox <- path_groups %>% filter(Redox_nucleotide == "redox") %>% pull(path) %>% unique()
length(redox)

path_abund2 <- path_abund %>%
  mutate(redox = ifelse(pathway %in% redox, "yes", "no"))
table(path_abund2$redox)

path_abund3 <- path_abund2 %>%
  mutate(nucleotide = ifelse(pathway %in% nucleotide, "yes", "no"))
table(path_abund3$nucleotide)

redox_abund <- path_abund3 %>%
  filter(redox == "yes")
length(unique(redox_abund$pathway))

nucleo_abund <- path_abund3 %>%
  filter(nucleotide == "yes")
length(unique(nucleo_abund$pathway))
  
```

test model code 

```{r}

test <- redox_abund %>%
  filter(pathway == "PWY-7977: L-methionine biosynthesis IV")

main <- lm(clr_abundance ~ group + age_s + sex + fruit_s + rain_s, data = test)
temp <- lm(clr_abundance ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = test)
summary(temp)

 hist(test$clr_abundance)

AICctab(main, temp, weights = TRUE)

round(summary(temp)$coefficients["temp_s", 1], 4)

round(summary(temp)$coefficients["temp_s", 2], 4)

```

redox loop

```{r}

redox_coef <- list()
redox_aic <- list()
redox_model_results <- list()

for (i in unique(redox_abund$pathway)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(redox_abund, pathway == i))

    # Fit models
    m_full <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s,
                   data = data)

    m_main <- lmer(clr_abundance ~ age_s + rain_s + sex + group + fruit_s,
                   data = data)

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    redox_aic[[i]] <- aic_df
    redox_coef[[i]] <- results_df
    redox_model_results[[i]] <- model_results_df

  })
}

# combine results
redox_aic <- do.call(rbind, redox_aic)
redox_coef <- do.call(rbind, redox_coef)
redox_model_results <- do.call(rbind, redox_model_results)

```

redox filter 

```{r}

redox_filter <- redox_aic %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(redox_filter)) 

redox_coef2 <- redox_coef %>% filter(path_id %in% redox_filter)
length(redox_coef2$path_id)

redox_aic2 <- redox_aic %>% filter(path_id %in% redox_filter)
length(redox_aic2$path_id)


```

nucleotide loop and filter 

```{r}

nuc_coef <- list()
nuc_aic <- list()
nuc_model_results <- list()

for (i in unique(nucleo_abund$pathway)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(nucleo_abund, pathway == i))

    # Fit models
    m_full <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s,
                   data = data)

    m_main <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s,
                   data = data)

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    nuc_aic[[i]] <- aic_df
    nuc_coef[[i]] <- results_df
    nuc_model_results[[i]] <- model_results_df

  })
}

# combine results
nuc_aic <- do.call(rbind, nuc_aic)
nuc_coef <- do.call(rbind, nuc_coef)
nuc_model_results <- do.call(rbind, nuc_model_results)

```


```{r}

nuc_filter <- nuc_aic %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(redox_filter)) 

redox_coef2 <- redox_coef %>% filter(path_id %in% redox_filter)
length(redox_coef2$path_id)

redox_aic2 <- redox_aic %>% filter(path_id %in% redox_filter)
length(redox_aic2$path_id)


```

try all of the paths

```{r}


path_coef <- list()
path_aic <- list()
path_model_results <- list()

for (i in unique(path_abund3$pathway)) {
  tryCatch({

    # Subset based on taxa
    data <- droplevels(subset(path_abund3, pathway == i))

    # Fit models
    m_full <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s + temp_s,
                   data = data)

    m_main <- lm(clr_abundance ~ age_s + rain_s + sex + group + fruit_s,
                   data = data)

    # AICc Calculation
    AICc <- as.data.frame(AICctab(m_full, m_main, weights = TRUE))
    AICc$model <- rownames(AICc)

    # Extract coefficient and standard error for "temp_s"
    coef_value <- round(summary(m_full)$coefficients["temp_s", 1], 4)
    std_value <- round(summary(m_full)$coefficients["temp_s", 2], 4)

    # Calculate Confidence Intervals
    Lower_95 <- coef_value - 1.96 * std_value
    Upper_95 <- coef_value + 1.96 * std_value
    Lower_50 <- coef_value - 0.6745 * std_value
    Upper_50 <- coef_value + 0.6745 * std_value

    # Create data frames
    aic_df <- data.frame(
      path_id = i, 
      model = AICc$model,
      dAICc = round(AICc$dAICc, 3),
      weight = round(AICc$weight, 3),
      Warning = warning_msg,
      stringsAsFactors = FALSE
    )

    results_df <- data.frame(
      path_id = i, 
      coef = coef_value,
      std = std_value,
      low_ci_95 = Lower_95,
      high_ci_95 = Upper_95,
      low_ci_50 = Lower_50,
      high_ci_50 = Upper_50,
      stringsAsFactors = FALSE
    )

    model_results_df <- data.frame(
      path_id = i,
      Predictor = rownames(summary(m_full)$coefficients)[-1], # Exclude intercept
      Coefficient = round(summary(m_full)$coefficients[-1, 1], 4),
      StdError = round(summary(m_full)$coefficients[-1, 2], 4),
      stringsAsFactors = FALSE
    )

    # Save results
    path_aic[[i]] <- aic_df
    path_coef[[i]] <- results_df
    path_model_results[[i]] <- model_results_df

  })
}

# combine results
path_aic <- do.call(rbind, path_aic)
path_coef <- do.call(rbind, path_coef)
path_model_results <- do.call(rbind, path_model_results)

```


```{r}

path_filter <- path_aic %>%
  group_by(path_id) %>%
  filter(model == "m_full" & weight > 0.5) %>%
  pull(path_id) 
length(unique(path_filter)) 

path_coef2 <- path_coef %>% filter(path_id %in% path_filter)
length(path_coef2$path_id)

path_aic2 <- path_aic %>% filter(path_id %in% path_filter)
length(path_aic2$path_id)

```







