---
title: "Untitled"
author: "Lucore"
date: "2026-04-28"
output: html_document
---

Notes: Here we are going to try the ratio analysis with a neopterin interaction 

We find that an interaction between neopterin and tempature is not as good at explaining the ration values as the model with only temperature, compared to the base model. So it is not really worthwhile to present the interaction results in addition to the temperature ration results. 

Going to save this in case I ever need to remind myself I tried this. Important to know that this was quick and dirty. I did not remake the taxa rations for this specific dataset. The pathway ratios aligned with this data set. If we ever wanted to publish this, would need to incorporate this more carefully. For now, I think it is fine. 

# 1. Load environment

```{r}

# packages
library(tidyverse)
library(lme4)
library(conflicted)
library(mgcv)
library(gratia)
library(bbmle)
library(ggpubr)
library(factoextra)
library(gridExtra)
library(ggnewscale)
library(zCompositions)
library(ggtree)
library(tidytree)
library(ape)
library(jsonlite)
library(file2meco)
library(phyloseq)
library(flextable)
library(gridtext)
library(grid)

#physeq objects
tax_physeq <- readRDS("phyloseq_objects/2024-12-18_phyloseq_16s.rds")
path_physeq_u50 <- readRDS("phyloseq_objects/pathway_phylo_u50.rds")

# functions
source("functions/my_scale.R")
source("functions/inverse_scale.R")
source("functions/kegg_parse_json.R")
source("functions/make_coef_plot.R")
source("functions/model_weight_plot.R")
source("functions/model_weight_plot.R")
source("functions/generate_shades.R")

# conflicts
conflict_prefer("select", "dplyr")
conflicts_prefer(dplyr::filter)
conflicts_prefer(cowplot::get_legend)


```


# 2. Taxa analysis 

## A. Set up (16s seqs)

### i. Metadata filtering

using 277 samples that have feeding and neopterin metadata, neopterin value is neopterin averaged 3 days before and after GM sample

```{r}

tax_m <- data.frame(sample_data(tax_physeq)) %>% 
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, fruit, residuals, neo)

tax_m_filtered <- tax_m[!is.na(tax_m$fruit), ] # remove samples with no diet data
tax_m_filtered <- tax_m_filtered[!is.na(tax_m_filtered$neo), ] # remove samples with no neo, 277 samples left

tax_physeq2 <- prune_samples(rownames(tax_m_filtered), tax_physeq) # filter physeq object

tax_physeq2 <- prune_taxa(taxa_sums(tax_physeq2) > 0, tax_physeq2) # remove ASV's with zero abundance

tax_physeq2

```

### ii. Prevalence filtering 

Identifies taxa that are not present in at least 0.5% of the total samples (at least two samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth.  

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(tax_physeq2),
                MARGIN = ifelse(taxa_are_rows(tax_physeq2), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(tax_physeq2),
                      tax_table(tax_physeq2))

# define prevalence threshold
prevalenceThreshold = 0.005 * nsamples(tax_physeq2)  
prevalenceThreshold # must appear in at least two samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

ggsave("figures/supp/supp_fig4.pdf", prev_fig, width = 6, height = 4)

# summary stats
asv_onesample <- nrow(prevdf[prevdf$Prevalence == 1,]) # number of ASV's present only in one sample
total_asv <- nrow(prevdf)
percent_onesample <- (asv_onesample/total_asv) * 100 # percent ASV's in only one sample

asv_nophyla <- nrow(prevdf[is.na(prevdf$Phylum),]) # number of ASV's with no assigned phylum 
percent_nophyla <- (asv_nophyla/total_asv) * 100 # percent number of ASV's with no phylum

## execute prevalence filter ##
tax_phylo_filtered <- prune_taxa((prev0 > prevalenceThreshold), tax_physeq2) 
tax_phylo_filtered

## remove taxa with unnasigned phyla ## 
tax_phylo_filtered2 <- subset_taxa(tax_phylo_filtered, !is.na(Phylum)) # remove 102 taxa 
tax_phylo_filtered2

# calc reads per sample
reads_filtered <- as.data.frame(sample_sums(tax_phylo_filtered2)) %>% 
  rownames_to_column(var = "sample") %>%
  rename(filtered_reads = 'sample_sums(tax_phylo_filtered2)')
  
# add reads per sample to metadata
tax_m_filtered2 <- tax_m_filtered %>%
  left_join(reads_filtered, by = "sample") 

# % assignment for remaining taxonomy levels

asign_class <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Class"])))/1159)*100)

asign_order <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Order"])))/1159)*100)

asign_family <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Family"])))/1159)*100)

asign_genus <- 100-(((sum(is.na(tax_table(tax_phylo_filtered2)[, "Genus"])))/1159)*100)

######## evaluate which phlya have greater than 1% rel abund ############

phylum_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Phylum"))
  
phylum_otu_table <- as.data.frame(otu_table(phylum_aggregated))
phylum_tax_table <- as.data.frame(tax_table(phylum_aggregated))

phylum_otu_ids <- rownames(phylum_otu_table)
phylum_tax_ids <- rownames(phylum_tax_table)

phylum_names <- phylum_tax_table[phylum_otu_ids, "Phylum"]

rownames(phylum_otu_table) <- phylum_names

# across samples
phylum_otu_table2 <- phylum_otu_table %>%
  rownames_to_column("Phylum") %>%
  pivot_longer(-Phylum, names_to = "sample", values_to = "Count") %>%
  left_join(reads_filtered, by = "sample") %>%
  group_by(sample) %>%
  mutate(RelativeAbundance = Count / sum(Count),
          Phylum_grouped = case_when(
           str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
           RelativeAbundance > 0.01 ~ Phylum,  
           TRUE ~ "Less than 1% rel abund"  
         )) %>%
  ungroup()

df_total <- phylum_otu_table2 %>%
  mutate(Phylum_grouped = case_when(
    str_detect(Phylum, "Firmicutes") ~ "Firmicutes",  
    TRUE ~ Phylum)) %>%
  group_by(Phylum_grouped) %>%
  summarise(TotalCount = sum(Count)) %>%
  ungroup() %>%
  mutate(RelativeAbundance = TotalCount / sum(TotalCount)) %>%
  mutate(Phylum_grouped = case_when(RelativeAbundance > 0.01 ~ Phylum_grouped, 
                                    TRUE ~ "Less than 1% rel abund")) %>%
  group_by(Phylum_grouped) %>%
  summarise(RA = sum(RelativeAbundance)) %>%
  ungroup() %>%
  mutate(RA = 100*RA,
         Phylum_grouped = fct_reorder(Phylum_grouped, RA))

ggplot(df_total, aes(x = Phylum_grouped, y = RA, fill = Phylum_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Relative Abundance (%)", fill = "Phylum") +
  theme_minimal() +
  coord_flip() 

```

###  iii. Generate alpha diversity metrics

```{r}

shannon <- estimate_richness(tax_phylo_filtered2, measures = "Shannon") # calc shannon
shannon <- shannon %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

hist(shannon$shannon) # normal dist

observed <- estimate_richness(tax_phylo_filtered2, measures = "Observed") # calc observed
observed <- observed %>%
  rownames_to_column(var = "sample") %>%
  rename(observed = Observed) %>%
  mutate( sample = str_replace_all(sample, "^X", ""),  
          sample = str_replace_all(sample, "\\.", "-"))

hist(observed$observed) # gamma dist


tax_m_filtered3 <- tax_m_filtered2 %>% # add to metadata
  left_join(shannon, by = "sample") %>%
  left_join(observed, by = "sample") 

```

### iv. CLR transformationand PCA

The rows need to be samples when doing the PCA because results show how samples cluster based on ASV's instead of the samples themselves

Use the CLR transformation and PCA analysis because of Gloor et al., 2017, Baniel publications, Gloor and Ried 2016, and this tutorial https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot reccomend that normalizing for sequencing depth is best done with a CLR transformation (as opposed to rarefaction) and that sample distances are best estimated using Aitchisons distances as opposed to Bray Curtis or Unifrac because they do not reflect the compositional nature of the data. Aitchison distance is the euclidean distances between the samples after the CLR transformation.

Retain PCA axes 1-3 because they explain >5% of the variance

```{r}

######### transformation and PCA #########

otu_table <- as.matrix(t(otu_table(tax_phylo_filtered)))
otu_table[1:5,1:5] # rows are samples 

set.seed(123)
czm <- cmultRepl(otu_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
czm[1:5,1:5] # rows are samples 

clr <- t(apply(czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
clr[1:5,1:5] # rows are samples 

pca <- prcomp(clr) # calc pca
plot(pca$x[,1],pca$x[,2]) # plot 

d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="") # PC1 14.4%
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="") # PC2 8.1%

######### make scree plot #########

eigenvalues <- pca$sdev^2  # square the standard deviations to get eigenvalues
percent_explained <- 100 * eigenvalues / sum(eigenvalues) 

# scree plot
pe_plot <- tibble(pe = percent_explained, axis = 1:length(percent_explained)) %>%
  ggplot(aes(x = axis, y = pe)) +
  geom_line() +
  geom_point() +
  labs(x = "Principal component axis", y = "Percent variance explained") +
  scale_y_continuous(breaks = seq(0, max(percent_explained), by = 2.5)) +
  theme_minimal() +
  theme(axis.line = element_line(color = "black", linewidth = .28),
        axis.text = element_text(size = 12, color = "black"), 
        axis.ticks.y = element_line(color = "black", linewidth = 0.3),
        axis.ticks.x = element_line(color = "black", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank())

ggsave("figures/supp/supp_fig6.pdf", pe_plot, width = 6, height = 6)

######### Add PCA values to metadata #########

df<-data.frame(pca$x)
df$sample<-rownames(df)
df_small <- df[, c(1:3,278)]

# add to metadata
tax_m_filtered4 <- tax_m_filtered3 %>%
  left_join(df_small, by = "sample") 

```

### v. prep data for modeling

scale variables and make categorical variables factors. Make all palmas sampels tenori, only 6 palmas samples not enough to estimate effect and they have overlapping home ranges.

```{r}

# scale variables
tax_model <- tax_m_filtered4 %>%
  mutate(across(c(age, temp, rain, fruit, residuals, neo), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           TRUE ~ group)),
         ID = as.factor(ID))  

```

## B. Ratio modeling

### i. Preprocessing

clean up taxa names in phyloseq object, use `tax_glom()` to combine taxa at the genus level and at the family level (i.e. sum abundances of the ASV's that are grouped together into genus or family). Change row names from ASV ID to taxa name. 

```{r}

# get rid of all the weird, unnassigned names and turn them all into NA's
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("Unassigned*",NA,x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("uncultured*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("unidentified*",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("gut metagenome",NA, x)))
tax_table(tax_phylo_filtered2)<-apply(tax_table(tax_phylo_filtered2),2, function(x) (gsub("metagenome",NA, x)))

# Tax glom 
genus_aggregated <- suppressMessages(phyloseq::tax_glom(tax_phylo_filtered2, "Genus"))

# extract tables
genus_otu_table <- as.data.frame(otu_table(genus_aggregated))
genus_tax_table <- as.data.frame(tax_table(genus_aggregated))
genus_metadata <- sample_data(genus_aggregated)

#change OTU rownames to ASV family names
# Ensure that the OTU IDs are the same in both tables (rows in otu_df and tax_df)
genus_otu_ids <- rownames(genus_otu_table)
genus_tax_ids <- rownames(genus_tax_table)

# Extract the taxa names from the taxonomy table
genus_names <- genus_tax_table[genus_otu_ids, "Genus"]

# Set the row names of the OTU table to the family names
rownames(genus_otu_table) <- genus_names

```

### ii. Calc rel abundance 

Calculate the relative abundance by dividing the raw abundance counts by the total reads per sample. Using the "filtered_reads" column because we are using the phyloseq object `tax_phylo_filtered2` that has already been filtered at a 0.5% threshold. 

```{r}

# extract table for DA testing
genus_da <-psmelt(genus_aggregated)

# Extract variables of interest 
genus_da1 <- genus_da %>% 
  mutate(Taxa = Genus) %>% # make genus the taxa values 
  select(Taxa, Abundance, Sample, age, sex, ID, max_t_days15, rain_days4, fruit, group, Phylum, Family, Class) %>% 
  rename(sample = Sample,
         temp = max_t_days15,
         rain = rain_days4) %>%
  left_join(reads_filtered, by = "sample") %>%
  mutate(level = "Genus") %>%
  mutate(RA = (Abundance/filtered_reads)*100)

# n = 277
length(unique(genus_da1$sample))  

```

### iii. Filter out low abundance taxa

Calculate the average relative abundance for each taxa across all of the samples and then keep the taxa that have at lease 0.2% relative abundance across all samples. Left with 128 genera 

```{r}

#Calculate average relative abundance of each taxa across all samples
genus_a1<-sapply(unique(genus_da1$Taxa),function(x) mean(genus_da1$RA[genus_da1$Taxa==x],na.rm=TRUE))
genus_a2<-data.frame(cbind(Taxa=as.character(unique(genus_da1$Taxa)),RA=genus_a1))
genus_a2$RA<-as.numeric(as.character(genus_a2$RA))

#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

genus_da2 <-droplevels(subset(genus_da1, genus_da1$Taxa %!in% NA)) # remove asv NAs
genus_da3 <-droplevels(subset(genus_da2, genus_da2$Taxa %in% unique(genus_a2$Taxa[genus_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(genus_da2$Taxa))
length(unique(genus_da3$Taxa)) # 134 genera

```

### iv. Scale variables

scale variables for modeling, 128 total taxa for data 

not going to combine palmas and tenori

```{r}

# Scale variables
da4 <- genus_da3 %>%
   mutate(across(c(temp, rain, age, fruit), my_scale, .names = "{.col}_s"),
         ID = as.factor(ID),
         sex = as.factor(sex)) 
  #mutate(group = as.factor(case_when(group == "Palmas" ~ "Tenori",
                           #TRUE ~ group)))

length(unique(da4$Taxa)) # 134 taxa total

#pull out genera classification to make ration assignments 
#genus <- da4 %>% select(Taxa, Phylum, Family, Class) %>% distinct()
#write.csv(genus, "data/genera.csv", row.names = FALSE)

```

### v. Creat ratios 

We aggregated counts to the genus level used multiplicative zero replacement `cmultRepl()` to replace zeros. For each predefined, biologically informed contrast, we computed the log-contrast as the difference of mean log-abundances between the numerator and denominator genus sets, i.e., the log of the ratio of their geometric means. Because balances are log-ratios, they are controlled for sequencing depth and live in Aitchison log-ratio space. 

REMEMBER STILL NEED TO UPDATE SOURCE LIST FOR CELLS IN YELLOW ON 2026-04-14_genera_categories.csv

```{r}

# make data a long format and convert to matrix  
genera_long <- da4 %>%
  select(Taxa, Abundance, sample) %>%
  pivot_wider(names_from = Taxa, values_from = Abundance, values_fill = 0) %>%
  arrange(sample) %>% 
  column_to_rownames("sample") %>%
  as.matrix()

# add psuedo count 
genera_czm <- cmultRepl(genera_long, label=0, method="CZM", z.warning = 1, z.delete = 1)
rs <- rowSums(as.matrix(genera_czm)) # check that each sample equals 1

#### build groups ###

groups <- da4 %>% select(Taxa, Phylum) %>% distinct()
  
taxa_groups <- read_csv("data/2026-04-21_genera_categories.csv") %>%
  select(Taxa, Oxygen_tolerance)


test <- anti_join(groups, taxa_groups, by = "Taxa")

groups2 <- groups %>%
  left_join(taxa_groups, by = "Taxa") 


# phyla group 
groups3 <- groups2 %>%
  mutate(Phylum_level = case_when(Phylum == "Firmicutes_A" ~ "firmi",
                                  Phylum == "Firmicutes_C" ~ "firmi",
                                  Phylum == "Firmicutes_D" ~ "firmi",
                                  Phylum == "Proteobacteria" ~ "proteo"))

####### make ratios ########

num_oxy <- groups3 %>% filter(Oxygen_tolerance == "facultative_anaerobe") %>% pull(Taxa) %>% unique()
den_oxy <- groups3 %>% filter(Oxygen_tolerance == "obligate_anaerobe") %>% pull(Taxa) %>% unique()

num_phy <- groups3 %>% filter(Phylum_level == "proteo") %>% pull(Taxa) %>% unique()
den_phy <- groups3 %>% filter(Phylum_level == "firmi") %>% pull(Taxa) %>% unique()

# compute balances
source("functions/make_balance_from_matrix.R")

B_oxy <- make_balance_from_matrix(genera_czm, num_oxy, den_oxy)
B_phy <- make_balance_from_matrix(genera_czm, num_phy, den_phy)

balances <- data.frame(
  sample = rownames(genera_czm),
  bal_oxygen   = as.numeric(B_oxy),
  bal_phylum   = as.numeric(B_phy))

# add to mediation data

tax_model2 <- tax_model %>%
  left_join(balances, by = "sample")
  

```

### vi. Model ratios

```{r}

# phylum
main <- lmer(bal_phylum ~ group + age_s + sex + fruit_s + rain_s + (1 | ID), data = tax_model2)
temp <- lmer(bal_phylum ~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID), data = tax_model2)
summary(temp)

neo <- lmer(bal_phylum ~ group + age_s + sex + fruit_s + rain_s + temp_s * neo_s + (1 | ID), data = tax_model2)
summary(neo)

AICctab(main, temp, neo, weights = TRUE)

#oxygen
main <- lmer(bal_oxygen ~ group + age_s + sex + fruit_s + rain_s + (1 | ID), data = tax_model2)
temp <- lmer(bal_oxygen ~ group + age_s + sex + fruit_s + rain_s + temp_s + (1 | ID), data = tax_model2)
summary(temp)

neo <- lmer(bal_oxygen ~ group + age_s + sex + fruit_s + rain_s + temp_s * neo_s + (1 | ID), data = tax_model2)
summary(neo)

AICctab(main, temp, neo, weights = TRUE)


```

# 3. Pathways analysis (shotgun seqs)

## A. set up 

### i. Metadata filtering

using 71 samples that have feeding and neotperin metadata

```{r}

path_m <- data.frame(sample_data(path_physeq_u50)) %>% # extract metadata 
  rownames_to_column("sample") %>%
  rename(temp = max_t_days15,
         rain = rain_days4,
         neo = total_3_day) %>%
  select(sample, ID, time, date, group, sex, age, repro, temp, rain, neo, fruit, residuals, neo, reads)

path_m_filtered <- path_m[!is.na(path_m$fruit), ] # remove samples with no diet data, drops 8 samples

path_m_filtered <- path_m_filtered[!is.na(path_m_filtered$neo), ] # remove samples with no neopterin data, down to 71 samples


# filter phyloseq objects to remove samples with no diet data 
path_physeq2 <- prune_samples(path_m_filtered$sample, path_physeq_u50) # filter physeq object
path_physeq2 <- prune_taxa(taxa_sums(path_physeq2) > 0, path_physeq2) # remove ASV's with zero abundance

```

### ii. Prevalence filtering

Remove unmapped and unintegrated values. Unmapped values represent the total abundance of reads that failed to align to any known gene sequences and unintegrated values represent genes that did not belong to a known pathway.

```{r}

path_filtered <- subset_taxa(path_physeq2, 
                                !(grepl("UNMAPPED", pathway) | grepl("UNINTEGRATED", pathway)))

```

### iii. Generate alpha diversity metrics

going to calculate this on the raw RPK's and add an offset for seq depth, use the number of paired reads from the kneaddata file

Ignore the warning, `estimate_richness()` prefers count data but it gives you exactly the same values as if you calculated it with vegan. 

```{r}

shannon <- estimate_richness(path_filtered, measures = "Shannon") %>%
  rownames_to_column(var = "sample") %>%
  rename(shannon = Shannon) 

path_alpha <- as.data.frame(colSums(otu_table(path_filtered) > 0)) %>%
  rename(observed = 'colSums(otu_table(path_filtered) > 0)') %>%
  rownames_to_column(var = "sample") %>%
  full_join(shannon, by = "sample")

hist(path_alpha$shannon) 
hist(path_alpha$observed)

path_m1 <- path_m_filtered %>% # add to metadata
  left_join(path_alpha, by = "sample")

```

### iv. CLR transformation and PCA 

```{r}

######### transformation and PCA #########

path_table <- as.matrix(t(otu_table(path_filtered)))
path_table[1:5,1:5] # rows are samples 

set.seed(123) # set seed for reproducibility
path_czm <- cmultRepl(path_table,label=0, method="CZM", z.warning = 0.90, z.delete = 0.90)
path_czm[1:5,1:5] # rows are samples 

path_clr <- t(apply(path_czm, 1, function(x){log(x) - mean(log(x))})) # must transpose again
path_clr[1:5,1:5] # rows are samples 


path_pca <- prcomp(path_clr) # calc pca
plot(path_pca$x[,1], path_pca$x[,2]) # plot 

path_d.mvar <- sum(path_pca$sdev^2) # total variance
path_PC1 <- paste("PC1 ","(",round(sum(path_pca$sdev[1]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC1 21.5%
path_PC2 <- paste("PC2 ","(",round(sum(path_pca$sdev[2]^2)/path_d.mvar, 3)*100,"%",")",sep="") # PC2 13.1%

######### Add PCA values to metadata #########

path_df<-data.frame(path_pca$x)
path_df$sample<-rownames(path_df)
path_df_small <- path_df %>% 
  select(c(1:3,72))

path_m2 <- path_m1 %>%
  left_join(path_df_small, by = "sample")

```

### v. Prep data to model

scale variables and make categorical variables factors

```{r}

path_model <- path_m2 %>%
  mutate(across(c(age, temp, rain, fruit, neo, residuals), my_scale, .names = "{.col}_s")) %>%
  mutate(group = as.factor(group),
         ID = as.factor(ID))

```

## B. Ratio modeling
### i. Filter by prevalence

Identifies pathways that are not present in at least 30% of the total samples (at least 27 samples) and removed them. Add column "filtered reads" to metadata with the reads retained per sample to control for for sequencing depth. 

```{r}

# calculate many samples did each taxa appear at least once
prev0 = apply(X = otu_table(path_filtered),
                MARGIN = ifelse(taxa_are_rows(path_filtered), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})

# make a dataframe that shows the prevalence (how many sample did the taxa show up in) and total abundance (total reads of that taxa) with taxonomy 
prevdf = data.frame(Prevalence = prev0,
                      TotalAbundance = taxa_sums(path_filtered),
                      tax_table(path_filtered))

# define prevalence threshold
prevalenceThreshold = 0.3 * nsamples(path_filtered)  
prevalenceThreshold # must appear in at least 27 samples

# plot prevalence threshold
prev_fig <- ggplot(prevdf, aes(TotalAbundance, Prevalence)) +
  geom_hline(yintercept = prevalenceThreshold, alpha = 0.5, linetype = 2) +
  geom_point(size = 2, alpha = 0.5) +
  scale_y_log10() + scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence (# of samples)") +
  theme_minimal()

## execute prevalence filter ##
path_phylo_filtered2 <- prune_taxa((prev0 > prevalenceThreshold), path_filtered) 
path_phylo_filtered2

```

### ii. Filter by rel abundance

```{r}

# extract data for DA testing

path_da <-psmelt(path_phylo_filtered2)

# Extract variables of interest 
path_da1 <- path_da %>% 
  select(OTU, Abundance, Sample, reads) %>% 
  rename(gm_sample = Sample,
         path = OTU) %>%
  mutate(RA = (Abundance/reads)*100)

# n = 87
length(unique(path_da1$gm_sample))  

#Calculate average relative abundance of each taxa across all samples
path_a1<-sapply(unique(path_da1$path),function(x) mean(path_da1$RA[path_da1$path==x],na.rm=TRUE))
path_a2<-data.frame(cbind(path=as.character(unique(path_da1$path)),RA=path_a1))
path_a2$RA<-as.numeric(as.character(path_a2$RA))

#Filter taxa for relative abundance threshold 

'%!in%'<-function(x,y)!('%in%'(x,y))

path_da2 <-droplevels(subset(path_da1, path_da1$path %!in% NA)) # remove asv NAs
path_da3 <-droplevels(subset(path_da2, path_da2$path %in% unique(path_a2$path[path_a2$RA>=0.002]))) # keep taxa only >= .2% rel abundance in all samples

length(unique(path_da3$path))


paths <- path_da3 %>% select(path) %>% distinct()

nrow(paths) # 270

#list of paths for category assignment
#write.csv(paths, "data/paths.csv", row.names = FALSE)

```

### iii. Ratio analysis 

#### a. create ratios 

Immun ratio
  Immune antagonist : functions typical of commensals/immune regulatory
  
Facul ratio
  typical of aerotolerant bacteria : typical of obligately anaerobic bacteria 
  ** does not incude general respiration pathas as not trying to show how repsiratory mode shifts with temp but rather community function related to health 

```{r}

# make data a long format and convert to matrix  
path_long <- path_da3 %>%
  select(path, Abundance, gm_sample) %>%
  pivot_wider(names_from = path, values_from = Abundance, values_fill = 0) %>%
  arrange(gm_sample) %>% 
  column_to_rownames("gm_sample") %>%
  as.matrix()

# add psuedo count 
path_czm <- cmultRepl(path_long, label=0, method="CZM", z.warning = 1, z.delete = 1)
rs <- rowSums(as.matrix(path_czm)) # check that each sample equals 1


####### make ratios ########

path_groups <- read_csv("data/2026-04-20_path categories.csv") # load categories 

num_imm <- path_groups %>% filter(Immune_antagonist == "antagonist") %>% pull(path) %>% unique()
den_imm <- path_groups %>% filter(Immune_antagonist == "commensal") %>% pull(path) %>% unique()

num_facul <- path_groups %>% filter(Facul_obligate == "facultative") %>% pull(path) %>% unique()
den_facul <- path_groups %>% filter(Facul_obligate == "obligate") %>% pull(path) %>% unique()

# compute balances

B_imm <- make_balance_from_matrix(path_czm, num_imm, den_imm)
B_facul <- make_balance_from_matrix(path_czm, num_facul, den_facul)

balances <- data.frame(
  sample = rownames(path_czm),
  bal_facul = as.numeric(B_facul),
  bal_imm = as.numeric(B_imm))

# add to modeling data

path_model2 <- path_model %>%
  left_join(balances, by = "sample")

```

#### d. manual modeling

for checking code and ration content

```{r}

main <- lm(bal_imm  ~ group + age_s + sex + fruit_s + rain_s, data = path_model2)
temp <- lm(bal_imm ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = path_model2)
summary(temp)

neo <- lmer(bal_imm ~ group + age_s + sex + fruit_s + rain_s + temp_s * neo_s + (1 | ID), data = path_model2)
summary(neo)

AICctab(main, temp, neo, weights = TRUE)


main <- lm(bal_facul ~ group + age_s + sex + fruit_s + rain_s, data = path_model2)
temp <- lm(bal_facul ~ group + age_s + sex + fruit_s + rain_s + temp_s, data = path_model2)
summary(temp)

neo <- lmer(bal_facul ~ group + age_s + sex + fruit_s + rain_s + temp_s * neo_s + (1 | ID), data = path_model2)
summary(neo)

AICctab(main, temp, neo, weights = TRUE)


```



